---
title: "01_preprocess.Rmd"
author: "Alyssa N. Easton"
date: '2026-05-13'
output:
---

## Inputs
- `data/raw/VNS_behavior_250314.xlsx`
- `data/raw/VNS_datafinalpaper_240513.xlsx`
- `data/raw/metaphlan/*.txt`

## Outputs
- `data/intermediate/01_meta_A.rds`
- `data/intermediate/01_meta_B.rds`

Outputs contain processed microbiome data, metadata, cytokine PCA,
and higher-order taxonomic information for species assignments

```{r "setup", include=FALSE}
source(here::here("R", "paths.R"))
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
knitr::opts_knit$set(root.dir = PROJECT_ROOT)
options(digits = 10)
library(tidyverse)
library(data.table)
library(magrittr)
library(readxl)
library(janitor)
source(here::here("R", "utils.R"))
source(here::here("R", "cleaning.R"))
source(here::here("R", "microbiome_preprocess.R"))
source(here::here("R", "diversity_utils.R"))
source(here::here("R", "constants.R"))
source(here::here("R", "pca_utils.R"))
```

## Prepare Metadata
```{r read}
# updated behavior data
pd_data <- read_xlsx(here::here("data", "raw", "VNS_behavior_250314.xlsx")) %>%
  dplyr::select(animal_code, startle_habituation, startle_PPI, PD_choice,
                EtOH_24h_intakeave, EtOH_pref_24h, EtOH_8pm_intakeave) %>%
  rename(PD_choice_filt = PD_choice,
         EtOH_24h_intakeave_final = EtOH_24h_intakeave,
         EtOH_pref_24h_final = EtOH_pref_24h,
         EtOH_8pm_intakeave_final = EtOH_8pm_intakeave)

# rest of data
metadata <- read_xlsx(here::here("data", "raw", "VNS_datafinalpaper_240513.xlsx")) %>%
  dplyr::select(where(~ any(!is.na(.)))) %>%           # remove empty cols
  mutate(animal_code = as.integer(animal_code)) %>%    # convert animal_code to int
  dplyr::select(-matches("p__|o__")) %>%               # remove old microbiome-related columns
  dplyr::select(-c(observed, chao1, shannon, PC1, PC2)) %>%
  left_join(pd_data, by = "animal_code") %>%           # merge in the behavior data
  mutate(
    # make Injury into factor, capitalize
    Injury = factor(Injury, levels = c("sham","blast")) %>%
      fct_recode("Sham" = "sham", "Blast" = "blast"),
    # make Treatment into factor, capitalize
    Treatment = factor(Treatment, levels = c("sham","VNS")) %>%
      fct_recode("Sham" = "sham", "VNS" = "VNS"),
    # make Round_main into factor
    Round_main = factor(Round_main, levels = c("1", "2", "3", "4", "5", "6", "7", "8", "9")),
    # make Cage id into factor
    Cage = factor(Cage),
    # add a feature called "cohort", which is either A or B
    cohort = case_when(
      feces_tp == 0.16 ~ "A",
      feces_tp == 1 ~ "B"),
    cohort = factor(cohort),
    # add a feature for seq_name, which is the ID listed on all sequencing files
    seq_name = case_when(
      animal_code == 1774 ~ 1775L,
      animal_code == 1363 ~ 1362L,
      animal_code == 1372 ~ 1377L,
      TRUE ~ animal_code)
    ) %>%
  # sort table by the SEQUENCING ID
  arrange(seq_name) %>%
  # replace special characters for later processing, because I'm not sure how
  # they are handled in Linux shell scripts
  rename_with(.fn = clean_vec) %>%
  rename(`brain_IFNg` = `brain_INFg`) %>%       # it's IFN, not INF
  # rename(
  #   c("serum_IL-1a" = `serum_IL-1α`,
  #     "brain_IL-1a" = `brain_IL-1α`,
  #     "brain_IFN_gamma" = `brain_INFγ`)
  # ) %>%
  # add a feature called "inj_trt_group" which has our 4 treatment groups
  mutate(
    inj_trt_group = factor(
      paste0(.$Injury, " + ", .$Treatment),
      levels = c("Sham + Sham", "Blast + Sham", "Sham + VNS", "Blast + VNS")
      ),
    blast         = ifelse(Injury    == "Blast",                   1, 0),
    vns           = ifelse(Treatment == "VNS",                     1, 0),
    blast_and_vns = ifelse(Injury == "Blast" & Treatment == "VNS", 1, 0),
    seq_name      = as.character(seq_name)
  )
```

```{r}
meta_a <- clean_meta(filter(metadata, cohort == "A"))
meta_b <- clean_meta(filter(metadata, cohort == "B"))
meta_b_smaller <- meta_b %>%
  dplyr::select(-any_of(extra_B_cols))
```

# Cytokine PCA
Tidy data import/PCA
Here, I'm using a "helper" function `run_.pca.join_pca` to automatically
   • filter to one cohort
   • select & scale a block of columns
   • run PCA
   • pick PCs ≥ threshold
   • augment metadata with those PCs

The resulting object is a list, including
   • meta: augmented metadata df including PCs ≥ threshold `res$meta$PCx`
   • pca: the output of prcomp() `res$pca`
   • variance: the variance explained per PC `res$variance`
   • loadings: loadings of each feature in each PC `res$loadings`
   
```{r}
res_A <- run_pca(meta_a, block = "serum|brain")

# extract PCs explaining ≥ 5% total variance
ck_pcs <- res_A$vars %>%
  filter(`Proportion of Variance` >= 0.05) %>%
  pull(PC)

pca_loadings <- res_A$loadings %>%
  dplyr::select(any_of(c("Feature", ck_pcs)))

pca_scores <- res_A$meta %>%
  dplyr::select(any_of(c("seq_name", ck_pcs)))

# update meta_a to include PCs
meta_a_withPCs <- meta_a %>%
  left_join(pca_scores, by = "seq_name")
```

# Microbiome
### Extract species-level counts
Process MetaPhlAn output into *species-level count table*
```{r}
files <- list.files("/proj/gibbons/2024_easton_vns/metaphlan_2/metaphlan", 
                    pattern = "_metaphlan\\.txt$", full.names = TRUE)

dt <- rbindlist(lapply(files, function(f) {
  fread(f)[, seq_name := sub("_metaphlan\\.txt$", "", basename(f))]
}), fill = TRUE) %>%
  rename(
    clade_name = `#clade_name`,
    est_n_reads = estimated_number_of_reads_from_the_clade
    ) %>%
  filter(seq_name %in% metadata$seq_name)

s_long <- dt %>%                 # we can do this for any taxonomic level
  filter(
    grepl("\\|s__", clade_name) | clade_name == "UNCLASSIFIED",
    !grepl("\\|t__", clade_name)
  ) %>%
  mutate(
    clade_name = ifelse(
      clade_name == "UNCLASSIFIED",
      "k__unclassified|p__unclassified|c__unclassified|o__unclassified|f__unclassified|g__unclassified|s__unclassified",
      clade_name
      )
  )

full_taxonomy <- data.frame(full_name = unique(s_long$clade_name)) %>%
  separate(
    full_name,
    sep = "\\|",
    into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"), # no "Strain"
    fill = "right",
    remove = FALSE
  ) %>%
  mutate(
    Kingdom = gsub("k__", "", Kingdom),
    Phylum  = gsub("p__", "", Phylum),
    Class   = gsub("c__", "", Class),
    Order   = gsub("o__", "", Order),
    Family  = gsub("f__", "", Family),
    Genus   = gsub("g__", "", Genus),
    Species = gsub("s__", "", Species)
  )

# check that all species names are still unique
length(unique(full_taxonomy$Species)) == length(unique(full_taxonomy$full_name))

s_long <- s_long %>%
  left_join(
    full_taxonomy %>%
      dplyr::select(full_name, Species),
    by = c("clade_name" = "full_name")
    ) %>%
  mutate(
    clade_name = coalesce(Species, clade_name)
    ) %>%
  dplyr::select(-Species)

s_counts <- s_long %>%
  pivot_wider(
    ., id_cols = "seq_name", names_from = "clade_name", values_from = "est_n_reads"
    ) %>%
  dplyr::select(-unclassified)   # remove because some values are negative

s_relabund <- s_long %>%
  pivot_wider(., id_cols = "seq_name", names_from = "clade_name", values_from = "relative_abundance")
```
  
```{r}
# Get sequence IDs for each cohort
seq_a <- metadata %>% filter(cohort == "A") %>% pull(seq_name)
seq_b <- metadata %>% filter(cohort == "B") %>% pull(seq_name)

# separate into 2 raw count tables
counts_both <- s_counts %>%
  filter(seq_name %in% c(seq_a, seq_b)) %>%
  dplyr::select(where(~ !all(is.na(.)))) %>%
  mutate(across(everything(), ~replace_na(., 0))) %>%
  column_to_rownames("seq_name")

counts_a <- s_counts %>%
  filter(seq_name %in% seq_a) %>%
  dplyr::select(where(~ !all(is.na(.)))) %>%
  mutate(across(everything(), ~replace_na(., 0))) %>%
  column_to_rownames("seq_name")

counts_b <- s_counts %>%
  filter(seq_name %in% seq_b) %>%
  dplyr::select(where(~ !all(is.na(.)))) %>%
  mutate(across(everything(), ~replace_na(., 0))) %>%
  column_to_rownames("seq_name")

# also just go ahead and separate the relative abundance into a and b
relabund_a <- s_relabund %>%
  filter(seq_name %in% seq_a) %>%
  dplyr::select(where(~ !all(is.na(.))))
relabund_b <- s_relabund %>%
  filter(seq_name %in% seq_b) %>%
  dplyr::select(where(~ !all(is.na(.))))
```

### Rarefy counts
for calculation of alpha and beta diversity metrics
```{r}
set.seed(seeds$rarefy)
raref_both <- rarefy(counts_both)  # 256 species
raref_a <- rarefy(counts_a)        # 214 species
raref_b <- rarefy(counts_b)        # 231 species


# curious about presence after rarefying
presence_a <- return.presence(raref_a, colnames(raref_a))
presence_b <- return.presence(raref_b, colnames(raref_b))
presence_both <- return.presence(raref_both, colnames(raref_both))

hist(presence_a)
hist(presence_b)
hist(presence_both)
```



### Alpha Diversity
Calculate per-sample Alpha Diversity indices at the species level
    *Shannon Index (richness + evenness)*
    *Berger-Parker Dominance*
    *Observed features (richness)*
```{r}
alph_a <- alpha_div(raref_a)
alph_b <- alpha_div(raref_b)

alph_a
alph_b
```

### Get cross-sectional information on each microbe
i.e. presence across treatments, within treatment groups
     for use during filtering
```{r}
a_bug_infos <- list(
  "presence_df" = counts_a %>%
    mutate(across(everything(), ~ . != 0)) %>%
    rownames_to_column("seq_name"),

  "per_sample_presence" = counts_a %>%
    mutate(across(everything(), ~ . != 0)) %>%
    {rowSums(.) / ncol(.)}
)

a_bug_infos$presence_by_batch <- meta_a %>%
  transmute(seq_name, Round_main) %>%
  left_join(a_bug_infos$presence_df, by = "seq_name") %>%
  column_to_rownames("seq_name") %>%
  presence_by_category(     # defined in shannon_div_functions
    data = .,
    category_col = "Round_main"
  )

a_bug_infos$info_df <- a_bug_infos$presence_df %>%
  dplyr::select(-seq_name) %>%
  summarize(across(everything(), ~ sum(.x) / n())) %>% # calculate total presence
  pivot_longer(
    everything(),
    names_to = "microbe",
    values_to = "pres_total"
  ) %>%
  left_join(a_bug_infos$presence_by_batch, by = "microbe") %>%
  rename(
    pres_batch1 = `1`,
    pres_batch3 = `3`
  )
```

```{r}
b_bug_infos <- list(
  "presence_df" = counts_b %>%
    mutate(across(everything(), ~ . != 0)) %>%
    rownames_to_column("seq_name"),

  "per_sample_presence" = counts_b %>%
    mutate(across(everything(), ~ . != 0)) %>%
    {rowSums(.) / ncol(.)}
)

b_bug_infos$presence_by_batch <- meta_b %>%
  transmute(seq_name, Round_main) %>%
  left_join(b_bug_infos$presence_df, by = "seq_name") %>%
  column_to_rownames("seq_name") %>%
  presence_by_category(     # defined in shannon_div_functions
    data = .,
    category_col = "Round_main"
  )

b_bug_infos$info_df <- b_bug_infos$presence_df %>%
  dplyr::select(-seq_name) %>%
  summarize(across(everything(), ~ sum(.x) / n())) %>% # calculate total presence
  pivot_longer(
    everything(),
    names_to = "microbe",
    values_to = "pres_total"
  ) %>%
  left_join(b_bug_infos$presence_by_batch, by = "microbe") %>%
  rename(
    pres_batch4 = `4`,
    pres_batch5 = `5`,
    pres_batch6 = `6`,
    pres_batch7 = `7`,
    pres_batch8 = `8`,
    pres_batch9 = `9`
  )

b_bug_infos$info_df
```

### Do CLR
```{r}
a_clr <- filter_and_clr(
  bug_info = a_bug_infos$info_df,
  counts = counts_a,
  total_presence = 0.3,
  batch_presence = 0.1,
  batch_mode = "all_batches",
  impute_method = "GBM",
  filt_before_clr = TRUE,
  scale_clr = FALSE
)

a_features <- join_bugs_and_div(
  raw_clr_df = a_clr$clr_df_raw,
  raw_alpha_df = alph_a
)

a_features
```


  
`a_features` is now very similar to `"../data/processed/A_clr_df_withalpha_0206.rds"`, except it doesn't have metadata or cytokine PCs yet.

```{r}
b_clr <- filter_and_clr(
  bug_info = b_bug_infos$info_df,
  counts = counts_b,
  total_presence = 0.30,
  batch_presence = 0.1,
  batch_mode = "min_batches",
  min_batches = 2,
  impute_method = "GBM",
  filt_before_clr = TRUE,
  scale_clr = TRUE
)

b_features <- join_bugs_and_div(
  raw_clr_df = b_clr$clr_df_raw,
  raw_alpha_df = alph_b
)

b_features
```


# SAVE
```{r}
a_species_data <- list(
  "meta"           = meta_a_withPCs, # contains animal_code and seq_id
  "counts"         = counts_a,       # row.names = seq_id
  "counts_raref"   = raref_a,
  "relative"       = relabund_a,
  "presence"       = a_bug_infos$presence_df,
  "presence_stats" = a_bug_infos$info_df,
  "clr_alpha"      = a_features$raw_feat,
  "clr_norm"       = a_features$norm_feat,
  "taxonomy_map"   = full_taxonomy,
  "PCA_loadings"   = pca_loadings,
  "PCA_variance"   = ck_pc_df,
  "PCA_raw"        = res_A$pca
)

b_species_data <- list(
  "meta"           = meta_b_smaller, # contains animal_code and seq_id
  "counts"         = counts_b,       # row.names = seq_id
  "counts_raref"   = raref_b,
  "relative"       = relabund_b,
  "presence"       = b_bug_infos$presence_df,
  "presence_stats" = b_bug_infos$info_df,
  "clr_alpha"      = b_features$raw_feat,
  "clr_norm"       = b_features$norm_feat,
  "taxonomy_map"   = full_taxonomy
)
```

Save
```{r, eval = FALSE}
save_output(a_species_data, "meta_A", "data/intermediate")
save_output(b_species_data, "meta_B", "data/intermediate")
```