---
title: "01_preprocess.Rmd"
author: "Alyssa N. Easton"
date: '2026-05-13'
output:
---

Outputs: `data/1_preprocessed/01_preprocess_metadata_cleaned_meta.rds`

```{r "setup", include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
knitr::opts_knit$set(root.dir = "/proj/gibbons/2024_easton_vns/data")
options(digits = 10) ## Formats output to 10 digits
library(tidyverse)
library(data.table)
library(magrittr)
library(readxl)
source("/proj/gibbons/2024_easton_vns/workflow/functions/utils.R")
```

## Prepare Metadata
```{r read}
# updated behavior data
pd_data <- read_xlsx("metadata/VNS_behavior_250314.xlsx") %>%
  dplyr::select(animal_code, startle_habituation, startle_PPI, PD_choice,
                EtOH_24h_intakeave, EtOH_pref_24h, EtOH_8pm_intakeave) %>%
  rename(PD_choice_filt = PD_choice,
         EtOH_24h_intakeave_final = EtOH_24h_intakeave,
         EtOH_pref_24h_final = EtOH_pref_24h,
         EtOH_8pm_intakeave_final = EtOH_8pm_intakeave)

# rest of data
metadata <- read_xlsx("metadata/VNS_datafinalpaper_240513.xlsx") %>%
  dplyr::select(where(~ any(!is.na(.)))) %>%           # remove empty cols
  mutate(animal_code = as.integer(animal_code)) %>%    # convert animal_code to int
  dplyr::select(-matches("p__|o__")) %>%               # remove old microbiome-related columns
  dplyr::select(-c(observed, chao1, shannon, PC1, PC2)) %>%
  left_join(pd_data, by = "animal_code") %>%           # merge in the behavior data
  mutate(
    # make Injury into factor, capitalize
    Injury = factor(Injury, levels = c("sham","blast")) %>%
      fct_recode("Sham" = "sham", "Blast" = "blast"),
    # make Treatment into factor, capitalize
    Treatment = factor(Treatment, levels = c("sham","VNS")) %>%
      fct_recode("Sham" = "sham", "VNS" = "VNS"),
    # make Round_main into factor
    Round_main = factor(Round_main, levels = c("1", "2", "3", "4", "5", "6", "7", "8", "9")),
    # make Cage id into factor
    Cage = factor(Cage),
    # add a feature called "cohort", which is either A or B
    cohort = case_when(
      feces_tp == 0.16 ~ "A",
      feces_tp == 1 ~ "B"),
    cohort = factor(cohort),
    # add a feature for seq_name, which is the ID listed on all sequencing files
    seq_name = case_when(
      animal_code == 1774 ~ 1775L,
      animal_code == 1363 ~ 1362L,
      animal_code == 1372 ~ 1377L,
      TRUE ~ animal_code)
    ) %>%
  # sort table by the SEQUENCING ID
  arrange(seq_name) %>%
  # replace special characters for later processing, because I'm not sure how
  # they are handled in Linux shell scripts
  rename(
    c("serum_IL-1a" = `serum_IL-1α`,
      "brain_IL-1a" = `brain_IL-1α`,
      "brain_INF_gamma" = `brain_INFγ`)
  ) %>%
  # add a feature called "inj_trt_group" which has our 4 treatment groups
  mutate(
    inj_trt_group = factor(
      paste0(.$Injury, " + ", .$Treatment),
      levels = c("Sham + Sham", "Blast + Sham", "Sham + VNS", "Blast + VNS")
      ),
    blast         = ifelse(Injury    == "Blast",                   1, 0),
    vns           = ifelse(Treatment == "VNS",                     1, 0),
    blast_and_vns = ifelse(Injury == "Blast" & Treatment == "VNS", 1, 0),
    seq_name      = as.character(seq_name),
  )
```

```{r}
clean_meta <- function(meta) {
  meta %>%
    mutate(Round_main    = droplevels(Round_main)) %>%
    dplyr::select(where(~ !all(is.na(.)))) %>%  # same as meta <- meta[, colSums(is.na(meta)) != nrow(meta)]
    relocate(c("cohort", "inj_trt_group"), .before = 3) %>%
    relocate(c("animal_code", "seq_name"), .before = 1) %>%
    dplyr::select(-c("cohort", "file_name", "feces_tp"))
}

meta_a <- clean_meta(filter(metadata, cohort == "A"))
meta_b <- clean_meta(filter(metadata, cohort == "B"))

meta_b_smaller <- meta_b %>%
  dplyr::select(-any_of(extra_B_cols))
```


Note: The extra mouse was 1444 (sham-sham). I accidentally used `EtOH_pref_24h` instead of `EtOH_pref_24h_final`. 

Save
```{r, eval = FALSE}
# Old way:
# saveRDS(metadata, "1_preprocessed/metadata.rds")

# New organized way:
save_output(metadata, "cleaned_meta", "data/1_preprocessed")
```


# Preprocess Microbiome data
```{r}
seq_a <- metadata %>% filter(cohort == "A") %>% pull(seq_name)
seq_b <- metadata %>% filter(cohort == "B") %>% pull(seq_name)
```

Preprocess new MetaPhlAn output myself - get *species-level count table*
```{r}
files <- list.files("/proj/gibbons/2024_easton_vns/metaphlan_2/metaphlan", 
                    pattern = "_metaphlan\\.txt$", full.names = TRUE)

dt <- rbindlist(lapply(files, function(f) {
  fread(f)[, seq_name := sub("_metaphlan\\.txt$", "", basename(f))]
}), fill = TRUE) %>%
  rename(
    clade_name = `#clade_name`,
    est_n_reads = estimated_number_of_reads_from_the_clade
    ) %>%
  filter(seq_name %in% metadata$seq_name)

s_long <- dt %>%                 # we can do this for any taxonomic level
  filter(
    grepl("\\|s__", clade_name) | clade_name == "UNCLASSIFIED",
    !grepl("\\|t__", clade_name)
  ) %>%
  mutate(
    clade_name = ifelse(
      clade_name == "UNCLASSIFIED",
      "k__unclassified|p__unclassified|c__unclassified|o__unclassified|f__unclassified|g__unclassified|s__unclassified",
      clade_name
      )
  )

full_taxonomy <- data.frame(full_name = unique(s_long$clade_name)) %>%
  separate(
    full_name,
    sep = "\\|",
    into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"), # no "Strain"
    fill = "right",
    remove = FALSE
  ) %>%
  mutate(
    Kingdom = gsub("k__", "", Kingdom),
    Phylum  = gsub("p__", "", Phylum),
    Class   = gsub("c__", "", Class),
    Order   = gsub("o__", "", Order),
    Family  = gsub("f__", "", Family),
    Genus   = gsub("g__", "", Genus),
    Species = gsub("s__", "", Species)
  )

# check that all species names are still unique
length(unique(full_taxonomy$Species)) == length(unique(full_taxonomy$full_name))

s_long <- s_long %>%
  left_join(
    full_taxonomy %>%
      dplyr::select(full_name, Species),
    by = c("clade_name" = "full_name")
    ) %>%
  mutate(
    clade_name = coalesce(Species, clade_name)
    ) %>%
  dplyr::select(-Species)

s_counts <- s_long %>%
  pivot_wider(
    ., id_cols = "seq_name", names_from = "clade_name", values_from = "est_n_reads"
    ) %>%
  dplyr::select(-unclassified)   # remove because some values are negative

s_relabund <- s_long %>%
  pivot_wider(., id_cols = "seq_name", names_from = "clade_name", values_from = "relative_abundance")

```

Rarefy count table to lowest n_reads (for each cohort)
  cohort A worst depth is sample *1413* (575,229 reads), followed by *1419*
  cohort B worst depth is sample *1766* (154,195 reads), followed by *1765* <- keep an eye on this 1766 - might need to drop?
  
```{r}
counts_both <- s_counts %>%
  filter(seq_name %in% c(seq_a, seq_b)) %>%
  dplyr::select(where(~ !all(is.na(.)))) %>%
  mutate(across(everything(), ~replace_na(., 0))) %>%
  column_to_rownames("seq_name")

counts_a <- s_counts %>%
  filter(seq_name %in% seq_a) %>%
  dplyr::select(where(~ !all(is.na(.)))) %>%
  mutate(across(everything(), ~replace_na(., 0))) %>%
  column_to_rownames("seq_name")

counts_b <- s_counts %>%
  filter(seq_name %in% seq_b) %>%
  dplyr::select(where(~ !all(is.na(.)))) %>%
  mutate(across(everything(), ~replace_na(., 0))) %>%
  column_to_rownames("seq_name")

# also just go ahead and separate the relative abundance into a and b
relabund_a <- s_relabund %>%
  filter(seq_name %in% seq_a) %>%
  dplyr::select(where(~ !all(is.na(.))))
relabund_b <- s_relabund %>%
  filter(seq_name %in% seq_b) %>%
  dplyr::select(where(~ !all(is.na(.))))
```

## Do rarefaction by hand
(this uses my own function `rarefy`, which is in `shannon_div_functions.R`)
```{r}
set.seed(145)
raref_both <- rarefy(counts_both)  # 256 species
raref_a <- rarefy(counts_a)        # 214 species
raref_b <- rarefy(counts_b)        # 231 species


# curious about presence after rarefying
presence_a <- return.presence(raref_a, colnames(raref_a))
presence_b <- return.presence(raref_b, colnames(raref_b))
presence_both <- return.presence(raref_both, colnames(raref_both))

hist(presence_a)
hist(presence_b)
hist(presence_both)
```



## Alpha Diversity Metrics
Calculate per-sample Alpha Diversity indices:
    *Shannon Index (richness + evenness), Berger-Parker Dominance, Observed features (richness)*
```{r}
alph_a <- alpha_div(raref_a)
alph_b <- alpha_div(raref_b)

alph_a
alph_b
```

## Get cross-sectional information on each microbe
 (presence across treatments, within treatment groups)
```{r}
a_bug_infos <- list(
  "presence_df" = counts_a %>%
    mutate(across(everything(), ~ . != 0)) %>%
    rownames_to_column("seq_name"),

  "per_sample_presence" = counts_a %>%
    mutate(across(everything(), ~ . != 0)) %>%
    {rowSums(.) / ncol(.)}
)

a_bug_infos$presence_by_batch <- meta_a %>%
  transmute(seq_name, Round_main) %>%
  left_join(a_bug_infos$presence_df, by = "seq_name") %>%
  column_to_rownames("seq_name") %>%
  presence_by_category(     # defined in shannon_div_functions
    data = .,
    category_col = "Round_main"
  )

a_bug_infos$info_df <- a_bug_infos$presence_df %>%
  dplyr::select(-seq_name) %>%
  summarize(across(everything(), ~ sum(.x) / n())) %>% # calculate total presence
  pivot_longer(
    everything(),
    names_to = "microbe",
    values_to = "pres_total"
  ) %>%
  left_join(a_bug_infos$presence_by_batch, by = "microbe") %>%
  rename(
    pres_batch1 = `1`,
    pres_batch3 = `3`
  )

```


## Processing
```{r}
a_clr <- filter_and_clr(
  bug_info = a_bug_infos$info_df,
  counts = counts_a,
  total_presence = 0.3,
  batch_presence = 0.1,
  impute_method = "GBM",
  filt_before_clr = TRUE,
  ctrl_batch = TRUE
)

a_features <- join_bugs_and_div(
  raw_clr_df = a_clr$clr_df_raw,
  raw_alpha_df = alph_a
)

a_features
```

Merge all data into one df 
  Treatments (exposures) are binary
  Microbes are filtered, CLR-transformed, and then z-scored
  Outcome variables are z-scored
  *Alpha diversity is z-scored!*
  
`a_features` is now very similar to `"../data/processed/A_clr_df_withalpha_0206.rds"`, except it doesn't have metadata or cytokine PCs yet.