---
title: "alphadiv_final.Rmd"
author: "Alyssa N. Easton"
date: '2025-09-29'
output: 
---

Originally `shannon_div.Rmd`, but had to re-run MetaPhlAn to get estimated counts,
as n_reads was associated with Shannon diversity.

Questions:
1. Across the metadata, what is associated with alpha diversity?
2. Does treatment affect alpha diversity?
3. Is treatment's effect on alpha diversity confounded by any variables (batch, weight)?

## Setup
```{r "setup", include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
knitr::opts_knit$set(root.dir = file.path(Sys.getenv("VNS_ROOT"), "data"))  # stored VNS_ROOT in ~/.Renviron 
options(digits = 10) ## Formats output to 10 digits
library(tidyverse)
library(data.table)
library(magrittr)
library(patchwork)
library(stats)
library(ggsignif)
library(lme4)
library(viridis)
library(pheatmap)
library(broom)
library(janitor)
library(openxlsx)
library(zCompositions)
library(compositions)
library(nortest)
source("/proj/gibbons/2024_easton_vns/scripts/shannon_div_functions.R")

```

## Import metadata
```{r}
# sort by sequencing id
metadata <- readRDS("1_preprocessed/metadata.rds") %>%
  relocate(seq_name, .before = 6) %>%
  arrange(seq_name)

seq_a <- metadata %>% filter(cohort == "A") %>% pull(seq_name)
seq_b <- metadata %>% filter(cohort == "B") %>% pull(seq_name)

# species-level count table

# take a random sub-sample of X counts (find lowest total counts/sample)

# calculate Shannon diversity, species observed, and another dominance metric
```

# Process Microbiome data
## Import + Preprocess Taxonomic Counts
Preprocess new MetaPhlAn output myself - get *species-level count table*
```{r}
files <- list.files("/proj/gibbons/2024_easton_vns/metaphlan_2/metaphlan", 
                    pattern = "_metaphlan\\.txt$", full.names = TRUE)

dt <- rbindlist(lapply(files, function(f) {
  fread(f)[, seq_name := sub("_metaphlan\\.txt$", "", basename(f))]
}), fill = TRUE) %>%
  rename(
    clade_name = `#clade_name`,
    est_n_reads = estimated_number_of_reads_from_the_clade
    ) %>%
  filter(seq_name %in% metadata$seq_name)

s_long <- dt %>%                 # we can do this for any taxonomic level
  filter(
    grepl("\\|s__", clade_name) | clade_name == "UNCLASSIFIED",
    !grepl("\\|t__", clade_name)
  ) %>%
  mutate(
    clade_name = ifelse(
      clade_name == "UNCLASSIFIED",
      "k__unclassified|p__unclassified|c__unclassified|o__unclassified|f__unclassified|g__unclassified|s__unclassified",
      clade_name
      )
  )

full_taxonomy <- data.frame(full_name = unique(s_long$clade_name)) %>%
  separate(
    full_name,
    sep = "\\|",
    into = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"), # no "Strain"
    fill = "right",
    remove = FALSE
  ) %>%
  mutate(
    Kingdom = gsub("k__", "", Kingdom),
    Phylum  = gsub("p__", "", Phylum),
    Class   = gsub("c__", "", Class),
    Order   = gsub("o__", "", Order),
    Family  = gsub("f__", "", Family),
    Genus   = gsub("g__", "", Genus),
    Species = gsub("s__", "", Species)
  )

# check that all species names are still unique
length(unique(full_taxonomy$Species)) == length(unique(full_taxonomy$full_name))

s_long <- s_long %>%
  left_join(
    full_taxonomy %>%
      dplyr::select(full_name, Species),
    by = c("clade_name" = "full_name")
    ) %>%
  mutate(
    clade_name = coalesce(Species, clade_name)
    ) %>%
  dplyr::select(-Species)

s_counts <- s_long %>%
  pivot_wider(
    ., id_cols = "seq_name", names_from = "clade_name", values_from = "est_n_reads"
    ) %>%
  dplyr::select(-unclassified)   # remove because some values are negative

s_relabund <- s_long %>%
  pivot_wider(., id_cols = "seq_name", names_from = "clade_name", values_from = "relative_abundance")

```

Rarefy count table to lowest n_reads (for each cohort)
  cohort A worst depth is sample *1413* (575,229 reads), followed by *1419*
  cohort B worst depth is sample *1766* (154,195 reads), followed by *1765* <- keep an eye on this 1766 - might need to drop?
```{r}
counts_both <- s_counts %>%
  filter(seq_name %in% c(seq_a, seq_b)) %>%
  dplyr::select(where(~ !all(is.na(.)))) %>%
  mutate(across(everything(), ~replace_na(., 0))) %>%
  column_to_rownames("seq_name")

counts_a <- s_counts %>%
  filter(seq_name %in% seq_a) %>%
  dplyr::select(where(~ !all(is.na(.)))) %>%
  mutate(across(everything(), ~replace_na(., 0))) %>%
  column_to_rownames("seq_name")

counts_b <- s_counts %>%
  filter(seq_name %in% seq_b) %>%
  dplyr::select(where(~ !all(is.na(.)))) %>%
  mutate(across(everything(), ~replace_na(., 0))) %>%
  column_to_rownames("seq_name")

# also just go ahead and separate the relative abundance into a and b
relabund_a <- s_relabund %>%
  filter(seq_name %in% seq_a) %>%
  dplyr::select(where(~ !all(is.na(.))))
relabund_b <- s_relabund %>%
  filter(seq_name %in% seq_b) %>%
  dplyr::select(where(~ !all(is.na(.))))
```

## Do rarefaction by hand
(this uses my own function `rarefy`, which is in `shannon_div_functions.R`)
```{r}
set.seed(145)
raref_both <- rarefy(counts_both)  # 256 species
raref_a <- rarefy(counts_a)        # 214 species
raref_b <- rarefy(counts_b)        # 231 species
```

Just curious about presence after rarefying
```{r}
presence_a <- return.presence(raref_a, colnames(raref_a))
presence_b <- return.presence(raref_b, colnames(raref_b))
presence_both <- return.presence(raref_both, colnames(raref_both))

hist(presence_a)
hist(presence_b)
hist(presence_both)
```


## Clean metadata
```{r}

clean_meta <- function(meta) {
  meta %>%
    mutate(
      Round_main    = droplevels(Round_main),
      blast         = ifelse(Injury    == "Blast",                   1, 0),
      vns           = ifelse(Treatment == "VNS",                     1, 0),
      blast_and_vns = ifelse(Injury == "Blast" & Treatment == "VNS", 1, 0),
      seq_name      = as.character(seq_name),
      inj_trt_group = fct_relevel(
        inj_trt_group, "Sham + Sham", "Blast + Sham", "Sham + VNS", "Blast + VNS"
        )
      ) %>%
    dplyr::select(where(~ !all(is.na(.)))) %>%  # same as meta <- meta[, colSums(is.na(meta)) != nrow(meta)]
    relocate(c("cohort", "inj_trt_group"), .before = 3) %>%
    relocate(c("animal_code", "seq_name"), .before = 1) %>%
    dplyr::select(-c("cohort", "file_name", "feces_tp"))
}

meta_a <- clean_meta(filter(metadata, cohort == "A"))
meta_b <- clean_meta(filter(metadata, cohort == "B"))

meta_b_smaller <- meta_b %>%
  dplyr::select(-c(
    LORR_1, LORR_2, LORR_3,
    ppi_mV_max_2, ppi_mV_max_10, ppi_mV_max_20, ppi_mV_max_50, ppi_mV_max_100, ppi_mV_max_200, ppi_mV_max_500, ppi_mV_max_1000,
    PD_0perc, PD_25perc, PD_50perc, PD_75perc, PD_100perc,
    PD_0RHElate, PD_25RHElate, PD_50RHElate, PD_75RHElate, PD_100RHElate,
    PD_0HEITI, PD_25HEITI, PD_50HEITI, PD_75HEITI, PD_100HEITI,
    EtOH_24h_1, EtOH_24h_2, EtOH_24h_3, EtOH_24h_4, EtOH_24h_5, EtOH_24h_6, EtOH_24h_7, EtOH_24h_8, EtOH_24h_9, EtOH_24h_10,
    EtOH_24h_11, EtOH_24h_12, EtOH_pref_24h_1, EtOH_pref_24h_2, EtOH_pref_24h_3, EtOH_pref_24h_4, EtOH_pref_24h_5,
    EtOH_pref_24h_6, EtOH_pref_24h_7, EtOH_pref_24h_8, EtOH_pref_24h_9, EtOH_pref_24h_10, EtOH_pref_24h_11, EtOH_pref_24h_12,
    EtOH_24h, EtOH_pref_24h, EtOH_8pm
  ))
```

# Tried filtering before CLR transform - it didn't change much
Maybe I should CLR-transform here just for the sake of it... ?
```{r}
# thresh    <- 0.25
# 
# counts    <- counts_a
# pres      <- counts > 0
# pct_pres  <- colSums(pres) / nrow(pres)
# good_taxa <- names(pct_pres)[pct_pres >= thresh] # 138 "good" taxa
# filt_counts  <- counts[, good_taxa]
# cm           <- cmultRepl(filt_counts, label = 0, method = "CZM", z.delete = FALSE)   # or method="GBM"
# clr_mat      <- t(apply(cm, 1, compositions::clr))
# 
# clr.transform <- function(count_mat, thresh = 0.25, method = "CZM") {
#   pres         <- count_mat > 0
#   pct_pres     <- colSums(pres) / nrow(pres)
#   good_taxa    <- names(pct_pres)[pct_pres >= thresh] # 138 "good" taxa
#   filt_counts  <- count_mat[, good_taxa]
#   cm           <- cmultRepl(filt_counts, label = 0, method = method, z.delete = FALSE)   # or method="GBM"
#   clr_mat      <- t(apply(cm, 1, compositions::clr))
#   return(clr_mat)
# }
# 
# # how many do these samples have ? they have some info but not much for whatever reason ...
# filt_counts["1366", filt_counts["1366", ] != 0]  # only 19 taxa observed/annotated
# filt_counts["1419", filt_counts["1419", ] != 0]  # only 26 taxa observed/annotated
```

## Try doing CLR + multiplicative replacement 

```{r}
# find_best_threshold <- function(count_mat, thresholds = seq(0.1, 0.5, by = 0.05),
#                                 methods = c("CZM", "GBM")) {
#   
#   results <- lapply(thresholds, function(thresh) {
#     pres         <- count_mat > 0
#     pct_pres     <- colSums(pres) / nrow(pres)
#     good_taxa    <- names(pct_pres)[pct_pres >= thresh]
#     if (length(good_taxa) < 2) return(NULL)  # skip if too few taxa
#     
#     filt_counts  <- count_mat[, good_taxa, drop = FALSE]
#     
#     method_results <- lapply(methods, function(m) {
#       cm <- cmultRepl(filt_counts, label = 0, method = m, z.delete = FALSE)
#       clr_mat <- t(apply(cm, 1, clr))
#       
#       # ---- per-feature AD tests ----
#       ad_stats <- apply(clr_mat, 2, function(feature_vals) {
#         # AD test on this taxon's clr distribution across samples
#         test <- tryCatch(ad.test(feature_vals), error = function(e) NULL)
#         if (is.null(test)) return(NA)
#         test$statistic
#       })
#       
#       # average statistic across taxa
#       mean_ad <- mean(ad_stats, na.rm = TRUE)
#       
#       data.frame(
#         threshold = thresh,
#         method = m,
#         n_taxa = ncol(clr_mat),
#         mean_ad_stat = mean_ad
#       )
#     })
#     
#     do.call(rbind, method_results)
#   })
#   
#   out <- do.call(rbind, results)
#   
#   # best threshold per method = lowest average AD statistic
#   best <- out %>%
#     group_by(method) %>%
#     slice_min(mean_ad_stat, n = 1, with_ties = FALSE) %>%
#     ungroup()
#   
#   list(results = out, best = best)
# }
# 
# # Example usage:
# res_a <- find_best_threshold(
#   counts_a,
#   thresholds = seq(0.1, 0.5, by = 0.05),
#   methods = c("CZM", "GBM")
#   )
# 
# res_b <- find_best_threshold(
#   counts_b,
#   thresholds = seq(0.1, 0.5, by = 0.05),
#   methods = c("CZM", "GBM")
# )
# 
# res_a$best
# res_b$best
```

```{r}
# clr_a <- clr.transform(counts_a, thresh = 0.25, method = "GBM")
# clr_b <- clr.transform(counts_b, thresh = 0.20, method = "GBM")
# 
# # Look through AD stats just because I'm curious for clr B
# ad_stats <- apply(clr_b, 2, function(feature_vals) {
#   # AD test on this taxon's clr distribution across samples
#   test <- tryCatch(ad.test(feature_vals), error = function(e) NULL)
#   if (is.null(test)) return(NA)
#   test$statistic
# })
# 
# # look at one of the worst ones (AD = 4.36)
# hist(clr_b[, "Adlercreutzia_mucosicola"])
# hist(clr_b[, "GGB31822_SGB45198"]) # this is the very worst (AD = 5.5)
```
meh, not horrible ...
# Redo clr the way I did it originallly (filtering afterwards)
```{r}
clr.transform <- function(relab, thresh = 0.25) {
  relab    <- relab %>%   # already has zero cols removed
    column_to_rownames("seq_name") %>%
    mutate(across(everything(), ~replace_na(., 0)/100))
  pres      <- relab > 0
  pct_pres  <- colSums(pres) / nrow(pres)
  good_taxa <- names(pct_pres)[pct_pres >= thresh]
  relab     <- relab + 1e-8                        # add a pseudocount
  clr_mat   <- relab %>%
    apply(1, function(row) {
      log(row / exp(mean(log(row))))
    }) %>%
    t() %>%
    as.data.frame() %>%
    setNames(colnames(relab))
  clr_filt <- clr_mat[, good_taxa]
  return(clr_filt)
}

clr_a <- clr.transform(relabund_a, 0.25)
clr_b <- clr.transform(relabund_b, 0.25)
```



# Export some of this preprocessed data
```{r}
species_processed <- list(
  # main outputs for A
  a_counts      = counts_a,
  a_counts_rare = raref_a,
  a_meta        = meta_a,
  a_relabund    = relabund_a,
  a_clr         = clr_a,
  # main outputs for B
  b_counts      = counts_b,
  b_counts_rare = raref_b,
  b_meta        = meta_b_smaller,
  b_relabund    = relabund_b,
  b_clr         = clr_b,
  # other useful info
  taxonomy_map  = full_taxonomy,
  all_counts    = s_counts,
  all_abund     = s_relabund,
  all_raref     = raref_both
)

saveRDS(species_processed, "/proj/gibbons/2024_easton_vns/data/1_preprocessed/species_october.rds")
```



# Alpha Diversity
## Calculate per-sample Alpha Diversity indices:
    *Shannon Index (richness + evenness), Berger-Parker Dominance, Observed features (richness)*
```{r}
alpha_div <- function(raref_counts) {
  shannon    <- vegan::diversity(raref_counts, index = "shannon", MARGIN = 1) # row-wise
  n_feats    <- rowSums(raref_counts >= 1)
  simpson    <- vegan::diversity(raref_counts, index = "simpson", MARGIN = 1)
  invsimpson <- vegan::diversity(raref_counts, index = "invsimpson")
  richness   <- rowSums(raref_counts > 0)
  pielou     <- shannon / log(richness)
  berg_park  <- apply(raref_counts, 1, function(x) {
    max(x) / sum(x)
  })
  
  data.frame(
    shannon       = shannon,
    simpson       = simpson,
    invsimpson    = invsimpson,
    richness      = richness,
    pielou        = pielou,
    berger_parker = berg_park
  )
}


alph_a <- alpha_div(raref_a)
alph_b <- alpha_div(raref_b)

alph_a
alph_b

saveRDS(alph_a, file = "/proj/gibbons/2024_easton_vns/data/diversity/A_alpha.rds")
saveRDS(alph_b, file = "/proj/gibbons/2024_easton_vns/data/diversity/B_alpha.rds")
```

## Merge diversity into metadata
```{r}
alpha_div_cols <- colnames(alph_a)

merge_a <- meta_a %>%
  left_join(alph_a %>% rownames_to_column("seq_name"), by = "seq_name") %>%
  relocate(all_of(alpha_div_cols), .before = 9)

merge_b <- meta_b_smaller %>%
  left_join(alph_b %>% rownames_to_column("seq_name"), by = "seq_name") %>%
  relocate(all_of(alpha_div_cols), .before = 9)
```

Testing function w/ signif bars
```{r}
injtrt_pal <- c(
  "Sham + Sham"  = "#1032ff",
  "Sham + VNS"   = "#1032ff",
  "Blast + Sham" = "#ff9200",
  "Blast + VNS"  = "#ff9200"
  )

# Manually assign x-positions of boxes
pos_map <- c(                # custom spacing
  "Sham + Sham"  = 4/30,
  "Blast + Sham" = 10/30,
  "Sham + VNS"   = 20/30,
  "Blast + VNS"  = 26/30
  )
box_wid <- 4/30
```

```{r}
# box_easy <- function(
#     data,
#     dv,
#     y_lab,
#     groups      = "inj_trt_group",
#     palette     = injtrt_pal,
#     x_positions = c(4,10,20,26),
#     box_width   = 4,
#     axis_length = 30,
#     show_legend = FALSE
#     ) {
#   # --- Rescale positions and widths to proportions ---
#   pos_map <- x_positions / axis_length
#   names(pos_map) <- levels(factor(data[[groups]]))  # match group order
#   box_width <- box_width / axis_length              # rescale width
#   error_width <- box_width/2
#   
#   # assign numeric xpos to each row
#   data$xpos <- pos_map[as.character(data[[groups]])]
# 
#   # calculate base ymax and step for significance bar stacking
#   dv_vals <- data[[dv]]
#   base    <- max(dv_vals, na.rm = TRUE)
#   step    <- 0.2 * diff(range(dv_vals, na.rm = TRUE))  # spacing between bars
#   y_max   <- base + 4 * step
#   
#   ggplot(data, aes(x = xpos, y = .data[[dv]], group = .data[[groups]], fill = .data[[groups]])) +
#     geom_boxplot(
#       width    = box_width,
#       color    = "grey",
#       fill     = "transparent",
#       outliers = FALSE,
#       coef     = 0
#     ) +
#     stat_summary(
#       geom = "errorbar",
#       fun.data = "mean_cl_boot",
#       width = error_width
#     ) +
#     geom_point(
#       position = position_jitter(width = box_width/4, height = 0), # jitter relative to box width
#       shape    = 22,
#       size     = 5,
#       aes(fill = .data[[groups]]),
#       color    = "black",
#       stroke   = 1.2
#     ) +
#     scale_x_continuous(               # x-axis scaling/locations
#       limits = c(0, 1),
#       breaks = pos_map,
#       labels = names(pos_map)
#     ) +
#     # scale_y_continuous(
#     #   limits = c(min(dv_vals, na.rm = TRUE), y_max)
#     # ) +
#     scale_fill_manual(values = palette) +
#     coord_fixed(
#       ratio = 1,
#       clip = "off"
#       ) +
#     labs(x = NULL, y = y_lab, fill = NULL) +
#     # geom_signif(
#     #   comparisons = list(
#     #     c("Sham + Sham" , "Blast + Sham"),
#     #     c("Sham + VNS"  , "Blast + VNS"),
#     #     c("Sham + Sham" , "Sham + VNS"),
#     #     c("Blast + Sham", "Blast + VNS")
#     #   ),
#     #   map_signif_level = FALSE, 
#     #   annotations      = rep(" ", 4),
#     #   y_position       = base + c(1,1,2,3) * step,
#     #   tip_length       = 0.03
#     # ) +
#     theme_classic() +
#     theme(
#       legend.justification = "left",
#       legend.direction     = "vertical",
#       legend.background    = element_blank(),
#       legend.box.margin    = margin(0, 0, 10, 0),
#       # plot.margin          = margin(10, 10, 10, 10),
#       aspect.ratio         = 1, # square plotting area
#       axis.title.y         = element_text(size = 33.75),
#       axis.text.x          = element_blank(),                # element_text(size = 20),
#       axis.text.y          = element_text(size = 30),
#       axis.ticks.x         = element_blank(),
#       axis.line            = element_line(colour = "black", linewidth = 1)
#     ) +
#     theme(legend.position = if(!show_legend) "none" else "right")
# }

```
  
## Visualize treatment effects across 4 treatment groups
```{r}
datasets <- list(
  A = merge_a,
  B = merge_b
  )
metrics  <- list(
  shannon       = "Shannon Index",
  simpson       = "Simpson Index", 
  invsimpson    = "Inverse Simpson",
  richness      = "Observed Species",
  pielou        = "Pielou's Evenness",
  berger_parker = "Berger-Parker Index"
  )

walk2(
  rep(names(datasets), each = length(metrics)),
  rep(names(metrics), times = length(datasets)),
  ~ {
    dset <- datasets[[.x]]
    metric <- .y
    y_lab <- metrics[[metric]]
    
    p <- box_easy(data = dset, dv = metric, y_lab = y_lab) + ggtitle(.x)
    ggsave(
      filename = sprintf("../figures/%s_%s_box.svg", .x, metric),
      plot = p,
      width = 5,
      device = "svg"
    )
  }
)

```


    Shannon Diversity
```{r}
shan_a <- box_easy(
  data  = merge_a,
  dv    = "shannon",
  y_lab = "Shannon Index"
  ) + ggtitle("A")

shan_b <- box_easy(
  data  = merge_b,
  dv    = "shannon",
  y_lab = "Shannon Index"
  ) + ggtitle("B")

# ggsave("../figures/AB_shannon_box.png", shan_a + shan_b, width = 11)
# ggsave("../figures/AB_shannon_box.svg", shan_a + shan_b, width = 11, device = "svg")

ggsave("../figures/A_shannon_box.svg", shan_a, width = 5, device = "svg")
ggsave("../figures/B_shannon_box.svg", shan_b, width = 5, device = "svg")

```
Simpson Index
```{r}
simp_a <- box_easy(data = merge_a, dv = "simpson", y_lab = "Simpson Index") + ggtitle("A")
simp_b <- box_easy(data = merge_b, dv = "simpson", y_lab = "Simpson Index") + ggtitle("B")

# ggsave("../figures/AB_simpson_box.png", simp, width = 11)
# ggsave("../figures/AB_simpson_box.svg", simp, width = 11, device = "svg")

ggsave("../figures/A_simpson_box.svg", simp_a, width = 5, device = "svg")
ggsave("../figures/B_simpson_box.svg", simp_b, width = 5, device = "svg")
```

Inverse Simpson
```{r}
# isimp_a <- split.boxplot(merge_a, "invsimpson", two_colors, y_lab = "Inverse Simpson") + ggtitle("A")
# isimp_b <- split.boxplot(merge_b, "invsimpson", two_colors, y_lab = "Inverse Simpson") + ggtitle("B")
# 
# isimp_a
# isimp_b
# ggsave("../figures/AB_invsimpson_box.png", isimp_a + isimp_b)
# 
# isimp_a <- split.barplot(merge_a, "invsimpson", two_colors, y_lab = "Inverse Simpson") + ggtitle("A")
# isimp_b <- split.barplot(merge_b, "invsimpson", two_colors, y_lab = "Inverse Simpson") + ggtitle("B")
# ggsave("../figures/AB_invsimpson_bar.png", isimp_a + isimp_b)

isimp <- (box_easy(data = merge_a, dv = "invsimpson", y_lab = "Inverse Simpson") + ggtitle("A")) +
         (box_easy(data = merge_b, dv = "invsimpson", y_lab = "Inverse Simpson") + ggtitle("B"))

ggsave("../figures/AB_invsimpson_box.png", isimp, width = 11)
ggsave("../figures/AB_invsimpson_box.svg", isimp, width = 11, device = "svg")
```

Observed Species
```{r}
# rich_a <- split.barplot(merge_a, "richness", two_colors, y_lab = "Observed Species") + ggtitle("A")
# rich_b <- split.barplot(merge_b, "richness", two_colors, y_lab = "Observed Species") + ggtitle("B")
# 
# rich_a
# rich_b
# ggsave("../figures/AB_richness_bar.png", rich_a + rich_b)

rich <- (box_easy(data = merge_a, dv = "richness", y_lab = "Observed Species") + ggtitle("A")) +
        (box_easy(data = merge_b, dv = "richness", y_lab = "Observed Species") + ggtitle("B"))

ggsave("../figures/AB_richness_box.png", rich, width = 11)
ggsave("../figures/AB_richness_box.svg", rich, width = 11, device = "svg")
```

Pielou's Evenness
```{r}
# even_a <- split.boxplot(merge_a, "pielou", two_colors, y_lab = "Pielou's Evenness") + ggtitle("A")
# even_b <- split.boxplot(merge_b, "pielou", two_colors, y_lab = "Pielou's Evenness") + ggtitle("B")
# 
# even_a
# even_b
# ggsave("../figures/AB_pielou_box.png", even_a + even_b)

even <- (box_easy(data = merge_a, dv = "pielou", y_lab = "Pielou's Evenness") + ggtitle("A")) +
        (box_easy(data = merge_b, dv = "pielou", y_lab = "Pielou's Evenness") + ggtitle("B"))

ggsave("../figures/AB_evenness_box.png", even, width = 11)
ggsave("../figures/AB_evenness_box.svg", even, width = 11, device = "svg")
```

Berger-Parker Dominance
```{r}
# bpa <- split.barplot(merge_a, "berger_parker", two_colors, y_lab = "Berger-Parker Index") + ggtitle("A")
# bpb <- split.barplot(merge_b, "berger_parker", two_colors, y_lab = "Berger-Parker Index") + ggtitle("B")
# 
# bpa + bpb
# ggsave("../figures/AB_bergerparker_bar.png", bpa + bpb)

bp <- (box_easy(data = merge_a, dv = "berger_parker", y_lab = "Berger-Parker Index") + ggtitle("A")) +
      (box_easy(data = merge_b, dv = "berger_parker", y_lab = "Berger-Parker Index") + ggtitle("B"))

ggsave("../figures/AB_bergerparker_box.png", bp, width = 11)
ggsave("../figures/AB_bergerparker_box.svg", bp, width = 11, device = "svg")
```

### Test for treatment effects with 2-way ANOVA
```{r}
anova_a <- list()
for (col in alpha_div_cols) {
  f <- as.formula(paste(col, "~ Injury * Treatment"))
  anova_a[[col]] <- aov(f, data = merge_a)
}

anova_b <- list()
for (col in alpha_div_cols) {
  f <- as.formula(paste(col, "~ Injury * Treatment"))
  anova_b[[col]] <- aov(f, data = merge_b)
}

# get a df for a single ANOVA
# anova_a_df <- tidy(anova_a[[1]]) %>%
#   mutate(
#     across(
#       c(sumsq, meansq, statistic, p.value),
#       ~ round(.x, 3))
#     )

# get df for ALL ANOVAs
anova_a_df <- map2_dfr(anova_a, names(anova_a), ~ {
  tidy(.x) %>%
    mutate(
      y_variable = .y,
      across(
        c(sumsq, meansq, statistic, p.value),
        ~ round(.x, 3)
      )
    )
}) %>%
  filter(y_variable %in% c("shannon", "richness", "berger_parker"))

anova_b_df <- map2_dfr(anova_b, names(anova_b), ~ {
  tidy(.x) %>%
    mutate(
      y_variable = .y,
      across(
        c(sumsq, meansq, statistic, p.value),
        ~ round(.x, 3)
      )
    )
}) %>%
  filter(y_variable %in% c("shannon", "richness", "berger_parker"))   # I actually only want these 3

anova_a_df
anova_b_df
```

Same results as last time for *Shannon Index*. 
  All Alpha diversity measures are significantly ~ Injury in cohort A
  Only Pielou's Evenness ~ Injury in cohort B

post-hoc pairwise comparisons
```{r}
post_hoc <- function(y_var, data) {
  f <- as.formula(paste(y_var, "~ inj_trt_group"))
  tests <- list(
    `Sham-Sham;Blast-Sham` = t.test(
      f, data = subset(data, inj_trt_group %in% c("Sham + Sham","Blast + Sham"))
    ),
    `Sham-VNS;Blast-VNS` = t.test(
      f, data = subset(data, inj_trt_group %in% c("Sham + VNS","Blast + VNS"))
    ),
    `Sham-Sham;Sham-VNS` = t.test(
      f, data = subset(data, inj_trt_group %in% c("Sham + Sham","Sham + VNS"))
    ),
    `Blast-Sham;Blast-VNS` = t.test(
      f, data = subset(data, inj_trt_group %in% c("Blast + Sham","Blast + VNS"))
    )
  )
  raw_p <- sapply(tests, \(x) x$p.value)
  p.adjust(raw_p, method = "bonferroni")
}

shan_tests_a <- post_hoc("shannon", merge_a)
shan_tests_b <- post_hoc("shannon", merge_b)
rich_tests_a <- post_hoc("richness", merge_a)
rich_tests_b <- post_hoc("richness", merge_b)
berg_tests_a <- post_hoc("berger_parker", merge_a)
berg_tests_b <- post_hoc("berger_parker", merge_b)

a_post_hocs <- data.frame(
  shannon = shan_tests_a,
  n_species = rich_tests_a,
  berger_parker = berg_tests_a
)

b_post_hocs <- data.frame(
  shannon = shan_tests_b,
  n_species = rich_tests_b,
  berger_parker = berg_tests_b
)
```
  


# Save tables to excel
*We have stats in tables* `anova_a_df`, `anova_b_df`, ``
Let's save them to an excel file
```{r}
wb <- createWorkbook()

addWorksheet(wb, "A_anova")
addWorksheet(wb, "A_posthoc")
addWorksheet(wb, "B_anova")
addWorksheet(wb, "B_posthoc")

writeData(wb, "A_anova",   anova_a_df)
writeData(wb, "A_posthoc", a_post_hocs)
writeData(wb, "B_anova",   anova_b_df)
writeData(wb, "B_posthoc", b_post_hocs)

saveWorkbook(wb, file = "final/alpha_div_tests.xlsx", overwrite = FALSE)
```

