---
title: "02_diversity.Rmd"
author: "Alyssa N. Easton"
date: '2025-09-29'
output: 
---

## Inputs
- `data/intermediate/01_meta_A.rds`
- `data/intermediate/01_meta_B.rds`

## Outputs
- 

Had to re-run MetaPhlAn to get estimated counts and rarefy counts myself,
as originally raw (non-rarefied) n_reads was associated with Shannon diversity.

Questions:
1. Across the metadata, what is associated with alpha diversity?
2. Does treatment affect alpha diversity?
3. Is treatment's effect on alpha diversity confounded by any variables (batch, weight)?

## Setup
```{r "setup", include=FALSE}
source(here::here("R", "paths.R"))
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
knitr::opts_knit$set(root.dir = PROJECT_ROOT)
options(digits = 10) ## Formats output to 10 digits
library(tidyverse)
library(data.table)
library(magrittr)
library(patchwork)
library(stats)
library(ggsignif)
library(lme4)
library(viridis)
library(pheatmap)
library(broom)
library(janitor)
library(openxlsx)
library(zCompositions)
library(compositions)
library(nortest)
source(here::here("R", "utils.R"))
source(here::here("R", "cleaning.R"))
source(here::here("R", "microbiome_preprocess.R"))
source(here::here("R", "diversity_utils.R"))
```

## Imports
```{r}
a_data <- readRDS(here::here("data", "intermediate", "01_meta_A.rds"))
b_data <- readRDS(here::here("data", "intermediate", "01_meta_B.rds"))
```


# Alpha Diversity
## Calculate per-sample Alpha Diversity indices:
    *Shannon Index (richness + evenness), Berger-Parker Dominance, Observed features (richness)*
```{r}
alpha_div <- function(raref_counts) {
  shannon    <- vegan::diversity(raref_counts, index = "shannon", MARGIN = 1) # row-wise
  n_feats    <- rowSums(raref_counts >= 1)
  simpson    <- vegan::diversity(raref_counts, index = "simpson", MARGIN = 1)
  invsimpson <- vegan::diversity(raref_counts, index = "invsimpson")
  richness   <- rowSums(raref_counts > 0)
  pielou     <- shannon / log(richness)
  berg_park  <- apply(raref_counts, 1, function(x) {
    max(x) / sum(x)
  })
  
  data.frame(
    shannon       = shannon,
    simpson       = simpson,
    invsimpson    = invsimpson,
    richness      = richness,
    pielou        = pielou,
    berger_parker = berg_park
  )
}


alph_a <- alpha_div(raref_a)
alph_b <- alpha_div(raref_b)

alph_a
alph_b

saveRDS(alph_a, file = "/proj/gibbons/2024_easton_vns/data/diversity/A_alpha.rds")
saveRDS(alph_b, file = "/proj/gibbons/2024_easton_vns/data/diversity/B_alpha.rds")
```

## Merge diversity into metadata
```{r}
alpha_div_cols <- colnames(alph_a)

merge_a <- meta_a %>%
  left_join(alph_a %>% rownames_to_column("seq_name"), by = "seq_name") %>%
  relocate(all_of(alpha_div_cols), .before = 9)

merge_b <- meta_b_smaller %>%
  left_join(alph_b %>% rownames_to_column("seq_name"), by = "seq_name") %>%
  relocate(all_of(alpha_div_cols), .before = 9)
```

Testing function w/ signif bars
```{r}
injtrt_pal <- c(
  "Sham + Sham"  = "#1032ff",
  "Sham + VNS"   = "#1032ff",
  "Blast + Sham" = "#ff9200",
  "Blast + VNS"  = "#ff9200"
  )

# Manually assign x-positions of boxes
pos_map <- c(                # custom spacing
  "Sham + Sham"  = 4/30,
  "Blast + Sham" = 10/30,
  "Sham + VNS"   = 20/30,
  "Blast + VNS"  = 26/30
  )
box_wid <- 4/30
```

```{r}
# box_easy <- function(
#     data,
#     dv,
#     y_lab,
#     groups      = "inj_trt_group",
#     palette     = injtrt_pal,
#     x_positions = c(4,10,20,26),
#     box_width   = 4,
#     axis_length = 30,
#     show_legend = FALSE
#     ) {
#   # --- Rescale positions and widths to proportions ---
#   pos_map <- x_positions / axis_length
#   names(pos_map) <- levels(factor(data[[groups]]))  # match group order
#   box_width <- box_width / axis_length              # rescale width
#   error_width <- box_width/2
#   
#   # assign numeric xpos to each row
#   data$xpos <- pos_map[as.character(data[[groups]])]
# 
#   # calculate base ymax and step for significance bar stacking
#   dv_vals <- data[[dv]]
#   base    <- max(dv_vals, na.rm = TRUE)
#   step    <- 0.2 * diff(range(dv_vals, na.rm = TRUE))  # spacing between bars
#   y_max   <- base + 4 * step
#   
#   ggplot(data, aes(x = xpos, y = .data[[dv]], group = .data[[groups]], fill = .data[[groups]])) +
#     geom_boxplot(
#       width    = box_width,
#       color    = "grey",
#       fill     = "transparent",
#       outliers = FALSE,
#       coef     = 0
#     ) +
#     stat_summary(
#       geom = "errorbar",
#       fun.data = "mean_cl_boot",
#       width = error_width
#     ) +
#     geom_point(
#       position = position_jitter(width = box_width/4, height = 0), # jitter relative to box width
#       shape    = 22,
#       size     = 5,
#       aes(fill = .data[[groups]]),
#       color    = "black",
#       stroke   = 1.2
#     ) +
#     scale_x_continuous(               # x-axis scaling/locations
#       limits = c(0, 1),
#       breaks = pos_map,
#       labels = names(pos_map)
#     ) +
#     # scale_y_continuous(
#     #   limits = c(min(dv_vals, na.rm = TRUE), y_max)
#     # ) +
#     scale_fill_manual(values = palette) +
#     coord_fixed(
#       ratio = 1,
#       clip = "off"
#       ) +
#     labs(x = NULL, y = y_lab, fill = NULL) +
#     # geom_signif(
#     #   comparisons = list(
#     #     c("Sham + Sham" , "Blast + Sham"),
#     #     c("Sham + VNS"  , "Blast + VNS"),
#     #     c("Sham + Sham" , "Sham + VNS"),
#     #     c("Blast + Sham", "Blast + VNS")
#     #   ),
#     #   map_signif_level = FALSE, 
#     #   annotations      = rep(" ", 4),
#     #   y_position       = base + c(1,1,2,3) * step,
#     #   tip_length       = 0.03
#     # ) +
#     theme_classic() +
#     theme(
#       legend.justification = "left",
#       legend.direction     = "vertical",
#       legend.background    = element_blank(),
#       legend.box.margin    = margin(0, 0, 10, 0),
#       # plot.margin          = margin(10, 10, 10, 10),
#       aspect.ratio         = 1, # square plotting area
#       axis.title.y         = element_text(size = 33.75),
#       axis.text.x          = element_blank(),                # element_text(size = 20),
#       axis.text.y          = element_text(size = 30),
#       axis.ticks.x         = element_blank(),
#       axis.line            = element_line(colour = "black", linewidth = 1)
#     ) +
#     theme(legend.position = if(!show_legend) "none" else "right")
# }

```
  
## Visualize treatment effects across 4 treatment groups
```{r}
datasets <- list(
  A = merge_a,
  B = merge_b
  )
metrics  <- list(
  shannon       = "Shannon Index",
  simpson       = "Simpson Index", 
  invsimpson    = "Inverse Simpson",
  richness      = "Observed Species",
  pielou        = "Pielou's Evenness",
  berger_parker = "Berger-Parker Index"
  )

walk2(
  rep(names(datasets), each = length(metrics)),
  rep(names(metrics), times = length(datasets)),
  ~ {
    dset <- datasets[[.x]]
    metric <- .y
    y_lab <- metrics[[metric]]
    
    p <- box_easy(data = dset, dv = metric, y_lab = y_lab) + ggtitle(.x)
    ggsave(
      filename = sprintf("../figures/%s_%s_box.svg", .x, metric),
      plot = p,
      width = 5,
      device = "svg"
    )
  }
)

```


    Shannon Diversity
```{r}
shan_a <- box_easy(
  data  = merge_a,
  dv    = "shannon",
  y_lab = "Shannon Index"
  ) + ggtitle("A")

shan_b <- box_easy(
  data  = merge_b,
  dv    = "shannon",
  y_lab = "Shannon Index"
  ) + ggtitle("B")

# ggsave("../figures/AB_shannon_box.png", shan_a + shan_b, width = 11)
# ggsave("../figures/AB_shannon_box.svg", shan_a + shan_b, width = 11, device = "svg")

ggsave("../figures/A_shannon_box.svg", shan_a, width = 5, device = "svg")
ggsave("../figures/B_shannon_box.svg", shan_b, width = 5, device = "svg")

```
Simpson Index
```{r}
simp_a <- box_easy(data = merge_a, dv = "simpson", y_lab = "Simpson Index") + ggtitle("A")
simp_b <- box_easy(data = merge_b, dv = "simpson", y_lab = "Simpson Index") + ggtitle("B")

# ggsave("../figures/AB_simpson_box.png", simp, width = 11)
# ggsave("../figures/AB_simpson_box.svg", simp, width = 11, device = "svg")

ggsave("../figures/A_simpson_box.svg", simp_a, width = 5, device = "svg")
ggsave("../figures/B_simpson_box.svg", simp_b, width = 5, device = "svg")
```

Inverse Simpson
```{r}
# isimp_a <- split.boxplot(merge_a, "invsimpson", two_colors, y_lab = "Inverse Simpson") + ggtitle("A")
# isimp_b <- split.boxplot(merge_b, "invsimpson", two_colors, y_lab = "Inverse Simpson") + ggtitle("B")
# 
# isimp_a
# isimp_b
# ggsave("../figures/AB_invsimpson_box.png", isimp_a + isimp_b)
# 
# isimp_a <- split.barplot(merge_a, "invsimpson", two_colors, y_lab = "Inverse Simpson") + ggtitle("A")
# isimp_b <- split.barplot(merge_b, "invsimpson", two_colors, y_lab = "Inverse Simpson") + ggtitle("B")
# ggsave("../figures/AB_invsimpson_bar.png", isimp_a + isimp_b)

isimp <- (box_easy(data = merge_a, dv = "invsimpson", y_lab = "Inverse Simpson") + ggtitle("A")) +
         (box_easy(data = merge_b, dv = "invsimpson", y_lab = "Inverse Simpson") + ggtitle("B"))

ggsave("../figures/AB_invsimpson_box.png", isimp, width = 11)
ggsave("../figures/AB_invsimpson_box.svg", isimp, width = 11, device = "svg")
```

Observed Species
```{r}
# rich_a <- split.barplot(merge_a, "richness", two_colors, y_lab = "Observed Species") + ggtitle("A")
# rich_b <- split.barplot(merge_b, "richness", two_colors, y_lab = "Observed Species") + ggtitle("B")
# 
# rich_a
# rich_b
# ggsave("../figures/AB_richness_bar.png", rich_a + rich_b)

rich <- (box_easy(data = merge_a, dv = "richness", y_lab = "Observed Species") + ggtitle("A")) +
        (box_easy(data = merge_b, dv = "richness", y_lab = "Observed Species") + ggtitle("B"))

ggsave("../figures/AB_richness_box.png", rich, width = 11)
ggsave("../figures/AB_richness_box.svg", rich, width = 11, device = "svg")
```

Pielou's Evenness
```{r}
# even_a <- split.boxplot(merge_a, "pielou", two_colors, y_lab = "Pielou's Evenness") + ggtitle("A")
# even_b <- split.boxplot(merge_b, "pielou", two_colors, y_lab = "Pielou's Evenness") + ggtitle("B")
# 
# even_a
# even_b
# ggsave("../figures/AB_pielou_box.png", even_a + even_b)

even <- (box_easy(data = merge_a, dv = "pielou", y_lab = "Pielou's Evenness") + ggtitle("A")) +
        (box_easy(data = merge_b, dv = "pielou", y_lab = "Pielou's Evenness") + ggtitle("B"))

ggsave("../figures/AB_evenness_box.png", even, width = 11)
ggsave("../figures/AB_evenness_box.svg", even, width = 11, device = "svg")
```

Berger-Parker Dominance
```{r}
# bpa <- split.barplot(merge_a, "berger_parker", two_colors, y_lab = "Berger-Parker Index") + ggtitle("A")
# bpb <- split.barplot(merge_b, "berger_parker", two_colors, y_lab = "Berger-Parker Index") + ggtitle("B")
# 
# bpa + bpb
# ggsave("../figures/AB_bergerparker_bar.png", bpa + bpb)

bp <- (box_easy(data = merge_a, dv = "berger_parker", y_lab = "Berger-Parker Index") + ggtitle("A")) +
      (box_easy(data = merge_b, dv = "berger_parker", y_lab = "Berger-Parker Index") + ggtitle("B"))

ggsave("../figures/AB_bergerparker_box.png", bp, width = 11)
ggsave("../figures/AB_bergerparker_box.svg", bp, width = 11, device = "svg")
```

### Test for treatment effects with 2-way ANOVA
```{r}
anova_a <- list()
for (col in alpha_div_cols) {
  f <- as.formula(paste(col, "~ Injury * Treatment"))
  anova_a[[col]] <- aov(f, data = merge_a)
}

anova_b <- list()
for (col in alpha_div_cols) {
  f <- as.formula(paste(col, "~ Injury * Treatment"))
  anova_b[[col]] <- aov(f, data = merge_b)
}

# get a df for a single ANOVA
# anova_a_df <- tidy(anova_a[[1]]) %>%
#   mutate(
#     across(
#       c(sumsq, meansq, statistic, p.value),
#       ~ round(.x, 3))
#     )

# get df for ALL ANOVAs
anova_a_df <- map2_dfr(anova_a, names(anova_a), ~ {
  tidy(.x) %>%
    mutate(
      y_variable = .y,
      across(
        c(sumsq, meansq, statistic, p.value),
        ~ round(.x, 3)
      )
    )
}) %>%
  filter(y_variable %in% c("shannon", "richness", "berger_parker"))

anova_b_df <- map2_dfr(anova_b, names(anova_b), ~ {
  tidy(.x) %>%
    mutate(
      y_variable = .y,
      across(
        c(sumsq, meansq, statistic, p.value),
        ~ round(.x, 3)
      )
    )
}) %>%
  filter(y_variable %in% c("shannon", "richness", "berger_parker"))   # I actually only want these 3

anova_a_df
anova_b_df
```

Same results as last time for *Shannon Index*. 
  All Alpha diversity measures are significantly ~ Injury in cohort A
  Only Pielou's Evenness ~ Injury in cohort B

post-hoc pairwise comparisons
```{r}
post_hoc <- function(y_var, data) {
  f <- as.formula(paste(y_var, "~ inj_trt_group"))
  tests <- list(
    `Sham-Sham;Blast-Sham` = t.test(
      f, data = subset(data, inj_trt_group %in% c("Sham + Sham","Blast + Sham"))
    ),
    `Sham-VNS;Blast-VNS` = t.test(
      f, data = subset(data, inj_trt_group %in% c("Sham + VNS","Blast + VNS"))
    ),
    `Sham-Sham;Sham-VNS` = t.test(
      f, data = subset(data, inj_trt_group %in% c("Sham + Sham","Sham + VNS"))
    ),
    `Blast-Sham;Blast-VNS` = t.test(
      f, data = subset(data, inj_trt_group %in% c("Blast + Sham","Blast + VNS"))
    )
  )
  raw_p <- sapply(tests, \(x) x$p.value)
  p.adjust(raw_p, method = "bonferroni")
}

shan_tests_a <- post_hoc("shannon", merge_a)
shan_tests_b <- post_hoc("shannon", merge_b)
rich_tests_a <- post_hoc("richness", merge_a)
rich_tests_b <- post_hoc("richness", merge_b)
berg_tests_a <- post_hoc("berger_parker", merge_a)
berg_tests_b <- post_hoc("berger_parker", merge_b)

a_post_hocs <- data.frame(
  shannon = shan_tests_a,
  n_species = rich_tests_a,
  berger_parker = berg_tests_a
)

b_post_hocs <- data.frame(
  shannon = shan_tests_b,
  n_species = rich_tests_b,
  berger_parker = berg_tests_b
)
```
  


# Save tables to excel
*We have stats in tables* `anova_a_df`, `anova_b_df`, ``
Let's save them to an excel file
```{r}
wb <- createWorkbook()

addWorksheet(wb, "A_anova")
addWorksheet(wb, "A_posthoc")
addWorksheet(wb, "B_anova")
addWorksheet(wb, "B_posthoc")

writeData(wb, "A_anova",   anova_a_df)
writeData(wb, "A_posthoc", a_post_hocs)
writeData(wb, "B_anova",   anova_b_df)
writeData(wb, "B_posthoc", b_post_hocs)

saveWorkbook(wb, file = "final/alpha_div_tests.xlsx", overwrite = FALSE)
```

