---
title: "02_diversity.Rmd"
author: "Alyssa N. Easton"
date: '2025-09-29'
output: 
---

## Inputs
- `data/intermediate/01_meta_A.rds`
- `data/intermediate/01_meta_B.rds`

## Outputs
- `results/RDS/02_diversity_A.rds`
- `results/RDS/02_diversity_B.rds`

Had to re-run MetaPhlAn to get estimated counts and rarefy counts myself,
as originally raw (non-rarefied) n_reads was associated with Shannon diversity.

Questions:
1. Across the metadata, what is associated with alpha diversity?
2. Does treatment affect alpha diversity?
3. Is treatment's effect on alpha diversity confounded by any variables (batch, weight)?

```{r "setup", include=FALSE}
source("/proj/gibbons/2026_easton_vns/R/paths.R")

knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
knitr::opts_knit$set(root.dir = PROJECT_ROOT)

options(digits = 10)

library(tidyverse)
library(openxlsx)
library(janitor)
library(ggtext)       # wrapper function plot_multiple_dv uses all of these
library(broom)
library(broom.helpers)
library(emmeans)
library(ggpubr)
library(cowplot)
library(grid)
library(gridExtra)
library(gtable)
library(vegan)

source(file.path(PROJECT_ROOT, "R", "utils.R"))
source(file.path(PROJECT_ROOT, "R", "cleaning.R"))
source(file.path(PROJECT_ROOT, "R", "microbiome_preprocess.R"))
source(file.path(PROJECT_ROOT, "R", "diversity_utils.R"))
source(file.path(PROJECT_ROOT, "R", "plotting_utils.R")) # needs ggtext
source(file.path(PROJECT_ROOT, "R", "palettes.R"))
# source(file.path(PROJECT_ROOT, "R", "constants.R"))
# source(file.path(PROJECT_ROOT, "R", "pca_utils.R"))
```

## Imports
```{r}
a_data <- readRDS(file.path(PROJECT_ROOT, "data", "intermediate", "01_meta_A.rds"))
b_data <- readRDS(file.path(PROJECT_ROOT, "data", "intermediate", "01_meta_B.rds"))
```

Merge diversity metrics with metadata
 - Shannon Index (richness + evenness)
 - Berger-Parker Dominance
 - Observed features (richness)
```{r}
div_cols <- c("shannon", "richness", "berger_parker")

a_df <- a_data$clr_alpha %>%
  dplyr::select(all_of(c("seq_name", div_cols))) %>%
  left_join(a_data$meta, by = "seq_name")

b_df <- b_data$clr_alpha %>%
  dplyr::select(all_of(c("seq_name", div_cols))) %>%
  left_join(b_data$meta, by = "seq_name")
```

# Alpha Diversity
### Box Plots
```{r}
label_df <- data.frame(dv = div_cols) %>%
  mutate(label = clean_bug_names(div_cols))

a_bp <- plot_multiple_dv(
  data = a_df,
  dvs = div_cols,
  groups = "inj_trt_group",
  factorA = "Injury",
  factorB = "Treatment",
  x_positions = c(4, 10, 20, 26),
  axis_length = 30,
  palette = injtrt_pal,
  label_df = label_df,
  italicize = FALSE,
  ylab_size = 20,
  show_legend = FALSE,
  fit_fun = aov,
  comparison_mode = "reduced",
  factorA_level_for_factorB = "Blast",
  return_plots = TRUE
  ) %>%
  purrr::map(., ~
  ggdraw(.x) +
    draw_grob(
      make_manual_legend(blast_pal),
      x = -0.10, y = 1.36,
      hjust = 0, vjust = 1
    )
)

a_bp
```
 
```{r}
b_bp <- plot_multiple_dv(
  data = b_df,
  dvs = div_cols,
  groups = "inj_trt_group",
  factorA = "Injury",
  factorB = "Treatment",
  x_positions = c(4, 10, 20, 26),
  axis_length = 30,
  palette = injtrt_pal,
  label_df = label_df,
  italicize = FALSE,
  ylab_size = 20,
  show_legend = FALSE,
  fit_fun = aov,
  comparison_mode = "reduced",
  factorA_level_for_factorB = "Blast",
  return_plots = TRUE
  ) %>%
  purrr::map(., ~
  ggdraw(.x) +
    draw_grob(
      make_manual_legend(blast_pal),
      x = -0.10, y = 1.36,
      hjust = 0, vjust = 1
    )
)

b_bp
```

### 2-way ANOVA
```{r}
anova_a <- extract_two_way_anova_table(
  data = a_df,
  dvs = div_cols,
  factorA = "Injury",
  factorB = "Treatment"
)

anova_b <- extract_two_way_anova_table(
  data = b_df,
  dvs = div_cols,
  factorA = "Injury",
  factorB = "Treatment"
)

anova_a
anova_b
```

Same results as last time for *Shannon Index*. 
  All Alpha diversity measures are significantly ~ Injury in cohort A
  Only Pielou's Evenness ~ Injury in cohort B

# Beta Diversity

`count.betadiv.analyze` is a wrapper function that does several things:
  - generates a Bray-Curtis dissimilarity matrix
  - does traditional PCoA using `vegan::wcmdscale`
  - does conditioned PCoA (corrected for some nuisance variable) using `vegan::capscale`
  - estimates variance explained per PC
  - PERMANOVA given formula
      - by terms
      - by margin
  - combines these outputs into a large list

```{r}
bc_a <- count.betadiv.analyze(
  rarefied_counts = a_data$counts_raref,
  meta_filt    = a_data$meta,
  id_col       = "seq_name",
  perm_formula = "Round_main + blast + vns + blast_and_vns",
  condition    = "Round_main",
  n_perm       = 999
)

bc_a$perm_terms
bc_a$perm_marg
```

```{r}
term_perm_a <- bc_a$perm_terms %>%
  filter(term %in% c("Round_main", "blast", "vns", "blast_and_vns")) %>%
  transmute(
    Term = case_when(
      term == "Round_main"    ~ "Batch",
      term == "blast"         ~ "Injury",
      term == "vns"           ~ "Treatment",
      term == "blast_and_vns" ~ "Interaction"
    ),
    df = as.character(
      paste0(df, ", ", resid_df)
    ),
    p = p.value,
    R2 = R2
  ) %>%
  mutate(Term = factor(Term, levels = c("Interaction", "Injury", "Treatment", "Batch"))) %>%
  arrange(Term)

term_perm_a
```


```{r}
bc_b <- count.betadiv.analyze(
  rarefied_counts = b_data$counts_raref,
  meta_filt    = b_data$meta,
  id_col       = "seq_name",
  perm_formula = "Round_main + blast + vns + blast_and_vns",
  condition    = "Round_main",
  n_perm       = 999
)

bc_b$perm_terms
bc_b$perm_marg
```



```{r}
bc_b$perm_marg
resid_df <- bc_b$perm_marg %>% filter(term == "Residual") %>% pull(df)

term_perm_b <- bc_b$perm_term %>%
  filter(term %in% c("Round_main", "blast", "vns", "blast_and_vns")) %>%
  transmute(
    Term = case_when(
      term == "Round_main"    ~ "Batch",
      term == "blast"         ~ "Injury",
      term == "vns"           ~ "Treatment",
      term == "blast_and_vns" ~ "Interaction"
    ),
    df = as.character(
      paste0(df, ", ", resid_df)
    ),
    p = p.value,
    R2 = R2
  ) %>%
  mutate(Term = factor(Term, levels = c("Interaction", "Injury", "Treatment", "Batch"))) %>%
  arrange(Term)

term_perm_b
```

### Plot A
```{r}
# plot marginal biplot of first 2 PCs
bc_a$cpcoa_scores <- scores(
  bc_a$cpcoa, display = "sites", scaling = 1,
  choices = 1:(bc_a$cpcoa$CCA$rank + bc_a$cpcoa$CA$rank)
  ) %>% as.data.frame() %>%
  rownames_to_column("seq_name") %>%
  left_join(., bc_a$meta_filt %>% dplyr::select(inj_trt_group, seq_name, Treatment), by = "seq_name") %>%
  rename(Group = inj_trt_group) %>%
  mutate(Group = factor(Group, levels = c("Sham + Sham", "Blast + Sham", "Sham + VNS", "Blast + VNS")))

biplot_a <- combined.biplot(
  pc_df = bc_a$cpcoa_scores,
  prop_var = bc_a$cpcoa_var,
  x_axis = "MDS1", y_axis = "MDS2",
  color_var = "Group", color_palette = my_colors,
  shape_var = "Group", shape_palette = my_shapes,
  rel_widths = c(4, 1), point_size = 3,
  legend_text_size = 8
  )

biplot_a
```

### Plot B
```{r}
# plot marginal biplot of first 2 PCs
bc_b$cpcoa_scores <- scores(
  bc_b$cpcoa, display = "sites", scaling = 1,
  choices = 1:(bc_b$cpcoa$CCA$rank + bc_b$cpcoa$CA$rank)
  ) %>% as.data.frame() %>%
  rownames_to_column("seq_name") %>%
  left_join(., bc_b$meta_filt %>% dplyr::select(inj_trt_group, seq_name, Treatment), by = "seq_name") %>%
  rename(Group = inj_trt_group) %>%
  mutate(Group = factor(Group, levels = c("Sham + Sham", "Blast + Sham", "Sham + VNS", "Blast + VNS")))

biplot_b <- combined.biplot(
  pc_df = bc_b$cpcoa_scores,
  prop_var = bc_b$cpcoa_var,
  x_axis = "MDS1", y_axis = "MDS2",
  color_var = "Group", color_palette = my_colors,
  shape_var = "Group", shape_palette = my_shapes,
  rel_widths = c(4, 1), point_size = 3,
  legend_text_size = 8
  )

biplot_b
```

# SAVE
Let's save the tables as-is now, and save to excel later
```{r}
a_results <- list(
  "alphadiv" = list(
    "anova_table" = anova_a,
    "boxplots"     = a_bp
  ),
  "betadiv" = list(
    "permanova_table" = term_perm_a,
    "pcoa_biplot"     = biplot_a
  )
)

b_results <- list(
  "alphadiv" = list(
    "anova_table"     = anova_b,
    "boxplots"        = b_bp
  ),
  "betadiv" = list(
    "permanova_table" = term_perm_b,
    "pcoa_biplot"     = biplot_b
  )
)

save_output(a_results, "diversity_A", "results/RDS")
save_output(b_results, "diversity_B", "results/RDS")
```