---
title: "braycurtis_FINAL.Rmd"
author: "Alyssa N. Easton"
date: '2025-09-12'
output:
---
# Analysis of microbiome beta diversity

Here, I use PERMANOVA to test for associations between treatments and outcomes with overall microbiome composition (measured via Bray-Curtis Distance of MetaPhlAn-defined Species-level Genome Bins (SGBs))

I then use PCoA dimensional reduction to compress the maximum amount of variance in Microbiome Bray-Curtis Dissimilarity into several Principal Coordinates. I *condition* the PCoA on *Batch*, so that the variance explained by batch is removed from the PCs (this makes the treatment effects easier to see). The resulting coordinates are plotted over treatment groups, and tested for association with treatments and outcomes.

## Setup

```{r "setup", include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
knitr::opts_knit$set(root.dir = file.path(Sys.getenv("VNS_ROOT"), "data"))  # stored VNS_ROOT in ~/.Renviron 
options(digits = 10) ## Formats output to 10 digits
library(tidyverse)
library(magrittr)
library(openxlsx)
library(vegan)
library(broom)
library(cowplot)
library(purrr)
library(RColorBrewer)
source("/proj/gibbons/2024_easton_vns/scripts/biplot_boxplot.R")
source("/proj/gibbons/2024_easton_vns/scripts/shannon_div_functions.R")
```

### Import Data
```{r include=FALSE}
micro <- readRDS("/proj/gibbons/2024_easton_vns/data/1_preprocessed/species_october.rds")
metadata_both <- bind_rows(micro$a_meta, micro$b_meta) %>%
  mutate(cohort = factor(ifelse(Round_main %in% c("1", "3"), "A", "B")))
```

# All data combined
If we start by looking at both the acute and chronic data sets together, we quickly
realize that there are major differences, even just on the beta diversity level.
```{r}
rownames(metadata_both) <- metadata_both[["seq_name"]]
bc <- vegdist(micro$all_raref, method = "bray")
  
if (all(rownames(metadata_both) != rownames(bc))) {
  metadata_both <- metadata_both[match(labels(bc), rownames(metadata_both)), ]
}
  
# Do traditional PCoA, outputting variance plot
pcoa <- vegan::wcmdscale(bc, eig = TRUE)
pcoa_var <- var.explain(pcoa = pcoa)
plt_var <- var.part.plot(
  pcoa_var, "Variance in Bray-Curtis Beta Diversity Explained by PCo",
  num_pc = 8) + theme_classic()
  
# Do conditioned PCoA (corrected for some nuisance variable)
cpcoa_formula <- as.formula(paste("bc ~", "cohort"))
cpcoa <- vegan::capscale(cpcoa_formula, data = metadata_both)
  
# extract eigenvalues for CCA and CA
eig <- cpcoa$CA$eig                             # PCs orthogonal to batch (conditioned)
prop_var <- eig / sum(eig)                      # proportion of non-batch-attributed variance explained by each PC
```

## PERMANOVA to quantify batch/cohort effects
Do PERMANOVA, correcting for cohort
```{r}
# do PERMANOVA
perm_formula <- as.formula(paste("bc ~", "cohort + blast + vns + blast_and_vns"))

perm_marg <- adonis2(
  perm_formula,
  data = metadata_both,
  permutations = 999,
  by = "margin"
) %>%
  tidy() %>% mutate(across(c(SumOfSqs, R2, statistic), ~ round(.x, 3)))

perm_terms <- adonis2(
  perm_formula,         # bc ~ X + Y + Z
  data = metadata_both,
  permutations = 999,
  by = "terms"
) %>%
  tidy() %>% mutate(across(c(SumOfSqs, R2, statistic), ~ round(.x, 3)))

perm_marg
perm_terms
```
Cohort explains ~20% of variance between samples

Do PERMANOVA, correcting for batch (which is more fine-grained categorization of cohort)
```{r}
# do PERMANOVA
perm_formula <- as.formula(paste("bc ~", "Round_main + blast + vns + blast_and_vns"))

perm_marg_batch <- adonis2(
  perm_formula,
  data = metadata_both,
  permutations = 999,
  by = "margin"
) %>%
  tidy() %>% mutate(across(c(SumOfSqs, R2, statistic), ~ round(.x, 3)))

perm_terms_batch <- adonis2(
  perm_formula,         # bc ~ X + Y + Z
  data = metadata_both,
  permutations = 999,
  by = "terms"
) %>%
  tidy() %>% mutate(across(c(SumOfSqs, R2, statistic), ~ round(.x, 3)))


perm_marg_batch
perm_terms_batch
```

Batch explains an additional ~30% of variance

## Plot by cohort and by batch

Set up data for this plot
```{r}
# plot marginal biplot of first 2 PCs
pcoa_scores <- scores(pcoa) %>%
  as.data.frame() %>%
  rownames_to_column("seq_name") %>%
  left_join(., metadata_both %>% dplyr::select(inj_trt_group, seq_name, Treatment, Round_main, cohort), by = "seq_name") %>%
  rename(Group = inj_trt_group) %>%
  mutate(Group = factor(Group, levels = c("Sham + Sham", "Blast + Sham", "Sham + VNS", "Blast + VNS")))


# display.brewer.all() # View all available palettes
batch_colors <- brewer.pal(n = 8, name = "Dark2") # Get 5 colors from the "Set1" palette
names(batch_colors) <- unique(pcoa_scores$Round_main)

cohort_shapes <- c("A" = 15, "B" = 16)
```

First, make a plot colored by treatment so we can see that there isn't much of a pattern 
```{r}
biplot_bytreatment <- biplot_main(
  pc_df = pcoa_scores,
  prop_var = pcoa_var,
  x_axis = "Dim1", y_axis = "Dim2",
  color_var = "Group", color_palette = my_colors,
  shape_var = "Group", shape_palette = my_shapes,
  point_size = 3, legend_text_size = 8
  ) +
  theme_bw(base_size = 20) +
  labs(x = "PCo1 (20.4%)", y = "PCo2 (15.1%)") +
  theme(legend.position = "none")

biplot_bytreatment
```
We can tell just visually that there are clearly 3 separate groups

Then, color by batch - this is the main contributor
```{r}
biplot_bybatch <- biplot_main(
  pc_df = pcoa_scores,
  prop_var = pcoa_var,
  x_axis = "Dim1", y_axis = "Dim2",
  color_var = "Round_main", color_palette = batch_colors,
  # shape_var = "cohort", shape_palette = cohort_shapes,
  shape_var = "Group", shape_palette = my_shapes,
  point_size = 3, legend_text_size = 8
  ) + 
  theme_bw(base_size = 20) +
  labs(x = "PCo1 (20.4%)", y = "PCo2 (15.1%)") +
  theme(legend.position = "none")

biplot_bybatch
```

Now save them
```{r}
# ggsave("../figures/ALL_braycurtis_PCoA_bytrt.png", biplot_bytreatment, height = 5, width = 5)
# ggsave("../figures/ALL_braycurtis_PCoA_bybatch.png", biplot_bybatch, height = 5, width = 5)
# MOVED TO LATER
```

### Disregard
  
  I tried to do this with my function, but it didn't work because of the cPCoA and
  PCoA loadings are set up differently, I think
  
```{r}
# bc <- count.betadiv.analyze(
#   rarefied_counts = micro$all_raref,
#   meta_filt    = metadata_both,
#   id_col       = "seq_name",
#   perm_formula = "cohort + blast + vns + blast_and_vns",
#   condition    = NULL,
#   n_perm       = 999
# )
# 
# bc$perm_terms
# bc$perm_marg
# 
# bc_round <- count.betadiv.analyze(
#   rarefied_counts = micro$all_raref,
#   meta_filt    = metadata_both,
#   seq_name       = "seq_name",
#   perm_formula = "Round_main + blast + vns + blast_and_vns",
#   condition    = NULL,
#   n_perm       = 999
# )
# 
# bc_round$perm_terms
# bc_round$perm_marg
# 
# # rarefied count matrix
# count.betadiv.analyze <- c(
#     rarefied_counts,                 # rarefied count matrix
#     meta_filt,                       # metadata with a row (id_col) that matches count matrix row names/col names
#     id_col = "seq_name",             # name of column containing the ID
#     perm_formula,                    # RHS formula for PERMANOVA (i.e. "Round_main + blast + vns + blast_and_vns")
#     condition,                       # if doing a conditional PCoA (residualized), which variable are you removing?
#     n_perm = 999                     # number of permutations
# )
```

# Cohort A (acute, cytokine)
```{r}
bcA <- count.betadiv.analyze(
  rarefied_counts = micro$a_counts_rare,
  meta_filt    = micro$a_meta,
  id_col       = "seq_name",
  perm_formula = "Round_main + blast + vns + blast_and_vns",
  condition    = "Round_main",
  n_perm       = 999
)

bcA$perm_terms
bcA$perm_marg
```

```{r}

# plot marginal biplot of first 2 PCs
bcA$cpcoa_scores <- scores(
  bcA$cpcoa, display = "sites", scaling = 1,
  choices = 1:(bcA$cpcoa$CCA$rank + bcA$cpcoa$CA$rank)
  ) %>% as.data.frame() %>%
  rownames_to_column("seq_name") %>%
  left_join(., bcA$meta_filt %>% dplyr::select(inj_trt_group, seq_name, Treatment), by = "seq_name") %>%
  rename(Group = inj_trt_group) %>%
  mutate(Group = factor(Group, levels = c("Sham + Sham", "Blast + Sham", "Sham + VNS", "Blast + VNS")))

biplot_A <- combined.biplot(
  pc_df = bcA$cpcoa_scores,
  prop_var = bcA$cpcoa_var,
  x_axis = "MDS1", y_axis = "MDS2",
  color_var = "Group", color_palette = my_colors,
  shape_var = "Group", shape_palette = my_shapes,
  rel_widths = c(4, 1), point_size = 3,
  legend_text_size = 8
  )

biplot_A

```
**This was 27.3, 20.4**, now it's *26.1, 20.8* ...?

Now 
```{r}
# here's what we already did in our function
bcA$pcoa
bcA$pcoa_var
bcA$pcoa_var_plt

# now, make this into an easy df for plotting
A_pcoa_df <- bcA$pcoa$points %>% as.data.frame() %>%
  rownames_to_column("seq_name") %>%
  left_join(., bcA$meta_filt %>% dplyr::select(inj_trt_group, seq_name, Treatment, Round_main), by = "seq_name") %>%
  rename(Group = inj_trt_group) %>%
  mutate(Group = factor(Group, levels = c("Sham + Sham", "Blast + Sham", "Sham + VNS", "Blast + VNS")))

# plot by treatment
pcoa_biplot_A <- combined.biplot(
  pc_df = A_pcoa_df,
  prop_var = bcA$pcoa_var,
  x_axis = "Dim1", y_axis = "Dim2",
  color_var = "Group", color_palette = my_colors,
  shape_var = "Group", shape_palette = my_shapes,
  rel_widths = c(4, 1), point_size = 3,
  legend_text_size = 8
  )
  

# plot by batch
pcoa_biplot_A_batch <- biplot_main(
  pc_df = A_pcoa_df,
  prop_var = bcA$pcoa_var,
  x_axis = "Dim1", y_axis = "Dim2",
  color_var = "Round_main", color_palette = batch_colors,
  shape_var = "Group", shape_palette = my_shapes,
  point_size = 3, legend_text_size = 8
  ) + theme(legend.position = "none")

pcoa_biplot_A
pcoa_biplot_A_batch
```


## Cohort B (chronic, behavior)
```{r}
bcB <- count.betadiv.analyze(
  rarefied_counts = micro$b_counts_rare,
  meta_filt       = micro$b_meta,
  id_col          = "seq_name",
  perm_formula    = "Round_main + blast + vns + blast_and_vns",
  condition       = "Round_main",
  n_perm          = 999
)

bcB$perm_terms
bcB$perm_marg
```

```{r}
# PCoA variance explained + marginal biplot
# bcA$pcoa_var_plt

# cPCoA
# plot variance explained?

# plot marginal biplot of first 2 PCs
bcB$cpcoa_scores <- scores(
  bcB$cpcoa, display = "sites", scaling = 1,
  choices = 1:(bcB$cpcoa$CCA$rank + bcB$cpcoa$CA$rank)
  ) %>% as.data.frame() %>%
  rownames_to_column("seq_name") %>%
  left_join(., bcB$meta_filt %>% dplyr::select(inj_trt_group, seq_name, Treatment), by = "seq_name") %>%
  rename(Group = inj_trt_group) %>%
  mutate(Group = factor(Group, levels = c("Sham + Sham", "Blast + Sham", "Sham + VNS", "Blast + VNS")))

biplot_B <- combined.biplot(
  pc_df = bcB$cpcoa_scores,
  prop_var = bcB$cpcoa_var,
  x_axis = "MDS1", y_axis = "MDS2",
  color_var = "Group", color_palette = my_colors,
  shape_var = "Group", shape_palette = my_shapes,
  rel_widths = c(4, 1), point_size = 3,
  legend_text_size = 8
  )

biplot_B
```
**This was 22.3, 17.6**

**How do I go about doing Variance explained here?** Is MDS var explained a percent of the total variance, or of non-batch variance?
answer: We calculated MDS variance explained as a percent of *Total* variance!

Now do the original PCoA
```{r}
# here's what we already did in our function
bcB$pcoa
bcB$pcoa_var
bcB$pcoa_var_plt

# now, make this into an easy df for plotting
B_pcoa_df <- bcB$pcoa$points %>% as.data.frame() %>%
  rownames_to_column("seq_name") %>%
  left_join(., bcB$meta_filt %>% dplyr::select(inj_trt_group, seq_name, Treatment, Round_main), by = "seq_name") %>%
  rename(Group = inj_trt_group) %>%
  mutate(Group = factor(Group, levels = c("Sham + Sham", "Blast + Sham", "Sham + VNS", "Blast + VNS")))

# plot
pcoa_biplot_B <- combined.biplot(
  pc_df = B_pcoa_df,
  prop_var = bcB$pcoa_var,
  x_axis = "Dim1", y_axis = "Dim2",
  color_var = "Group", color_palette = my_colors,
  shape_var = "Group", shape_palette = my_shapes,
  rel_widths = c(4, 1), point_size = 3,
  legend_text_size = 8
  )

# plot by batch
pcoa_biplot_B_batch <- biplot_main(
  pc_df = B_pcoa_df,
  prop_var = bcB$pcoa_var,
  x_axis = "Dim1", y_axis = "Dim2",
  color_var = "Round_main", color_palette = batch_colors,
  shape_var = "Group", shape_palette = my_shapes,
  point_size = 3, legend_text_size = 8
  ) + theme(legend.position = "none")

pcoa_biplot_B
pcoa_biplot_B_batch
```


# Save plots as svg files
```{r}
# main figures: cPCoA separate
ggsave("../figures/beta/A_braycurtis_cPCoA_new.svg", biplot_A, height = 5, width = 5, device = "svg")
ggsave("../figures/beta/B_braycurtis_cPCoA_new.svg", biplot_B, height = 5, width = 5, device = "svg")

# supplemental
### PCoA of both cohorts
ggsave("../figures/beta/ALL_braycurtis_PCoA_bytrt.svg", biplot_bytreatment, height = 5, width = 5, device = "svg")
ggsave("../figures/beta/ALL_braycurtis_PCoA_bybatch.svg", biplot_bybatch, height = 5, width = 5, device = "svg")
### PCoA of just A
ggsave("../figures/beta/A_braycurtis_PCoA_bytrt.svg", pcoa_biplot_A, height = 5, width = 5, device = "svg")
ggsave("../figures/beta/A_braycurtis_PCoA_bybatch.svg", pcoa_biplot_A_batch, height = 5, width = 5, device = "svg")
### PCoA of just B
ggsave("../figures/beta/B_braycurtis_PCoA_bytrt.svg", pcoa_biplot_B, height = 5, width = 5, device = "svg")
ggsave("../figures/beta/B_braycurtis_PCoA_bybatch.svg", pcoa_biplot_B_batch, height = 5, width = 5, device = "svg")
```

# Save tables to excel
*We have stats in tables:*
    `perm_terms_table_a`   `perm_marg_table_a`    `ccanova_table_a`
  - `perm_terms_table_b`   `perm_marg_table_b`    `ccanova_table_b`
```{r}
stats_list <- list(
  "perm_terms_a" = bcA$perm_terms,
  "perm_marg_a"  = bcA$perm_marg,
  "perm_terms_b" = bcB$perm_terms,
  "perm_marg_b"  = bcB$perm_marg
)

wb <- createWorkbook()

imap(stats_list, ~ {
  addWorksheet(wb, .y)  # .y = names
  writeData(wb, .y, .x) # .x = values
})

saveWorkbook(wb, file = "final/beta_diversity_stats_UPDATED.xlsx", overwrite = TRUE)
```

