---
title: "vns_genera_barplots.Rmd"
author: "Alyssa N. Easton"
date: '2025-09-05'
output: html_document
---

# Preprocessing Thome files to combine

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
knitr::opts_knit$set(root.dir = file.path(Sys.getenv("VNS_ROOT"), "data"))  # stored VNS_ROOT in ~/.Renviron
options(digits = 10) ## Formats output to 10 digits
library(tidyverse)
library(forcats)
library(tibble)
library(data.table)
library(magrittr)
library(openxlsx)
library(janitor)
library(purrr)
library(broom)
library(pheatmap)
library(circlize)
library(cowplot)
library(ggpubr)
library(ComplexHeatmap)
library(svglite)
source("/proj/gibbons/2024_easton_vns/scripts/june_functions.R")
source("/proj/gibbons/2024_easton_vns/scripts/shannon_div_functions.R")

```

Data
```{r}
micro <- readRDS("/proj/gibbons/2024_easton_vns/data/1_preprocessed/species_october.rds")

a_relabund <- micro$all_abund %>% filter(seq_name %in% micro$a_meta$seq_name) %>%
  dplyr::select(where(~ !all(is.na(.)))) # %>%
  # mutate(across(everything(), ~replace_na(., 0)))
  
b_relabund <- micro$all_abund %>% filter(seq_name %in% micro$b_meta$seq_name) %>%
  dplyr::select(where(~ !all(is.na(.)))) # %>%
  # mutate(across(everything(), ~replace_na(., 0)))


# micro <- readRDS("/proj/gibbons/2024_easton_vns/data/1_preprocessed/microbe_ncbi.rds")
# species_counts <- micro$metaphlan_100 %>% mutate(seq_name = as.numeric(seq_name))        # now rel_abund
# taxa_map       <- micro$metaphlan_tax_100_map
# taxa_cols      <- colnames(species_counts) %>% setdiff("seq_name")
# merged         <- readRDS("/proj/gibbons/2024_easton_vns/data/1_preprocessed/meta_outcomePC_all.rds")
# merged_clr     <- clr.return.presence(data = merged,
#                                   taxa_cols = taxa_cols,
#                                   id_col = "seq_name")
```

```{r}
# a <- merged %>%
#   filter(cohort == "A") %>%
#   arrange(inj_trt_group, Round_main)
# 
# b <- merged %>%
#   filter(cohort == "B") %>%
#   arrange(inj_trt_group, Round_main)
# 
# four_colors <- c(
#   "Sham + Sham" = "#1032ff",
#   "Sham + VNS" = "#00fdff",
#   "Blast + Sham" = "#ff9200",
#   "Blast + VNS" = "#942493"
#   )
```

# sort by `inj_trt_group`, *then* by `Round_main` (so we'll sort of see batch effects)
```{r}
get.plot.data <- function(meta, rel_abund) {
  # arrange in order of Group, then batch, then cage.
  meta <- meta %>%
    arrange(inj_trt_group, Round_main, Cage) %>%
    mutate(seq_name = as.character(seq_name)) %>%
    inner_join(rel_abund, by = "seq_name") %>%
    mutate(
      inj_trt_group = fct_relevel(inj_trt_group, "Sham + Sham", "Blast + Sham", "Sham + VNS", "Blast + VNS")
    )
  
  taxa_cols <- colnames(rel_abund) %>% setdiff("seq_name")
  
  # meta$unclassified <- 100 - (
  #   meta %>%
  #     dplyr::select(all_of(taxa_cols)) %>%
  #     rowSums()
  #   )
  # 
  # taxa_cols <- c(taxa_cols, "unclassified")
  
  # pivot to long format, where "Feature" is a column that contains the species
  long_df <- meta %>%
    pivot_longer(
      cols = all_of(taxa_cols),
      names_to = "Species",
      values_to = "Abundance"
    )
  
  # get the top species
  top_features <- long_df %>%
    group_by(Species) %>%
    summarise(total_abundance = sum(Abundance, na.rm = TRUE)) %>%
    filter(Species != "unclassified") %>%
    slice_max(total_abundance, n = 10) %>%
    pull(Species)
  
  long_df <- long_df %>%
    mutate(
      Species = ifelse(Species %in% top_features, Species, "Other")
    ) %>%
    group_by(seq_name, Species) %>%   # Group by sample and species
    mutate(Abundance = sum(Abundance, na.rm = TRUE)) %>%  # Sum abundance
    ungroup() %>%
    distinct() %>%
    mutate(
      Species = fct_reorder(Species, Abundance, .fun = sum, .desc = TRUE),
      Species = fct_relevel(Species, "Other", after = 0)
      )
  
  return(long_df)
}

```

```{r}
plot_df_a <- get.plot.data(meta = micro$a_meta, rel_abund = a_relabund)
plot_df_b <- get.plot.data(meta = micro$b_meta, rel_abund = b_relabund)
```

Don't need to replot, but could change order of groups here
```{r}
plot_a <- ggplot(plot_df_a, aes(x = as.character(seq_name), y = Abundance, fill = Species)) +
  geom_bar(stat = "identity") +
  facet_grid(~inj_trt_group, scales = "free_x", space = "free") +  # groups samples visually by Group
  theme_classic() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        axis.line.y = element_blank()) +
  labs(x = "Sample", y = "Relative Abundance", fill = "Species") +
  scale_fill_manual(values = palette_16)
  # scale_fill_manual(values = new_palette)
plot_a
# ggsave("../figures/species_barplot_a.svg", plot_a, height = 3, width = 5, device = "svg")
```

Experimented with different orders, but decided the first one was the best
```{r}
plot_b <- ggplot(plot_df_b, aes(x = as.character(seq_name), y = Abundance, fill = Species)) +
  geom_bar(stat = "identity") +
  facet_grid(~inj_trt_group, scales = "free_x", space = "free") +  # groups samples visually by Group
  theme_classic() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        axis.line.y = element_blank()) +
  labs(x = "Sample", y = "Relative Abundance", fill = "Species") +
  scale_fill_manual(values = palette_16)
  # scale_fill_manual(values = new_palette)
plot_b
# ggsave("../figures/species_barplot_b.svg", plot_b, height = 4, width = 6, device = "svg")
```


# Old code + trying different colors
```{r}
all_taxa <- unique(c(plot_df_a$Species, plot_df_b$Species)) # this is what I used to develop a consistent palette
all_taxa
```

7d0225
CAADFF light purple
"#4CBB17", # Kelly green
"#E69F00", # orange
"#6A3D9A", # dark purple
"#EC5F6B", # watermelon"


order_a <- c(
  "Other", "Muribaculaceae_bacterium", "GGB46159_SGB63927", "GGB70852_SGB95305",
  "Dubosiella_newyorkensis", "Lactobacillus_johnsonii", "GGB27878_SGB40312",
  "Akkermansia_muciniphila", "Lachnospiraceae_bacterium",
  "Bacteroides_thetaiotaomicron", "GGB1515_SGB2095"
)

order_b <- c(
  "Other", "GGB27876_SGB40310", "Dubosiella_newyorkensis", "Lactobacillus_johnsonii",
  "Muribaculaceae_bacterium", "Alistipes_sp_DSM_112343", "GGB27918_SGB40356",
  "GGB27878_SGB40312", "Bifidobacterium_pseudolongum", "Akkermansia_muciniphila",
  "Turicibacter_sp_1E2"
)


plot_df_a <- plot_df_a %>%        # Set factor levels in your plotting data
  mutate(Species = factor(Species, levels = order_a)) %>%
  arrange()

plot_df_b <- plot_df_b %>%
  mutate(Species = factor(Species, levels = order_b))

order_b <- c(
  "Other", "Dubosiella_newyorkensis", "GGB27876_SGB40310", "Muribaculaceae_bacterium",
  "Lactobacillus_johnsonii", "GGB27918_SGB40356", "Alistipes_sp_DSM_112343",
  "GGB27878_SGB40312", "Bifidobacterium_pseudolongum", "Akkermansia_muciniphila",
  "Turicibacter_sp_1E2"
)
plot_df_b <- plot_df_b %>%
  mutate(Species = factor(Species, levels = order_b))