---
title: "B_joint_final.Rmd"
author: "Alyssa N. Easton"
date: '2025-01-08'
output:
---

Testing the hypothesis:
“Microbe mediates ANY component of the blast → outcome relationship”

## Setup
```{r "setup", include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
knitr::opts_knit$set(root.dir = file.path(Sys.getenv("VNS_ROOT"), "data"))  # stored VNS_ROOT in ~/.Renviron
options(expressions = 6000)  # Increase allowed recursion depth
options(digits = 4) ## Formats output to 10 digits
library(magrittr)
library(tidyverse)
library(broom)
library(broom.helpers)
library(rlang)
library(glue)
library(janitor)
library(purrr)
library(ggh4x)
library(zCompositions)
library(parallel)
library(parallelly)
library(pbapply)
library(furrr)      # future_pmap comes from here
library(future)
library(limma)
library(ggeffects)
library(patchwork)
library(lme4)
library(lmerTest)
library(performance)
library(MuMIn)
library(car)
library(scales)
library(future.apply)
library(ComplexHeatmap)
library(circlize)
library(cowplot)
library(Cairo)
library(gridExtra)
# library(lmeresampler)
source("/proj/gibbons/2024_easton_vns/workflow/functions/mdact.R")
source("/proj/gibbons/2024_easton_vns/workflow/functions/june_functions.R")
source("/proj/gibbons/2024_easton_vns/workflow/functions/shannon_div_functions.R")
source("/proj/gibbons/2024_easton_vns/workflow/functions/artifact_correct_functions.R")
source("/proj/gibbons/2024_easton_vns/workflow/functions/lmer_functions.R")
```

## Import and preprocess data
    metadata, microbe presence/absence, sample-wise presence/absence, batch-wise presence/absence
    compile into a summary `bug_info`
```{r}
# read in important data
all_data <- readRDS("/proj/gibbons/2024_easton_vns/data/1_preprocessed/species_october.rds")
# alpha_b <- readRDS("/proj/gibbons/2024_easton_vns/data/diversity/B_alpha.rds")
bug_info <- readRDS("../workflow/B_bug_info.rds")

presence_df <- all_data$b_counts %>%
  mutate(across(everything(), ~ . != 0)) %>%
  rownames_to_column("seq_name")

per_sample_presence <- all_data$b_counts %>%
  mutate(across(everything(), ~ . != 0)) %>%
  {rowSums(.) / ncol(.)}

# all_joint_regr <- readRDS("../B_pM_lmer_0106.rds")
# out_list <- readRDS("../B_outcome_lmer_0106.rds")

outcome_cols <- c(
  "OFB_Cdistance",  "OFB_Ctime",                "OFB_Centries",  
  "OFB_Clatency",   "startle_habituation",      "startle_PPI", 
  "PD_choice_filt", "EtOH_24h_intakeave_final", "EtOH_8pm_intakeave_final",
  "EtOH_pref_24h_final")
```

# 1: Statistical models

Processing
```{r}
args <- list(
  total_presence      =  0.30,
  batch_n_min         =     2,
  impute_method       = "GBM",
  filt_before_clr     =  TRUE,
  ctrl_batch          =  TRUE)

filt_bugs <- bug_info %>% 
  filter(
    pres_total     >= args$total_presence,
    n_batch        >= args$batch_n_min    ) %>%
  pull(microbe)

if (args$filt_before_clr) {
  filt_counts <- all_data$b_counts %>%
    dplyr::select(any_of(filt_bugs))
} else {
  filt_counts <- all_data$b_counts
}

imputed_counts <- cmultRepl(filt_counts, label = 0, method = args$impute_method, z.delete = FALSE)
gm <- exp(rowMeans(log(imputed_counts)))
clr_mat <- log(imputed_counts) - log(gm)
clr_df <- as_tibble(clr_mat) %>% scale() %>% as.data.frame() # NOW i've scaled it
colnames(clr_df) <- colnames(filt_counts)
rownames(clr_df) <- rownames(clr_mat)
```


Do we want to add alpha diversity features like we did with cohort A? Probably good to stay consistent
```{r}
alpha <- readRDS("/proj/gibbons/2024_easton_vns/data/diversity/B_alpha.rds")

adiv_features <- alpha %>%
  transmute(shannon, richness, berger_parker) %>%
  rownames_to_column("seq_name") %>%
  mutate(
    shannon = scale(shannon),
    richness = scale(richness),
    berger_parker = scale(log(berger_parker))
    )

# hist(alpha$berger_parker)
# hist(adiv_features$berger_parker)
```


Merge all data into one df 
  Treatments (exposures) are binary
  Microbes are filtered, CLR-transformed, and then z-scored
  Outcome variables are z-scored
```{r}
df <- all_data$b_meta %>%
  dplyr::select(all_of(c(
    "seq_name", "inj_trt_group", "blast", "vns", "blast_and_vns", "Round_main", outcome_cols
  ))) %>%
  mutate(seq_name = as.character(seq_name)) %>% 
  right_join((clr_df %>% rownames_to_column("seq_name")), by = "seq_name") %>%
  mutate(across(all_of(outcome_cols), ~as.numeric(scale(.)))) %>%
  left_join(adiv_features, by = "seq_name")

filt_bugs <- c(filt_bugs, "shannon", "richness", "berger_parker")

saveRDS(df, "../data/processed/B_clr_df_withalpha_0216.rds")
```

## Model microbe (M)

Differential abundance testing (contains p_m)
  > Added in testing for VNS effect given blast == 1
  
### LRT p-values
```{r}
model_microbe <- map_dfr(filt_bugs, function(bug_name) {

  form <- as.formula(paste0(
    bug_name,
    " ~ blast + vns + blast_and_vns + (1 | Round_main)"
  ))

  terms <- c("blast", "vns", "blast_and_vns") # removed "vns_on_blast"
  n_samples <- sum(!is.na(df[[bug_name]]))

  tryCatch({

    m <- lmer(form, data = df, REML = FALSE, na.action = "na.exclude")
    sm <- summary(m)$coefficients

    ################## Fixed effects ##################
    fe <- tibble(
      term      = rownames(sm),
      estimate  = sm[, "Estimate"],
      std.error = sm[, "Std. Error"],
      statistic = sm[, "t value"]
    ) %>%
      filter(term %in% c("blast", "vns", "blast_and_vns"))

    ################## P-values ##################
    # since there is no exact partial F-test for LMER, Chi-squared is the best approximation (asymptotic joint hypotheses)
    pvals <- tibble(
      term = terms,
      p.value = c(
        anova(update(m, . ~ . - blast - blast_and_vns), m, test = "Chisq")$`Pr(>Chisq)`[2], # any blast effect
        anova(update(m, . ~ . - vns - blast_and_vns), m, test = "Chisq")$`Pr(>Chisq)`[2],   # any vns effect
        anova(update(m, . ~ . - blast_and_vns), m, test = "Chisq")$`Pr(>Chisq)`[2]          # interaction
        # car::linearHypothesis(m,"vns + blast_and_vns = 0")$`Pr(>Chisq)`[2] # VNS effect when blast = 1 (linear contrast)
      )
    )

    ################## Combine ##################
    bind_rows(
      fe %>% dplyr::select(term, estimate, std.error, statistic) %>%
    #   tibble(term = "vns_on_blast", estimate = NA_real_, std.error = NA_real_, statistic = NA_real_)) %>%
      left_join(pvals, by = "term") %>%
      mutate(
        microbe = bug_name,
        n = n_samples
      )
    )

  }, error = function(e) {

    tibble(
      term      = terms,
      estimate  = NA_real_,
      std.error = NA_real_,
      statistic = NA_real_,
      p.value   = NA_real_,
      microbe   = bug_name,
      n         = n_samples
    )
  })
}) %>%
  group_by(term) %>%
  mutate(adj.p = p.adjust(p.value, method = "BH")) %>%
  ungroup()

# saveRDS(model_microbe, "../B_microbe_models_0107.rds")
saveRDS(model_microbe, "../data/3_newest/B_microbe_models_0216.rds")
```

### Coefficients and p-values (for heat maps)
```{r}
# form <- as.formula("Lachnospiraceae_unclassified_SGB41414 ~ blast + vns + blast_and_vns + (1|Round_main)")
# testlm <- lmer(form, data = df, REML = FALSE, na.action = "na.exclude")
# summary(testlm)

coef_microbe <- map_dfr(filt_bugs, function(bug_name) {
  
  form <- as.formula(paste0(bug_name, " ~ blast + vns + blast_and_vns + (1|Round_main)"))
  terms <- c("blast", "vns", "blast_and_vns", "(Intercept)") # Round_main random effects are stored differently
  n_samples <- sum(!is.na(df[[bug_name]]))

  tryCatch({

    m <- lmer(form, data = df, REML = FALSE, na.action = "na.exclude")
    sm <- summary(m)$coefficients
    
    ################## Fixed effects ##################
    fe <- tibble(
      term      = rownames(sm),
      estimate  = sm[, "Estimate"],
      std.error = sm[, "Std. Error"],
      statistic = sm[, "t value"],
      p.value   = sm[, "Pr(>|t|)"]
    ) %>%
      filter(term %in% terms)
  
    bind_rows(fe %>% mutate(microbe = bug_name, n = n_samples))
    
  }, error = function(e) {

    tibble(
      term      = terms,
      estimate  = NA_real_,
      std.error = NA_real_,
      statistic = NA_real_,
      p.value   = NA_real_,
      microbe   = bug_name,
      n         = n_samples
    )
  })
}) %>%
  group_by(term) %>%
  mutate(adj.p = ifelse(
    term == "(Intercept)", NA_real_,
    p.adjust(p.value, method = "BH")
    )) %>%
  ungroup()

saveRDS(coef_microbe, "../data/3_newest/B_microbe_coef_pvals_0216.rds")

head(coef_microbe)
```

## Model outcome (Y)

Test whether microbes explain any variance in outcome not explained by treatment

### LRT
```{r}
model_outcome <- outcome_cols %>% # used two_col to test
  set_names() %>%
  map(function(out) {
    map_dfr(filt_bugs, function(bug_name) {
      df$microbe <- df[[bug_name]]
      df <- df %>% filter(!is.na(out))      # added this to remove NA outcome samples
      form <- as.formula(paste0(out, " ~ microbe + blast + vns + blast_and_vns + (1 | Round_main)"))
      terms <- c("blast", "vns", "blast_and_vns", "microbe")  # , "vns_on_blast")
      n_samples <- sum(!is.na(df[[bug_name]])) # count non-missing samples for this bug

      tryCatch({
        m <- lmer(form, data = df, REML = FALSE, na.action = "na.exclude")
        sm <- summary(m)$coefficients   ## ---- fixed-effect estimates ----
        fe <- tibble(
          term      = rownames(sm),
          estimate  = sm[, "Estimate"],
          std.error = sm[, "Std. Error"],
          statistic = sm[, "t value"]
        ) %>%
          filter(term %in% terms)
        
        ################## P-values ##################
        pvals <- tibble(                ## ---- LRT p-values ----
          term = terms,
          p.value = c(                  # not currently using, but could use anova(, ddf = "Kenward-Roger")
            anova(update(m, . ~ . - blast - blast_and_vns), m, test = "Chisq")$`Pr(>Chisq)`[2], # any blast effect
            anova(update(m, . ~ . - vns - blast_and_vns), m, test = "Chisq")$`Pr(>Chisq)`[2],   # any vns effect
            anova(update(m, . ~ . - blast_and_vns), m, test = "Chisq")$`Pr(>Chisq)`[2],         # interaction / synergy
            anova(update(m, . ~ . - microbe), m, test = "Chisq")$`Pr(>Chisq)`[2]                # microbe
            # car::linearHypothesis(m, "vns + blast_and_vns = 0")$`Pr(>Chisq)`[2]   # vns on blast
          ))
        
        ################## Combine ##################
        bind_rows(
          fe %>% dplyr::select(term, estimate, std.error, statistic) #,
          # tibble(
          #   term = "vns_on_blast",
          #   estimate = NA_real_,
          #   std.error = NA_real_,
          #   statistic = NA_real_
          # )
        ) %>%
          left_join(pvals, by = "term") %>%
          mutate(
            microbe = bug_name,
            n = n_samples
          )

        fe %>%
          left_join(pvals, by = "term") %>%
          mutate(
            microbe = bug_name,
            n = n_samples
          )
      }, error = function(e) {
        tibble(
          term      = terms,
          estimate  = NA_real_,
          std.error = NA_real_,
          statistic = NA_real_,
          p.value   = NA_real_,
          microbe   = bug_name,
          n         = n_samples
        )
    })
  }) %>%
    mutate(outcome = out) %>%
    group_by(term) %>%
    mutate(adj.p = p.adjust(p.value, method = "BH")) %>%
    ungroup()
})

# saveRDS(model_outcome, "../B_outcome_models_0107.rds")
saveRDS(model_outcome, "../data/3_newest/B_outcome_models_0216.rds")
```

### Coefficient p-values
```{r}
coef_outcome <- outcome_cols %>% # used two_col to test
  set_names() %>%
  map(function(out) {
    
    map_dfr(filt_bugs, function(bug_name) {
    
      df$microbe <- df[[bug_name]]
      df <- df %>% filter(!is.na(out))      # added this to remove NA outcome samples
      form <- as.formula(paste0(out, " ~ microbe + blast + vns + blast_and_vns + (1 | Round_main)"))
      terms <- c("blast", "vns", "blast_and_vns", "microbe")  # , "vns_on_blast")
      n_samples <- sum(!is.na(df[[bug_name]])) # count non-missing samples for this bug
  
      tryCatch({
        m <- lmer(form, data = df, REML = FALSE, na.action = "na.exclude")
        sm <- summary(m)$coefficients
        
        ################## Fixed effects ##################
        fe <- tibble(
          term      = rownames(sm),
          estimate  = sm[, "Estimate"],
          std.error = sm[, "Std. Error"],
          statistic = sm[, "t value"],
          p.value   = sm[, "Pr(>|t|)"]
        ) %>%
          filter(term %in% terms)
      
        bind_rows(fe %>% mutate(microbe = bug_name, n = n_samples))
        
      }, error = function(e) {
    
        tibble(
          term      = terms,
          estimate  = NA_real_,
          std.error = NA_real_,
          statistic = NA_real_,
          p.value   = NA_real_,
          microbe   = bug_name,
          n         = n_samples
        )
      })
      
    }) %>%
      mutate(outcome = out) %>%
      group_by(term) %>%
      mutate(adj.p = ifelse(term == "(Intercept)", NA_real_, p.adjust(p.value, method = "BH"))) %>%
      ungroup()
  })

saveRDS(coef_outcome, "../data/3_newest/B_outcome_coefs_0216.rds")

# head(coef_outcome)
head(coef_outcome)
```


Organize p-values for mediation analysis
```{r}
p_m_df <- model_microbe %>%
  pivot_wider(id_cols = "microbe", names_from = "term",
    values_from = "p.value", names_prefix = "pm_"
    )

p_y_df <- model_outcome %>%
  map_dfr(function(one) {
    one %>%
      dplyr::filter(term == "microbe") %>%
      dplyr::select(microbe, p.value) %>%
      dplyr::mutate(outcome = unique(one$outcome))
  }) %>%
  pivot_wider(
    id_cols = "microbe",
    names_from = "outcome",
    values_from = "p.value"
  )

all_p_df <- p_m_df %>%
  left_join(p_y_df, by = "microbe")

head(all_p_df)

# saveRDS(all_p_df, "../B_p_values_0108.rds")
saveRDS(all_p_df, "../data/3_newest/B_LRT_p_values_0216.rds")
```

Also merge regular coefficient p-values
```{r}
m_coef_df <- coef_microbe %>%
  mutate(
    model = "m",
    term = ifelse(term == "blast_and_vns", "bv", term)
    ) %>%
  transmute(microbe, term, "est" = estimate, "p" = p.value) %>%
  filter(term %in% c("blast", "vns", "bv")) %>%
  pivot_wider(
    id_cols = "microbe", names_from = "term",
    values_from = c("est", "p"),
    names_glue = "{.value}_{term}"
    )

y_coef_df <- coef_outcome %>%
  map_dfr(function(one) {
    one %>%
      mutate(
        outcome = unique(one$outcome),
        model = "y",
        term = ifelse(term == "blast_and_vns", "bv", term)
      )
  }) %>%
  transmute(microbe, outcome, term, "est" = estimate, "p" = p.value) %>%
  filter(term %in% c("blast", "vns", "bv", "microbe")) %>%
  pivot_wider(
    id_cols = c("microbe", "outcome"),
    names_from = "term",
    values_from = c("est", "p"),
    names_glue = "{.value}_{term}"
    )

head(m_coef_df)
head(y_coef_df)
```
### Could remove batch effect on models with singular fit...

## M-DACT

```{r}
# set up a list with arguments for different M-DACT runs
i <- c("pm_blast", "pm_vns")
j <- outcome_cols

mdact_runs <- expand.grid(med_name = i, out_name = j) %>%
  split(seq_len(nrow(.))) %>% 
  map(as.list)

# run everything at FDR = 0.1
all_runs <- lapply(mdact_runs, function(x) {
  one_mdact(
    all_p_df = all_p_df,
    pm = x[["med_name"]],
    py = x[["out_name"]],
    fdr = 0.1
    )
})
```

```{r}

thr_long <- expand.grid(
  exp_name = i, out_name = j, fdr = c(0.100, 0.125, 0.150, 0.175, 0.200, 0.225, 0.250),
  stringsAsFactors=FALSE
  ) %>%
  pmap_dfr(function(exp_name, out_name, fdr) {
    one_mdact_thr(
      all_p_df = all_p_df,
      pm  = exp_name,
      py  = out_name,
      fdr = fdr
    ) %>%
      mutate(fdr = fdr)
  })

thr_wide <- thr_long %>%
  mutate(fdr = paste0("dact_thr_", fdr * 100)) %>%
  pivot_wider(
    names_from  = fdr,
    values_from = dact_thr
  )

head(thr_wide)
```
Final merge
```{r}
all_run_df <- bind_rows(all_runs) %>%
  left_join(thr_wide, by = c("exposure", "outcome")) %>%
  mutate(
    signif_at = case_when(
      dact_stat < dact_thr_10     ~ 0.100,
      dact_stat < `dact_thr_12.5` ~ 0.125,
      dact_stat < dact_thr_15     ~ 0.150,
      dact_stat < `dact_thr_17.5` ~ 0.175,
      dact_stat < dact_thr_20     ~ 0.200,
      dact_stat < `dact_thr_22.5` ~ 0.225,
      dact_stat < dact_thr_25     ~ 0.250,
      TRUE ~ NA_real_
    )
  )

all_run_df %>%
  arrange(dact_stat) %>%
  head()

saveRDS(all_run_df, "../data/3_newest/B_mediation_0216.rds")
```


# 2: Analyzing DA and M-DACT results

Since I already ran prior analyses, I can just import these dfs
```{r}
df            <- readRDS("../data/processed/B_clr_df_withalpha_0216.rds")
model_microbe <- readRDS("../data/3_newest/B_microbe_models_0216.rds")
model_outcome <- readRDS("../data/3_newest/B_outcome_models_0216.rds")
all_p_df      <- readRDS("../data/3_newest/B_LRT_p_values_0216.rds")
all_run_df    <- readRDS("../data/3_newest/B_mediation_0216.rds")
```

## Parsing LMER results
### Exposure -> microbe
```{r}
da_bugs <- model_microbe %>%
  filter(adj.p < 0.2, term %in% c("blast", "vns")) %>%
  pull(microbe) %>%
  unique()

lrt_pvals <- model_microbe %>%
  filter(microbe %in% da_bugs) %>%
  transmute(microbe, term, "p" = p.value, "q" = adj.p) %>%
  filter(term %in% c("blast", "vns")) %>%
  pivot_wider(
    id_cols = "microbe", names_from = "term",
    values_from = c("p", "q"),
    names_glue = "{.value}_{term}"
    ) %>%
  mutate(
    blast_signif = ifelse(q_blast < 0.2, TRUE, FALSE),
    vns_signif = ifelse(q_vns < 0.2, TRUE, FALSE),
    microbe = gsub("_", " ", microbe)
  )

da_blast <- setNames(lrt_pvals$blast_signif, nm = lrt_pvals$microbe)
da_vns   <- setNames(lrt_pvals$vns_signif, nm = lrt_pvals$microbe)
```

### Heat map
```{r, fig.width = 5, fig.height = 4.5}
m_coef_mat <- m_coef_df %>%
  filter(microbe %in% da_bugs) %>%
  mutate(microbe = gsub("_", " ", microbe)) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(contains("est")) %>%
  as.matrix()

colnames(m_coef_mat) <- c("Blast", "VNS", "Blast:VNS")

m_p_mat <- m_coef_df %>%
  filter(microbe %in% da_bugs) %>%
  mutate(microbe = gsub("_", " ", microbe)) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(contains("p")) %>%
  mutate(
    p_blast = case_when(p_blast < 0.001 ~ "**", p_blast < 0.05 ~ "*", .default = ""),
    p_vns   = case_when(p_vns   < 0.001 ~ "**", p_vns   < 0.05 ~ "*", .default = ""),
    p_bv    = case_when(p_bv    < 0.001 ~ "**", p_bv    < 0.05 ~ "*", .default = "")
    ) %>%
  as.matrix()

absmax <- max(abs(m_coef_mat))
col_fun <- colorRamp2(
  c(-absmax, 0, absmax),
  c("blue", "white", "red")
)

da_blast <- da_blast[rownames(m_coef_mat)]
da_vns   <- da_vns[rownames(m_coef_mat)]

blast_color <- c("TRUE" = "#0492c2", "FALSE" = "#FFFFFF00")
vns_color   <- c("TRUE" = "#FFc067", "FALSE" = "#FFFFFF00")
```

```{r}
lrt_ann_blast<- rowAnnotation(
  "LRT Blast FDR < 0.2" = da_blast,
  col = list("LRT Blast FDR < 0.2" = blast_color),
  show_annotation_name = FALSE,
  simple_anno_size = unit(0.2, "cm")
)

lrt_ann_vns<- rowAnnotation(
  "LRT VNS FDR < 0.2" = da_vns,
  col = list("LRT VNS FDR < 0.2" = vns_color),
  show_annotation_name = FALSE,
  simple_anno_size = unit(0.2, "cm")
)

# plot(lrt_ann_vns +lrt_ann_blast)
```

```{r, fig.width=6, fig.height=3.5}
ht <- Heatmap(m_coef_mat, name = "Estimate", col = col_fun,
  cluster_columns = FALSE,
  show_row_dend = FALSE,
  show_column_dend = FALSE,
  left_annotation = c(lrt_ann_vns, lrt_ann_blast),
  cluster_rows = TRUE,
  # row_split = da_vns,   # no VNS hits, so we don't need this
  # cluster_row_slices = TRUE,
  
  cell_fun = function(j, i, x, y, w, h, fill) {
    grid::grid.text(
      m_p_mat[i, j],
      x = x,
      y = y,
      gp = grid::gpar(
        fontsize = 14,
        col = "black",
        fontface = "bold"
      )
    )
  })

ht
```

```{r}
pdf("../figures/B_microbe_abundance_fdr20.pdf", width = 6, height = 3.5)
ht_drawn <- draw(ht)
# row_ord <- row_order(ht_drawn)
# da_bug_order <- rownames(trt_effect_matrix)[row_ord]
dev.off()
```


Romboutsia is generally increased by treatment, but especially blast
Clostridiales is generally increased by blast, but especially on VNS background
SGB41718 is generally decreased by blast, but especially on VNS background

Can also plot/test "presence/absence" if reviewers request, but the results did align well when I tested them for sparse features (as in, sparse bugs that were significant in the CLR regressions were actually even more significant if I did a presence/absence logistic regression, because CLR-transforming zeroes is just adding noise).

### Microbe  -> outcome (independent of exposure)
```{r}
outcome_hits <- map_dfr(model_outcome, function(df) {
  df %>% filter(term == "microbe", adj.p < 0.2)
}) %>% arrange(adj.p)

outcome_hits
```

These bugs may or may not be affected by treatment, but they explain additional variance in outcomes.
```{r}
form <- as.formula("startle_habituation ~ GGB28802_SGB41455 + blast + vns + blast_and_vns + (1 | Round_main)")
m    <- lmer(form, data = df, REML = FALSE, na.action = "na.exclude")
summary(m)
```

OK so SGB41455 is associated with increased startle_habituation (+0.402),
which is about the same effect size as blast, *but significant independent of blast*
```{r}
form <- as.formula("startle_PPI ~ GGB42598_SGB59794 + blast + vns + blast_and_vns + (1 | Round_main)")
m    <- lmer(form, data = df, REML = FALSE, na.action = "na.exclude")
summary(m)
```

I want to plot residual outcome vs. bug
!! Need to fix this by using lmer() instead of lm()
```{r}
residout_bug <- function(bug_name, data, outcome_name) {
  plt <- data.frame(
    row.names     = data$seq_name,
    bug_clr       = data[[bug_name]],
    inj_trt_group = data$inj_trt_group,
    blast         = data$blast,
    vns           = data$vns,
    blast_and_vns = data$blast_and_vns,
    Round_main    = data$Round_main,
    outcome       = data[[outcome_name]]
  ) %>%
    filter(!is.na(outcome))
  
  form <- as.formula("outcome ~ blast + vns + blast_and_vns + (1 | Round_main)")
  fit <- lmer(formula = form, data = plt)
  plt$out_resid <- resid(fit)
  
  ggplot(plt, aes(x = bug_clr, y = out_resid)) +     # fill = inj_trt_group, shape = Round_main
    geom_point(aes(color = inj_trt_group), size = 4, alpha = 0.7) +
    scale_fill_manual(values = injtrt_colors) +
    # scale_shape_manual(values = batch_shapes) +
    # geom_smooth(aes(color = inj_trt_group), method = "lm", se = FALSE, alpha = 0.2) +
    scale_color_manual(values = injtrt_colors) +
    geom_smooth(method = "lm", se = FALSE, color = "black",linetype = "dashed") +
    labs(
      x = paste0(gsub("_", " ", bug_name), " (24h)"),
      y = paste0("Residual ", outcome_name)
    ) +
    theme_bw() +
    theme(
      legend.position = "none",
      axis.text = element_text(size = 20),
      axis.title = element_text(size = 20)
      )
}
```
We could use bootstrapping to estimate these effects (slopes) as well. 

Save all these to a single pdf page
```{r}
n_plts <- nrow(outcome_hits)
n_cols <- 4
n_rows <- ceiling(n_plts / n_cols)

pdf_width <- 5 * n_cols
pdf_height <- 5 * n_rows

residout_plots <- vector("list", n_plts)

for (i in seq_len(nrow(outcome_hits))) {
  residout_plots[[i]] <- 
    residout_bug(outcome_hits[[i, "microbe"]], df, outcome_hits[[i, "outcome"]])
}

pdf("../figures/B_outcome_resid_plots.pdf", width = pdf_width, height = pdf_height)

grid.arrange(
  grobs = residout_plots,
  ncol = n_cols   # adjust layout
)

dev.off()
```
Function to split pdfs into columns
```{r}
pdf_split <- function(
    hit_df, data_df,
    sub_height, sub_width,
    n_cols,
    plot_function,
    filename
    ) {
  
  n_plts <- ifelse(
    plot_function == "box_easier", yes = length(hit_df),
    no = nrow(hit_df)
  )
  
  n_rows <- ceiling(n_plts / n_cols)
  
  pdf_width <- sub_width * n_cols
  pdf_height <- sub_height * n_rows
  
  plot_vector <- vector("list", n_plts)
  
  if (plot_function == "residout_bug") {
    for (i in seq_len(nrow(hit_df))) {
      plot_vector[[i]] <- residout_bug(
        bug_name     = hit_df[[i, "microbe"]],
        data         = data_df,
        outcome_name = hit_df[[i, "outcome"]]
      )}
    
  } else if (plot_function == "residout_residbug") {
    for (i in seq_len(nrow(hit_df))) {
      plot_vector[[i]] <- residout_residbug(
        bug_name     = hit_df[[i, "mediator"]],
        data         = data_df,
        outcome_name = hit_df[[i, "outcome"]]
      )
    }
    
  } else if (plot_function == "box_easier") {
    for (i in 1:length(hit_df)) {  # hit_df can be a vector here
      bug_names <- hit_df %>%
        str_replace_all("sp", "sp.") %>%
        str_replace_all("_", " ")
      
      plot_vector[[i]] <- box_easier(
        data_df,
        hit_df[i],
        bug_names[i]
      )
    }
    
  }
  
  pdf(filename, width = pdf_width, height = pdf_height)
  grid.arrange(grobs = plot_vector, ncol = n_cols)
  dev.off()
}
```

Function works for `residout_bug``
```{r}
pdf_split(
    hit_df        = outcome_hits,
    data_df       = df,
    sub_height    = 5,
    sub_width     = 5,
    n_cols        = 4,
    plot_function = "residout_bug",
    filename      = "../figures/B_outcome_resid_plots.pdf"
    )
```

### Exposure -> microbe -> outcome 
```{r}
med_hits <- all_run_df %>% filter(!is.na(signif_at))
med_hits
```

#### Group-wise box-plot (microbes)
```{r, fig.width = 5, fig.height = 5}
bugs_to_plot <- unique(c(da_bugs, outcome_hits$microbe, med_hits$mediator))
# bug_names <- bugs_to_plot %>%
#   str_replace_all("sp", "sp.") %>%
#   str_replace_all("_", " ")

# bug_boxplots <- list()

pdf_split(
    hit_df        = bugs_to_plot,
    data_df       = df,
    plot_function = "box_easier",
    filename      = "../figures/B_microbe_abundance.pdf",
    sub_height = 5, sub_width = 5, n_cols = 4,
    )

```

Save all these to a single pdf
```{r}
# # Open the PDF device
# pdf("../figures/B_microbe_plots.pdf", width = 5, height = 5) # You can adjust dimensions
# 
# # each loop adds a new pdf page
# for (i in 1:length(bugs_to_plot)) {
#   bug_boxplots[[i]] <- box_easier(df, bugs_to_plot[i], bug_names[i])
#   print(bug_boxplots[[i]])
# }
# 
# # Close the PDF device to save
# dev.off()
```


## ACME Bootstrap CI
```{r}
med_hits <- all_run_df %>% filter(!is.na(signif_at))

med_ci_input <- med_hits %>%
  transmute(exposure, mediator, outcome) %>%
  mutate(
    exposure = gsub("pm_", "", exposure),  # this assumes no exposures are "vns_on_blast"
    condition = case_when(
      exposure == "blast" ~ "vns",
      exposure == "vns" ~ "blast"
    )
  ) %>%
  split(seq_len(nrow(.))) %>% 
  map(as.list)

med_ci_input[[1]]
```
I need to stratify by `batch` to include as a random effect.
However, I should *also* stratify by `inj_trt_group` to ensure I have a balanced sample every time.

Which of these bugs actually had batch effects?
```{r}
lmer_list <- lapply(med_ci_input, function(x) {
  form_m <- as.formula(paste0(x$mediator," ~ blast + vns + blast_and_vns + (1 | Round_main)"))
  form_y <- as.formula(paste0(
    x$outcome, " ~ ", x$mediator," + blast + vns + blast_and_vns + (1 | Round_main)")
    )
  m <- lmer(form_m, data = df, REML = FALSE, na.action = "na.exclude")
  y <- lmer(form_y, data = df, REML = FALSE, na.action = "na.exclude")
  
  list("M" = m, "Y" = y)
})

singularity_df <- map2_dfr(
  lmer_list,
  med_ci_input,
  function(mods, info) {
    tibble(
      mediator = info$mediator,
      outcome  = info$outcome,
      M_singular = lme4::isSingular(mods$M),
      Y_singular = lme4::isSingular(mods$Y)
    )
  }
)
```


Convergence was OK on all of these, so I think I'm good.
```{r}
# lmer_list
summary(lmer_list$`2`$M)
summary(lmer_list$`2`$Y)
```

Resample within (batch x inj_trt_group) 
```{r}
n_hits <- length(med_ci_input)
n_workers <- ifelse(n_hits < 15, n_hits, 15)  # no point in having more workers than jobs

plan(multisession, workers = n_workers)

blast_ci_list <- future_lapply(med_ci_input, function(x) {
  strat_boot_lmer(
    data     = df,
    exposure = x$exposure,
    mediator = x$mediator,
    outcome  = x$outcome,
    n_boot = 999)
}, future.seed = 125)

ci_list <- blast_ci_list
ci_df <- bind_rows(ci_list)

saveRDS(ci_df, "/proj/gibbons/2024_easton_vns/data/3_newest/B_bootCI_df_0216.rds")
ci_df
# saveRDS(ci_df, "/proj/gibbons/2024_easton_vns/data/mediation/B_bootCI_blocked_df_0109.rds")
```
We had to do it this way (preserve sample size within batch) because there is a variance component to the random effect (1|Round_main), so it could become unstable if n dropped too low for a single batch.

#### troubleshooting things
testdata <- case_bootstrap_within_clust(data, "Round_main")

```{r}
x <- med_ci_input[[1]]
data <- df
exposure <- x$exposure
mediator <- x$mediator
outcome <- x$outcome
n_boot <- 5

test <- strat_boot_lmer(data, exposure, mediator, outcome, n_boot = 5)


data$treat_cell <- interaction(data$blast, data$vns, drop = TRUE)
data$strat <- interaction(data$treat_cell, data$Round_main, drop = TRUE)

# Pre-split once
boot_prep   <- prep_group_bootstrap(data, "strat")  # need to split on BOTH "treat_cell" AND "batch"
split_data  <- boot_prep$split_data
group_sizes <- boot_prep$group_sizes

as.vector(group_sizes)
```


```{r}
# formulas
form_m <- as.formula(paste0(mediator, " ~ blast + vns + blast_and_vns + (1 | Round_main)"))
form_y <- as.formula(paste0(outcome, " ~ ", mediator, " + blast + vns + blast_and_vns + (1 | Round_main)"))

# Storage matrix
boot_mat <- matrix(NA_real_, n_boot, 6)
colnames(boot_mat) <- c(
  "ACME_cond0", "ACME_cond1",
  "ADE_cond0",  "ADE_cond1",
  "Total_cond0","Total_cond1"
)

# for (i in seq_len(n_boot)) {
  
i <- 1
boot_data <- resample_groups(split_data, group_sizes) # need to resample within "treat_cell" and "batch" combinations 

fit_m <- lmer(form_m, boot_data, REML = FALSE)
fit_y <- lmer(form_y, boot_data, REML = FALSE)

fe_m <- fixef(fit_m)
fe_y <- fixef(fit_y)

# Check required coefficients exist
needed_m <- c(exposure, "blast_and_vns")
needed_y <- c(exposure, mediator, "blast_and_vns")

if (!all(needed_m %in% names(fe_m)) ||
    !all(needed_y %in% names(fe_y))) {
  next
}

b <- fe_y[mediator]

a_cond0 <- fe_m[exposure]
a_cond1 <- fe_m[exposure] + fe_m["blast_and_vns"]

c_cond0 <- fe_y[exposure]
c_cond1 <- fe_y[exposure] + fe_y["blast_and_vns"]

ACME_0  <- a_cond0 * b
ACME_1  <- a_cond1 * b

ADE_0   <- c_cond0
ADE_1   <- c_cond1

boot_mat[i, ] <- c(
  ACME_0,            # ACME cond 0
  ACME_1,
  ADE_0,             # ADE cond 0
  ADE_1,
  ACME_0 + ADE_0,    # total cond 0
  ACME_1 + ADE_1
)
# }

boot_df <- as_tibble(boot_mat) |> drop_na()

boot_df |>
  summarise(
    across(
      everything(),
      list(
        est   = mean,
        lower = ~ quantile(.x, 0.025),
        upper = ~ quantile(.x, 0.975)
      ),
      .names = "{.col}_{.fn}"
    )
  ) |>
  mutate(
    exposure = exposure,
    mediator = mediator,
    outcome  = outcome
  )
```



## Plot coefficients for mediators + annotate with p-values
  Get the order of microbes
### new code
```{r}
med_bugs <- unique(med_hits$mediator) # luckily they are all unique

med_bugs <- med_hits %>%
  filter(signif_at <= 0.2) %>%
  pull(mediator)

m_coef_mat <- m_coef_df %>%
  filter(microbe %in% med_bugs) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(contains("est")) %>%
  as.matrix()

m_p_mat <- m_coef_df %>%
  filter(microbe %in% med_bugs) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(contains("p")) %>%
  mutate(
    p_blast = case_when(p_blast < 0.001 ~ "**", p_blast < 0.05 ~ "*", .default = ""),
    p_vns   = case_when(p_vns   < 0.001 ~ "**", p_vns   < 0.05 ~ "*", .default = ""),
    p_bv    = case_when(p_bv    < 0.001 ~ "**", p_bv    < 0.05 ~ "*", .default = "")
    ) %>%
  as.matrix()

y_coef_mat <- y_coef_df %>%
  semi_join(med_hits, by = c("outcome", "microbe" = "mediator")) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(contains("est")) %>%
  as.matrix()

y_p_mat <- y_coef_df %>%
  semi_join(med_hits, by = c("outcome", "microbe" = "mediator")) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(contains("p")) %>%
  mutate(
    p_blast = case_when(p_blast < 0.001 ~ "**", p_blast < 0.05 ~ "*", .default = ""),
    p_vns   = case_when(p_vns   < 0.001 ~ "**", p_vns   < 0.05 ~ "*", .default = ""),
    p_bv    = case_when(p_bv    < 0.001 ~ "**", p_bv    < 0.05 ~ "*", .default = ""),
    p_microbe = case_when(p_microbe < 0.001 ~ "**", p_microbe    < 0.05 ~ "*", .default = "")
    ) %>%
  as.matrix()

row_order <- rownames(m_coef_mat)
y_coef_mat <- y_coef_mat[row_order, c(colnames(m_coef_mat), "est_microbe")]
y_p_mat <- y_p_mat[row_order, c(colnames(m_p_mat), "p_microbe")]

row_outcome <- med_hits %>%
  distinct(mediator, outcome) %>%
  filter(mediator %in% rownames(m_coef_mat)) %>%
  tibble::deframe()

row_exposure <- med_hits %>%
  distinct(mediator, exposure) %>%
  filter(mediator %in% rownames(m_coef_mat)) %>%
  tibble::deframe()

row_fdr <- med_hits$signif_at
names(row_fdr) <- med_hits$mediator

row_outcome <- row_outcome[rownames(m_coef_mat)]
row_exposure <- row_exposure[rownames(m_coef_mat)]
row_fdr <- row_fdr[rownames(m_coef_mat)]


stopifnot(all(names(row_outcome) == rownames(m_coef_mat)))
stopifnot(all(names(row_outcome) == rownames(y_coef_mat)))
stopifnot(all(names(row_fdr) == rownames(row_outcome)))

# fdr_fun <- colorRamp2(c(0.00, 0.30), c("darkgreen", "lightyellow"))
fdr_value_vector <- rev(c(0.1, 0.125, 0.15, 0.175, 0.20))

fdr_colors <- structure(
  RColorBrewer::brewer.pal(length(fdr_value_vector), "Greens"),
  names = fdr_value_vector
  )

absmax <- max(abs(c(m_coef_mat, y_coef_mat)))
col_fun <- colorRamp2(
  c(-absmax, 0, absmax),
  c("blue", "white", "red")
)

exposure_color <- c("pm_blast" = "#0492c2", "pm_vns" = "#FFc067")
```

Annotation first
```{r}
fdr_annotation <- rowAnnotation(
  "FDR <" = row_fdr,
  col = list("FDR <" = fdr_colors),
  show_annotation_name = FALSE,
  simple_anno_size = unit(0.2, "cm")
)

which_exposure <- rowAnnotation(
  "Exposure" = row_exposure,
  col = list("Exposure" = exposure_color),
  show_annotation_name = FALSE,
  simple_anno_size = unit(0.2, "cm")
)

plot(fdr_annotation + which_exposure)
```


```{r}
# heatmap model M coefficients
ht_m <- Heatmap(
  m_coef_mat,
  name = "M Estimate",
  col = col_fun,
  left_annotation = c(fdr_annotation, which_exposure),
  
  row_split = row_exposure,
  cluster_rows = TRUE,
  cluster_row_slices = TRUE,
  cluster_columns = FALSE,
  show_row_dend = FALSE,
  show_column_dend = FALSE,
  cell_fun = function(j, i, x, y, w, h, fill) {
    grid::grid.text(
      m_p_mat[i, j],
      x = x,
      y = y,
      gp = grid::gpar(
        fontsize = 14,
        col = "black",
        fontface = "bold"
      )
    )
  })

ht_m
```

```{r}
# heatmap model Y coefficients
ht_y <- Heatmap(
  y_coef_mat,
  name = "Y Estimate",
  col = col_fun,
  
  row_split = row_exposure,
  cluster_rows = TRUE,
  cluster_row_slices = TRUE,
  
  cluster_columns = FALSE,
  show_row_dend = FALSE,
  show_column_dend = FALSE,
  cell_fun = function(j, i, x, y, w, h, fill) {
    grid::grid.text(
      y_p_mat[i, j],
      x = x,
      y = y,
      gp = grid::gpar(
        fontsize = 14,
        col = "black",
        fontface = "bold"
      )
    )
  })

ht_y
```


```{r, fig.width = 8, fig.height = 4.5}
ht_list <- ht_m + ht_y

draw(ht_list, main_heatmap = "M Estimate")
draw(ht_list, main_heatmap = "Y Estimate")
```

Now save
```{r}
pdf("../figures/B_med_coef_heatmap.pdf", width = 8, height = 4.5)
ht_drawn <- draw(ht_list, main_heatmap = "M Estimate")
rows <- row_order(ht_drawn)
dev.off()
```

## Plotting blast mediation hits
*Doing this after heat maps so I can use same order as clustering*
```{r}
ht_row_order <- c(unlist(rows))
names(ht_row_order) <- NULL
ht_row_order

ci_row_order <- row_order[ht_row_order]
ci_row_order
```


```{r}
ci_plt <- ci_df %>%
  filter(mediator %in% ci_row_order) %>%
  mutate(
    exposure = factor(exposure, levels = c("blast", "vns")),
    mediator = factor(mediator, levels = ci_row_order),     # don't reverse
    outcome  = factor(outcome, levels = c("OFB_Clatency", "OFB_Cdistance", "PD_choice_filt", "EtOH_pref_24h_final")),
    exp_med_out  = as.character(interaction(exposure, mediator, outcome, sep = ";"))
    ) %>%
  pivot_longer(
    cols = matches("^(ACME|ADE|Total)_"),
    names_to = c("effect", "cond", ".value"),
    names_pattern = "^([^_]+)_([^_]+)_(.+)$"
  ) %>%
  mutate(
    effect = factor(effect, levels = c("ACME", "ADE", "Total")),
    cond   = factor(cond, levels = c("cond0", "cond1")),
    # exp_med_out = factor(paste0(exposure, ";", outcome, ";", mediator)),
    exp_med_out_cond = factor(paste0(exp_med_out, ";", cond))
    # microbe = factor(mediator)
  ) %>%
  arrange(mediator)

shade_bands <- ci_plt %>%
  group_by(mediator) %>% 
  arrange(by = mediator) %>%
  mutate(
    band_id = as.numeric(mediator),
    fill = ifelse(band_id %% 2 == 1, "grey95", "transparent")
    )

effect_colors <- c(
  "ADE"   = "#9C2121",
  "ACME"  = "#689835",
  "Total" = "#070E00"
)
```

```{r, fig.width = 7, fig.height = 3.5}

plt <- ggplot(ci_plt, aes(x = est, y = effect, color = effect)) +
  geom_rect(data = shade_bands, aes(
      xmin = -Inf, xmax = Inf,
      ymin = -Inf, ymax = Inf,
      fill = fill
      ), inherit.aes = FALSE, color = NA) +
  geom_point(size = 1.5) +
  geom_linerange(aes(xmin = lower, xmax = upper), linewidth = 0.4) +
  scale_fill_identity() +
  # geom_errorbarh(aes(xmin = mean - se, xmax = mean + se), linewidth = 0.3, height = 0.5) +
  scale_color_manual(values = effect_colors) +
  facet_grid(
    rows = vars(mediator), # if multiple exposure/outcomes, use exp_med_out
    cols   = vars(cond),
    space  = "free_y",
    # switch = "y", scales = "free_y",
    ) +
  # scale_x_continuous(limits = c(xmin, xmax)) +
  labs(x = NULL, y = NULL) +
  guides(fill = guide_legend(reverse = TRUE)) +
  theme_minimal(base_size = 14) +
  theme(
    strip.placement   = "outside",
    panel.spacing.y   = unit(0.1, "lines"),   # try 1-3 "lines" of spacing
    panel.spacing.x   = unit(0.1, "lines"),   # put close together
    axis.text.y       = element_blank(),
    axis.ticks.y      = element_blank(),
    axis.ticks.x      = element_line(color = "black", linewidth = 0.3),
    axis.line.x       = element_line(colour = "black", linewidth = 0.4),
    panel.grid        = element_blank(),
    panel.background  = element_blank()
  )

plt1 <- plt + theme(
  legend.position    = "none",
  strip.text.y.right = element_text(
    angle = 0, margin = margin(l = 4, unit = "pt"), hjust = 0
  ))

plt2 <- plt + theme(legend.position = "none", strip.text.y.right = element_blank())

plt_leg <- plt + theme(strip.text.y.right = element_blank())
legend_only <- get_legend(plt_leg)

plt1
```

do as device instead (so I can save plot with legend)
```{r}
row_height  <- 0.5
panel_width <- 2

pdf("../figures/B_med_blast_cond_blocked.pdf",
    width = (panel_width*2) + 3,
    height = (nlevels(shade_bands$mediator) * row_height))

print(plt1)
plot(legend_only)
dev.off()
```




#### Also plot residual outcome vs. residual bug for the mediators
I want to plot residual outcome vs. bug
```{r}
residout_residbug <- function(bug_name, data, outcome_name) {
  plt <- data.frame(
    row.names     = data$seq_name,
    bug_clr       = data[[bug_name]],
    inj_trt_group = data$inj_trt_group,
    blast         = data$blast,
    vns           = data$vns,
    blast_and_vns = data$blast_and_vns,
    Round_main    = data$Round_main,
    outcome       = data[[outcome_name]]
  ) %>%
    filter(!is.na(outcome))
  
  residy <- lmer(as.formula("outcome ~ blast + vns + blast_and_vns + (1|Round_main)"), data = plt, REML = FALSE)
  residx <- lmer(as.formula("bug_clr ~ blast + vns + blast_and_vns + (1|Round_main)"), data = plt, REML = FALSE)
  plt$out_resid <- resid(residy)
  plt$bug_resid <- resid(residx)
  
  ggplot(plt, aes(x = bug_resid, y = out_resid)) +     # fill = inj_trt_group, shape = Round_main
    geom_point(aes(color = inj_trt_group), size = 4, alpha = 0.7) +
    # scale_fill_manual(values = injtrt_colors) +
    # scale_shape_manual(values = batch_shapes) +
    # geom_smooth(aes(color = inj_trt_group), method = "lm", se = FALSE, alpha = 0.2) +
    scale_color_manual(values = injtrt_colors) +
    geom_smooth(method = "lm", se = FALSE, color = "black",linetype = "dashed") +
    labs(
      x = paste("Residual \n", gsub("_", " ", bug_name)),  # use expression(italic()) to italicize
      y = paste0("Residual ", gsub("acute_cyto_", "", outcome_name))
    ) +
    theme_bw() +
    theme(
      legend.position = "none",
      axis.text = element_text(size = 20),
      axis.title = element_text(size = 20)
      )
}
```

```{r, fig.width = 5, fig.height = 5.4}
testplot <- residout_residbug(med_hits[[1, "mediator"]], df, med_hits[[1, "outcome"]])
print(testplot)
```


Save all these to a single pdf
```{r, fig.width = 4, fig.height = 4}
# resid_resid_plots <- list()
# pdf("../figures/A_outcome_resid_resid_plots.pdf", width = 5, height = 5)
# 
# for (i in 1:nrow(med_hits)) {
#   resid_resid_plots[[i]] <- residout_residbug(med_hits[[i, "mediator"]], df, med_hits[[i, "outcome"]])
#   print(resid_resid_plots[[i]])
# }
# 
# # Close the PDF device to save
# dev.off()

pdf_split(
  hit_df        = med_hits,
  data_df       = df,
  plot_function = "residout_residbug",
  filename      = "../figures/B_outcome_resid_resid_plots.pdf",
  sub_height    = 5.4,
  sub_width     = 5.0,
  n_cols        = 4
)
```




### Plotting blast mediation hits (OLD)
```{r}
ci_plt <- ci_df %>%
  pivot_longer(
    cols = matches("^(ACME|ADE|Total)_"),
    names_to = c("effect", "cond", ".value"),
    names_pattern = "^([^_]+)_([^_]+)_(.+)$"
  ) %>%
  mutate(
    effect = factor(effect, levels = c("ACME", "ADE", "Total")),
    cond   = factor(cond, levels = c("cond0", "cond1")),
    exp_med_out = factor(paste0(exposure, ";", outcome, ";", mediator)),
    exp_med_out_cond = factor(paste0(exp_med_out, ";", cond)),
    microbe = factor(mediator)
  ) %>%
  arrange(effect)

shade_bands <- ci_plt %>%
  group_by(exp_med_out) %>% 
  arrange(by = exp_med_out) %>%
  mutate(
    band_id = as.numeric(exp_med_out),
    fill = ifelse(band_id %% 2 == 1, "grey95", "transparent")
    )

effect_colors <- c(
  "ADE"   = "#9C2121",
  "ACME"  = "#689835",
  "Total" = "#070E00"
)
```

```{r, fig.width = 7, fig.height = 3}
plt <- ggplot(ci_plt, aes(x = est, y = effect, color = effect)) +
  geom_rect(data = shade_bands, aes(
      xmin = -Inf, xmax = Inf,
      ymin = -Inf, ymax = Inf,
      fill = fill
      ), inherit.aes = FALSE, color = NA) +
  geom_point(size = 1.5) +
  geom_linerange(aes(xmin = lower, xmax = upper), linewidth = 0.4) +
  scale_fill_identity() +
  # geom_errorbarh(aes(xmin = mean - se, xmax = mean + se), linewidth = 0.3, height = 0.5) +
  scale_color_manual(values = effect_colors) +
  facet_grid(
    rows = vars(exp_med_out), # if multiple exposure/outcomes, use exp_med_out
    cols   = vars(cond),
    space  = "free_y",
    # switch = "y", scales = "free_y",
    ) +
  # scale_x_continuous(limits = c(xmin, xmax)) +
  labs(x = NULL, y = NULL) +
  guides(fill = guide_legend(reverse = TRUE)) +
  theme_minimal(base_size = 14) +
  theme(
    strip.placement   = "outside",
    panel.spacing.y   = unit(0.1, "lines"),   # try 1-3 "lines" of spacing
    panel.spacing.x   = unit(0.1, "lines"),   # put close together
    axis.text.y       = element_blank(),
    axis.ticks.y      = element_blank(),
    axis.ticks.x      = element_line(color = "black", linewidth = 0.3),
    axis.line.x       = element_line(colour = "black", linewidth = 0.4),
    panel.grid        = element_blank(),
    panel.background  = element_blank()
  )

plt1 <- plt + theme(
  legend.position    = "none",
  strip.text.y.right = element_text(
    angle = 0, margin = margin(l = 4, unit = "pt"), hjust = 0
  ))

plt2 <- plt + theme(legend.position = "none", strip.text.y.right = element_blank())

plt_leg <- plt + theme(strip.text.y.right = element_blank())
legend_only <- get_legend(plt_leg)

plt1
```


do as device instead (so I can save plot with legend)
```{r}
row_height  <- 0.5
panel_width <- 2

pdf("../figures/B_med_cond_cases.pdf",
    width = (panel_width*2) + 3,
    height = (nlevels(shade_bands$exp_med_out) * row_height))

print(plt1)
plot(legend_only)
dev.off()
```


### Make heatmap to go along with this (OLD)

```{r, fig.width = 5, fig.height = 4}
order_bugs <- c(
  "bacterium_1XD42_54", "Clostridiales_bacterium", "GGB29002_SGB41718",
  "GGB3171_SGB4185", "GGB28898_SGB41580"
)

trt_effect_matrix <- model_microbe %>%
  filter(microbe %in% med_hits$mediator) %>%
  mutate(microbe = factor(microbe, levels = order_bugs)) %>%
  dplyr::select(microbe, term, estimate) %>%
  pivot_wider(names_from = term, values_from = estimate) %>%
  arrange(microbe) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(c("blast", "vns", "blast_and_vns")) %>%
  as.matrix()

trt_p_matrix <- model_microbe %>%
  filter(microbe %in% med_hits$mediator) %>%
  mutate(
    microbe = factor(microbe, levels = order_bugs),
    p.ann = case_when(
      p.value < 0.005 ~ "**",
      p.value < 0.05 ~ "*",
      .default = ""
      )
    ) %>%
  dplyr::select(microbe, term, p.ann) %>%
  pivot_wider(names_from = term, values_from = p.ann) %>%
  arrange(microbe) %>%
  column_to_rownames("microbe") %>%
  as.matrix()

absmax <- max(abs(trt_effect_matrix))
col_fun <- colorRamp2(
  c(-absmax, 0, absmax),
  c("blue", "white", "red")
)

ht <- Heatmap(trt_effect_matrix, name = "Estimate", col = col_fun,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_dend = FALSE,
  show_column_dend = FALSE,
  row_order = order_bugs,
  cell_fun = function(j, i, x, y, w, h, fill) {
    grid::grid.text(
      trt_p_matrix[i, j],
      x = x,
      y = y,
      gp = grid::gpar(
        fontsize = 14,
        col = "black",
        fontface = "bold"
      )
    )
  })
```


```{r, fig.width = 5, fig.height = 5}
ht
```

```{r}
pdf("../figures/med_coef_heatmap.pdf", width = 5, height = 5)
ht_drawn <- draw(ht)
dev.off()
```

#### Also plot residual outcome vs. residual bug for the mediators
I want to plot residual outcome vs. bug
```{r}
residout_residbug <- function(bug_name, data, outcome_name) {
  plt <- data.frame(
    row.names     = data$seq_name,
    bug_clr       = data[[bug_name]],
    inj_trt_group = data$inj_trt_group,
    blast         = data$blast,
    vns           = data$vns,
    blast_and_vns = data$blast_and_vns,
    Round_main    = data$Round_main,
    outcome       = data[[outcome_name]]
  ) %>%
    filter(!is.na(outcome))
  
  formx <- as.formula("bug_clr ~ blast + vns + blast_and_vns + (1 | Round_main)")
  formy <- as.formula("outcome ~ blast + vns + blast_and_vns + (1 | Round_main)")
  fitx <- lmer(formula = formx, data = plt)
  fity <- lmer(formula = formy, data = plt)
  
  plt$bug_resid <- resid(fitx)
  plt$out_resid <- resid(fity)
  
  ggplot(plt, aes(x = bug_resid, y = out_resid)) +     # fill = inj_trt_group, shape = Round_main
    geom_point(aes(color = inj_trt_group), size = 4, alpha = 0.7) +
    scale_fill_manual(values = injtrt_colors) +
    # scale_shape_manual(values = batch_shapes) +
    # geom_smooth(aes(color = inj_trt_group), method = "lm", se = FALSE, alpha = 0.2) +
    scale_color_manual(values = injtrt_colors) +
    geom_smooth(method = "lm", se = FALSE, color = "black",linetype = "dashed") +
    labs(
      x = paste0(bug_name, " Abundance"),
      y = paste0("Residual normalized ", outcome_name)
    ) +
    theme_bw() + theme(legend.position = "none")
}
```

Save all these to a single pdf
```{r, fig.width = 4, fig.height = 4}
resid_resid_plots <- list()
pdf("../figures/B_outcome_resid_resid_plots.pdf", width = 5, height = 5)

for (i in 1:nrow(med_hits)) {
  resid_resid_plots[[i]] <- residout_residbug(med_hits[[i, "mediator"]], df, med_hits[[i, "outcome"]])
  print(resid_resid_plots[[i]])
}

# Close the PDF device to save
dev.off()
```


##### Minimum microbe p-value for each outcome?
```{r, fig.width = 4, fig.height = 4}
for (out in model_outcome) {
  p_df <- data.frame(
    microbe = out$microbe[out$term == "microbe"],
    p.value = out$p.value[out$term == "microbe"]
  )
  
  min_p <- min(p_df$p.value)
  outcome_name <- unique(out$outcome)
  
  p_hist <- ggplot(data.frame(p_df), aes(x = p.value)) +
    geom_histogram(bins = 20, color = "white") +
    labs(
      title = paste0(outcome_name, ", min p = ", round(min_p, digits = 4)),
      x     = "p_microbe (outcome ~ microbe + treatments)",
      y     = "count"
    ) + theme_bw()
  
  print(p_hist)
}
```