---
title: "cytokine_PCA.Rmd"
author: "Alyssa N. Easton"
date: '2025-10-03'
output: 
---
Edited to only contain cytokines (from `PCA_final.Rmd`) and use new microbiome data *2025-10-03*

# Setup
Do PCA of cytokines (cohort A) and behaviors (cohort B) to construct composite scores.
```{r "setup", include=FALSE}

knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
knitr::opts_knit$set(root.dir = file.path(Sys.getenv("VNS_ROOT"), "data"))  # stored VNS_ROOT in ~/.Renviron
options(digits = 10) ## Formats output to 10 digits
library(tidyverse)
library(data.table)
library(magrittr)
library(patchwork)
library(stats)
library(broom)
library(emmeans)
library(RColorBrewer)
library(cowplot)
library(openxlsx)
library(emmeans)
library(car)

# my own modules in behavior_cytokine_PCA.Rmd if I want them back
# for now, I'm just using functions

source("/proj/gibbons/2024_easton_vns/scripts/PCA_final_functions.R")
source("/proj/gibbons/2024_easton_vns/scripts/shannon_div_functions.R")

```

# Tidy data import/PCA using tidyverse/broom
Here, I'm using a "helper" function `run_.pca.join_pca` to automatically
   • filter to one cohort
   • select & scale a block of columns
   • run PCA
   • pick PCs ≥ threshold
   • augment metadata with those PCs

The resulting object is a list, including
   • meta: augmented metadata df including PCs ≥ threshold `res$meta$PCx`
   • pca: the output of prcomp() `res$pca`
   • variance: the variance explained per PC `res$variance`
   • loadings: loadings of each feature in each PC `res$loadings`

```{r}
micro <- readRDS("/proj/gibbons/2024_easton_vns/data/1_preprocessed/species_october.rds")


# 2. Run PCA for all serum + brain cytokines, combined
res_A  <- run_pca(micro$a_meta, block = "serum|brain")
res_A$loadings$plot_name <- res_A$loadings$Feature %>%
  gsub("serum_", "", .) %>%
  gsub("brain_", "", .) %>%
  gsub("gamma", "g", .) %>%
  gsub("INF", "IFN", .)

# saveRDS(res_A, "PCA/pca_A_cytokines.rds")
PC1_loadings <- res_A$loadings %>%
  dplyr::select(c(Feature, PC1))

all_load <- res_A$loadings %>%
  dplyr::select(c(Feature, PC1, PC2, PC3, PC4, PC5))

# write_csv(all_load, file = "/proj/gibbons/2024_easton_vns/data/PCA/cytokine_PCA_loadings.csv")
```

## Initial plotting of variance explained, PC1 and 2
### Variance explained per PC
```{r}
var_explained <- plot_pca_variance(res_A) + theme_minimal()
var_explained
```
```{r}
res_A$vars
```


```{r}
res_A$vars[[1, 2]] + res_A$vars[[2, 2]]
```
For cytokines (n = 27), the top 2 PCs account for 53% of variance

### Simple vectors/biplot of PC1 and PC2
```{r}
vectors <- plot_pca_vectors(res_A, i_vectors = 30)
vectors

biplot <- plot_pca_biplot(res_A, "inj_trt_group", four_colors, i_vectors = 0) + theme_classic()
biplot

```

- Sean said we can even look at PCs that explain ~5% variance
```{r}
ck_pc_df <- res_A$vars %>%
  filter(`Proportion of Variance` >= 0.05)

ck_pcs <- ck_pc_df %>% pull(PC)

ck_pc_var <- ck_pc_df %>% pull(`Proportion of Variance`)
names(ck_pc_var) <- ck_pcs
```

### 2-way ANOVA
```{r}
ck_anova_list <- map(ck_pcs, ~ {
  f <- reformulate(c("Injury*Treatment"), response = .x)
  aov(f, data = res_A$meta) %>%
    tidy() %>%
    mutate(across(c(sumsq, meansq, statistic), ~ round(.x, 3)),
           p.value = round(p.value, 8))
})

names(ck_anova_list) <- ck_pcs
anova_df <- map_dfr(ck_anova_list, bind_rows, .id = "PC")
```

Do post-hoc testing for these 5 things.
```{r}
ck_posthoc_list <- map(ck_pcs, ~ {
  post_hoc(.x, res_A$meta)
})

names(ck_posthoc_list) <- ck_pcs
posthoc_df <- map_dfr(ck_posthoc_list, bind_rows, .id = "PC")
```

```{r}
anova_df %>%
  filter(term != "Residuals") %>%
  transmute(PC, term, statistic, p.value) %>%
  pivot_wider(
    id_cols = "PC", names_from = "term",
    values_from = c("statistic", "p.value"),
    names_glue = "{.value}_{term}"
  )
```
# Do Levene test for dispersion differences

Functions
```{r}
# plotting functions
plot_hist <- function(df) {
  pc_name <- unique(df$PC)
  
  ggplot(df, aes(x = dev)) +
    geom_histogram(bins = 20) +
    facet_wrap(~inj_trt_group, scales = "free") +
    ggtitle(pc_name) +
    theme_classic()
}

plot_qq <- function(df) {
  pc_name <- unique(df$PC)
  ggplot(df, aes(sample = dev)) +
    stat_qq() +
    stat_qq_line() +
    facet_wrap(~inj_trt_group) +
    ggtitle(pc_name) +
    theme_classic()
}

plot_box <- function(df) {
  pc_name <- unique(df$PC)
  ggplot(df, aes(x = inj_trt_group, y = dev)) +
    geom_boxplot() +
    ggtitle(pc_name) +
    theme_classic()
}
```

First, look at deviation distributions  
```{r}
dev_df <- res_A$meta %>%
  select(all_of(c("inj_trt_group", ck_pcs))) %>%
  pivot_longer(cols = all_of(ck_pcs), names_to = "PC", values_to = "value") %>%
  group_by(PC, inj_trt_group) %>%
  mutate(dev = abs(value - median(value, na.rm = TRUE))) %>%
  ungroup()

pc_split <- dev_df %>%
  group_split(PC) %>%
  set_names(unique(dev_df$PC))

hist_list <- map(pc_split, plot_hist)
qq_list   <- map(pc_split, plot_qq)
box_list  <- map(pc_split, plot_box)
```

```{r, fig.width=5, fig.height=15}
combined_plots <- map(pc_split, ~ {
  plot_hist(.x) / plot_qq(.x) / plot_box(.x)
})

walk2(combined_plots, names(combined_plots), ~ {
  print(.x + plot_annotation(title = .y))
})
```
## Tried Levene test, but it was underpowered here
### Test Levene on one PC
```{r}
f <- as.formula("PC5 ~ inj_trt_group")    # reformulate(c("Injury*Treatment"), response = "PC5")
car::leveneTest(f, data = res_A$meta, center = median)


    as.data.frame() %>%
    rownames_to_column("term") %>%
    as_tibble() %>%
    rename(
      df = Df,
      statistic = `F value`,
      p.value = `Pr(>F)`
    ) %>%
    mutate(
      across(c(statistic), ~ round(.x, 3)),
      p.value = round(p.value, 8)
    )
```
### Do Levne test on all PCs
```{r}
ck_levene_list <- map(ck_pcs, ~ {
  f <- reformulate(c("Injury*Treatment"), response = .x)
  
  car::leveneTest(f, data = res_A$meta, center = median) %>%
    as.data.frame() %>%
    rownames_to_column("term") %>%
    as_tibble() %>%
    rename(
      df = Df,
      statistic = `F value`,
      p.value = `Pr(>F)`
    ) %>%
    mutate(
      across(c(statistic), ~ round(.x, 3)),
      p.value = round(p.value, 8)
    )
})
names(ck_levene_list) <- ck_pcs

ck_levene_posthoc_list <- map(ck_pcs, ~ {
  levene_posthoc(.x, res_A$meta)
})
names(ck_levene_posthoc_list) <- ck_pcs

levene_df <- map_dfr(ck_levene_list, bind_rows, .id = "PC")
levene_posthoc_df <- map_dfr(ck_levene_posthoc_list, bind_rows, .id = "PC")
```

df(3, 23) 
```{r}
levene_df %>% filter(term == "group") %>% dplyr::select(-c("term", "df"))
```

```{r}
levene_posthoc_df
```


Plot deviations of PC5
```{r}
df_dev <- res_A$meta %>%
    group_by(inj_trt_group) %>%
    mutate(dev = abs(.data[["PC5"]] - mean(.data[["PC5"]], na.rm = TRUE))) %>%
    ungroup()
df_dev

hist(df_dev$dev)
```
### try lm permutation p-values instead
```{r}
library(lmPerm)

ck_perm_list <- map(ck_pcs, ~ {
  
  df_dev <- res_A$meta %>%
    group_by(Injury, Treatment) %>%
    mutate(dev = abs(.data[[.x]] - median(.data[[.x]], na.rm = TRUE))) %>%
    ungroup()
  
  aovp(dev ~ Injury * Treatment, data = df_dev) %>%
    broom::tidy()
})

ck_perm_list
```

## Do fligner test instead - this worked
```{r}
ck_flig_list <- map(ck_pcs, ~ {
  f <- reformulate("inj_trt_group", response = .x)
  fligner.test(f, data = res_A$meta) %>%
    broom::tidy()
})

names(ck_flig_list) <- ck_pcs

ck_flig_df <- map_dfr(ck_flig_list, bind_rows, .id = "PC")

ck_flig_df
```

```{r}
levene_wilcox_posthoc <- function(y_var, data) {
  
  df_dev <- data %>%
    group_by(inj_trt_group) %>%
    mutate(dev = abs(.data[[y_var]] - median(.data[[y_var]], na.rm = TRUE))) %>%
    ungroup()
  
  f <- dev ~ inj_trt_group
  
  tests <- list(
    `Sham-Sham;Blast-Sham` = wilcox.test(
      f, data = subset(df_dev, inj_trt_group %in% c("Sham + Sham","Blast + Sham"))
    ),
    `Sham-VNS;Blast-VNS` = wilcox.test(
      f, data = subset(df_dev, inj_trt_group %in% c("Sham + VNS","Blast + VNS"))
    ),
    `Blast-Sham;Blast-VNS` = wilcox.test(
      f, data = subset(df_dev, inj_trt_group %in% c("Blast + Sham","Blast + VNS"))
    )
  )
  
  # raw_p <- 
  # sapply(tests, \(x) x$p.value)
  # p.adjust(raw_p, method = "bonferroni")
  tests
}
```


“Because dispersion estimates were non-normally distributed and sample sizes were small, we used the Fligner–Killeen test as a robust global test for heterogeneity of variance. Factor-specific effects were explored using two-way ANOVA on absolute deviations (Brown–Forsythe approach), and pairwise differences were assessed using Wilcoxon rank-sum tests.”

```{r}
ck_fligner_posthoc_list <- map(ck_pcs, ~ {
  levene_wilcox_posthoc(.x, res_A$meta)
})
names(ck_fligner_posthoc_list) <- ck_pcs

cfp_df <- map(ck_fligner_posthoc_list, ~ {
  sapply(.x, function(y){y$p.value})
}) %>% bind_rows(.id = "PC")

cfp_df
```


Two-way ANOVA on absolute deviations (Levene/Brown–Forsythe approach)
```{r}
ck_levene2_list <- map(ck_pcs, ~ {
  
  df_dev <- res_A$meta %>%
    group_by(Injury, Treatment) %>%
    mutate(dev = abs(.data[[.x]] - median(.data[[.x]], na.rm = TRUE))) %>%
    ungroup()
  
  f <- dev ~ Injury * Treatment
  
  aov(f, data = df_dev) %>%
    broom::tidy() %>%
    mutate(
      across(c(sumsq, meansq, statistic), ~ round(.x, 3)),
      p.value = round(p.value, 8)
    )
})

names(ck_levene2_list) <- ck_pcs

levene2_df <- map_dfr(ck_levene2_list, bind_rows, .id = "PC")

levene2_df
```

```{r}
res_A$meta %>% group_by(Injury) %>%
  summarise(MAD = mean(abs(PC5 - median(PC5))), .groups = "drop")
```
Mean absolute deviation (MAD) is *doubled* between Sham and Blast

```{r}
f <- as.formula("PC5 ~ Injury")
fligner.test(f, data = res_A$meta)

df_dev <- res_A$meta %>%
    group_by(Injury) %>%
    mutate(dev = abs(.data[["PC5"]] - mean(.data[["PC5"]], na.rm = TRUE))) %>%
    ungroup()
  
f <- as.formula("dev ~ Injury")
wilcox.test(f, data = df_dev)
```


```{r}
obs_diff <- df_dev %>%
  group_by(Injury) %>%
  summarise(MAD = mean(abs(PC5 - median(PC5))), .groups = "drop") %>%
  summarise(diff = MAD[Injury=="Blast"] - MAD[Injury=="Sham"]) %>%
  pull(diff)

perm_diff <- replicate(10000, {   # probably ought to stratify here
  shuffled <- sample(res_A$meta$PC5)
  df_perm <- res_A$meta %>% mutate(PC5 = shuffled)
  MADs <- df_perm %>% group_by(Injury) %>%
    summarise(MAD = mean(abs(PC5 - median(PC5))), .groups = "drop")
  MADs$MAD[MADs$Injury=="Blast"] - MADs$MAD[MADs$Injury=="Sham"]
})

p_value <- mean(abs(perm_diff) >= abs(obs_diff))
p_value
```

# Test for homogeneity of dispersion
```{r}
df_summary <- res_A$meta %>%
  group_by(Injury, Treatment) %>%
  summarise(MAD = mean(abs(PC5 - median(PC5))), .groups = "drop")

df_summary
```


```{r}
perm_2way_MAD <- function(df, var, factor1 = "Injury", factor2 = "Treatment", cell = "inj_trt_group",
                          nperm = 10000, seed = 123){
  set.seed(seed)
  
  # compute observed statistics
  df_summary <- df %>%
    group_by(.data[[factor1]], .data[[factor2]]) %>%
    summarise(MAD = mean(abs(.data[[var]] - median(.data[[var]]))), .groups = "drop")
  
  obs <- list(
    main1 = df_summary %>% group_by(.data[[factor1]]) %>%
      summarise(MAD_mean = mean(MAD)) %>%
      summarise(diff = MAD_mean[1] - MAD_mean[2]) %>% pull(diff),
    main2 = df_summary %>% group_by(.data[[factor2]]) %>%
      summarise(MAD_mean = mean(MAD)) %>%
      summarise(diff = MAD_mean[1] - MAD_mean[2]) %>% pull(diff),
    interaction = ((df_summary$MAD[1] - df_summary$MAD[2]) - (df_summary$MAD[3] - df_summary$MAD[4]))
  )
  
  # preallocate vectors
  perm_main1 <- perm_main2 <- perm_inter <- numeric(nperm)
  
  for(i in seq_len(nperm)){
    df_perm <- df
    
    # shuffle within cells for interaction test
    df_perm <- df_perm %>%
      group_by(.data[[cell]]) %>%
      mutate(perm_var = sample(.data[[var]])) %>%
      ungroup()
    
    df_summ <- df_perm %>%
      group_by(.data[[factor1]], .data[[factor2]]) %>%
      summarise(MAD = mean(abs(.data[["perm_var"]] - median(.data[["perm_var"]]))), .groups = "drop")
    
    # main effect 1: shuffle factor1 within factor2 levels
    df_main1 <- df_perm %>%
      group_by(.data[[factor2]]) %>%
      mutate(tmp_factor1 = sample(.data[[factor1]])) %>%
      ungroup() %>%
      group_by(tmp_factor1, .data[[factor2]]) %>%
      summarise(MAD = mean(abs(.data[["perm_var"]] - median(.data[["perm_var"]]))), .groups = "drop")
    
    perm_main1[i] <- df_main1 %>%
      group_by(tmp_factor1) %>%
      summarise(MAD_mean = mean(MAD)) %>%
      summarise(diff = MAD_mean[1] - MAD_mean[2]) %>% pull(diff)
    
    # main effect 2: shuffle factor2 within factor1 levels
    df_main2 <- df_perm %>%
      group_by(.data[[factor1]]) %>%
      mutate(tmp_factor2 = sample(.data[[factor2]])) %>%
      ungroup() %>%
      group_by(.data[[factor1]], tmp_factor2) %>%
      summarise(MAD = mean(abs(.data[["perm_var"]] - median(.data[["perm_var"]]))), .groups = "drop")
    
    perm_main2[i] <- df_main2 %>%
      group_by(tmp_factor2) %>%
      summarise(MAD_mean = mean(MAD)) %>%
      summarise(diff = MAD_mean[1] - MAD_mean[2]) %>% pull(diff)
    
    # interaction: already precomputed from permuted cells
    perm_inter[i] <- ((df_summ$MAD[1] - df_summ$MAD[2]) - (df_summ$MAD[3] - df_summ$MAD[4]))
  }
  
  p_values <- c(
    main1 = mean(abs(perm_main1) >= abs(obs$main1)),
    main2 = mean(abs(perm_main2) >= abs(obs$main2)),
    interaction = mean(abs(perm_inter) >= abs(obs$interaction))
  )
  
  list(observed = obs, p_values = p_values)
}
```

```{r}
results <- lapply(ck_pcs, function(v){
  perm_2way_MAD(res_A$meta, var = v, factor1 = "Injury", factor2 = "Treatment", cell = "inj_trt_group", nperm = 999)
})
names(results) <- ck_pcs
results$PC5
```
```{r}
# ----- Function to run stratified permutation 2-way ANOVA on MADs -----
perm_2way_MAD <- function(df, var, factor1 = "Injury", factor2 = "Treatment",
                          cell = "inj_trt_group", nperm = 5000, seed = 123){
  set.seed(seed)
  
  # Compute observed MADs
  df_summary <- df %>%
    group_by(.data[[factor1]], .data[[factor2]]) %>%
    summarise(MAD = mean(abs(.data[[var]] - median(.data[[var]]))), .groups = "drop")
  
  obs <- list(
    main1 = df_summary %>% group_by(.data[[factor1]]) %>%
      summarise(MAD_mean = mean(MAD)) %>%
      summarise(diff = MAD_mean[1] - MAD_mean[2]) %>% pull(diff),
    main2 = df_summary %>% group_by(.data[[factor2]]) %>%
      summarise(MAD_mean = mean(MAD)) %>%
      summarise(diff = MAD_mean[1] - MAD_mean[2]) %>% pull(diff),
    interaction = ((df_summary$MAD[1] - df_summary$MAD[2]) - (df_summary$MAD[3] - df_summary$MAD[4]))
  )
  
  # Preallocate permutation vectors
  perm_main1 <- perm_main2 <- perm_inter <- numeric(nperm)
  
  for(i in seq_len(nperm)){
    df_perm <- df
    
    # Interaction: shuffle within each 2x2 cell
    df_perm <- df_perm %>%
      group_by(.data[[cell]]) %>%
      mutate(perm_var = sample(.data[[var]])) %>%
      ungroup()
    
    # Compute MADs per cell
    df_summ <- df_perm %>%
      group_by(.data[[factor1]], .data[[factor2]]) %>%
      summarise(MAD = mean(abs(.data[["perm_var"]] - median(.data[["perm_var"]]))), .groups = "drop")
    
    # Main effect 1: shuffle factor1 within factor2
    df_main1 <- df_perm %>%
      group_by(.data[[factor2]]) %>%
      mutate(tmp_factor1 = sample(.data[[factor1]])) %>%
      ungroup() %>%
      group_by(tmp_factor1, .data[[factor2]]) %>%
      summarise(MAD = mean(abs(.data[["perm_var"]] - median(.data[["perm_var"]]))), .groups = "drop")
    
    perm_main1[i] <- df_main1 %>%
      group_by(tmp_factor1) %>%
      summarise(MAD_mean = mean(MAD)) %>%
      summarise(diff = MAD_mean[1] - MAD_mean[2]) %>% pull(diff)
    
    # Main effect 2: shuffle factor2 within factor1
    df_main2 <- df_perm %>%
      group_by(.data[[factor1]]) %>%
      mutate(tmp_factor2 = sample(.data[[factor2]])) %>%
      ungroup() %>%
      group_by(.data[[factor1]], tmp_factor2) %>%
      summarise(MAD = mean(abs(.data[["perm_var"]] - median(.data[["perm_var"]]))), .groups = "drop")
    
    perm_main2[i] <- df_main2 %>%
      group_by(tmp_factor2) %>%
      summarise(MAD_mean = mean(MAD)) %>%
      summarise(diff = MAD_mean[1] - MAD_mean[2]) %>% pull(diff)
    
    # Interaction effect from permuted cells
    perm_inter[i] <- ((df_summ$MAD[1] - df_summ$MAD[2]) - (df_summ$MAD[3] - df_summ$MAD[4]))
  }
  
  # Compute permutation p-values
  p_values <- c(
    main1 = mean(abs(perm_main1) >= abs(obs$main1)),
    main2 = mean(abs(perm_main2) >= abs(obs$main2)),
    interaction = mean(abs(perm_inter) >= abs(obs$interaction))
  )
  
  # Return list
  list(observed = obs, p_values = p_values)
}


# ----- Function to run across multiple variables and return tidy dataframe -----
perm_2way_MAD_multi <- function(df, vars, factor1 = "Injury", factor2 = "Treatment",
                                cell = "inj_trt_group", nperm = 5000, seed = 123){
  
  results <- map(vars, ~perm_2way_MAD(df, var = .x, factor1 = factor1, factor2 = factor2,
                                      cell = cell, nperm = nperm, seed = seed))
  names(results) <- vars
  
  # Convert to tidy dataframe
  map_dfr(names(results), function(v){
    res <- results[[v]]
    tibble(
      variable = v,
      effect = names(res$p_values),
      observed_diff = unlist(res$observed)[names(res$p_values)],
      p_value = unlist(res$p_values)
    )
  })
}

# ----- Example usage -----
results_df <- perm_2way_MAD_multi(res_A$meta, ck_pcs, nperm = 5000)

# Optional: pivot wider for reporting
results_wide <- results_df %>%
  pivot_wider(names_from = effect, values_from = c(observed_diff, p_value))

results_wide
```

```{r}
saveRDS(results_df, file = "1_preprocessed/cyto_PCA_dispersion.rds")
```

```{r}
results_wide
```
```{r}
# ----- Function to run stratified permutation 2-way ANOVA on variance -----
perm_2way_VAR <- function(df, var, factor1 = "Injury", factor2 = "Treatment",
                         cell = "inj_trt_group", nperm = 5000, seed = 123){
  set.seed(seed)
  
  # ensure consistent ordering
  df <- df %>%
    mutate(
      !!factor1 := factor(.data[[factor1]]),
      !!factor2 := factor(.data[[factor2]])
    )
  
  f1_levels <- levels(df[[factor1]])
  f2_levels <- levels(df[[factor2]])
  
  # ----- Observed statistics -----
  
  # Main effects (marginal — NOT per cell)
  obs_main1 <- df %>%
    group_by(.data[[factor1]]) %>%
    summarise(VAR = var(.data[[var]]), .groups = "drop") %>%   # or do I want var = mean((x-mean(x))^2)
    summarise(diff = VAR[.data[[factor1]] == f1_levels[2]] -
                     VAR[.data[[factor1]] == f1_levels[1]]) %>%
    pull(diff)
  
  obs_main2 <- df %>%
    group_by(.data[[factor2]]) %>%
    summarise(VAR = var(.data[[var]]), .groups = "drop") %>%    # or do I want var = mean((x-mean(x))^2)
    summarise(diff = VAR[.data[[factor2]] == f2_levels[2]] -
                     VAR[.data[[factor2]] == f2_levels[1]]) %>%
    pull(diff)
  
  # Interaction (cell-level — this part stays per-cell)
  df_cells <- df %>%
    group_by(.data[[factor1]], .data[[factor2]]) %>%
    summarise(VAR = var(.data[[var]]), .groups = "drop")   # or do I want var = mean((x-mean(x))^2)
  
  obs_inter <- with(df_cells,
  (VAR[df_cells[[factor1]] == f1_levels[2] & df_cells[[factor2]] == f2_levels[2]] -
   VAR[df_cells[[factor1]] == f1_levels[2] & df_cells[[factor2]] == f2_levels[1]]) -
  (VAR[df_cells[[factor1]] == f1_levels[1] & df_cells[[factor2]] == f2_levels[2]] -
   VAR[df_cells[[factor1]] == f1_levels[1] & df_cells[[factor2]] == f2_levels[1]])
)
  
  # ----- Permutations -----
  perm_main1 <- perm_main2 <- perm_inter <- numeric(nperm)
  
  for(i in seq_len(nperm)){
    
    # --- MAIN EFFECT 1: shuffle factor1 within factor2 ---
    df_m1 <- df %>%
      group_by(.data[[factor2]]) %>%
      mutate(tmp_f1 = sample(.data[[factor1]])) %>%
      ungroup()
    
    perm_main1[i] <- df_m1 %>%
      group_by(tmp_f1) %>%
      summarise(VAR = var(.data[[var]]), .groups = "drop") %>%
      summarise(diff = VAR[tmp_f1 == f1_levels[2]] -
                       VAR[tmp_f1 == f1_levels[1]]) %>%
      pull(diff)
    
    # --- MAIN EFFECT 2: shuffle factor2 within factor1 ---
    df_m2 <- df %>%
      group_by(.data[[factor1]]) %>%
      mutate(tmp_f2 = sample(.data[[factor2]])) %>%
      ungroup()
    
    perm_main2[i] <- df_m2 %>%
      group_by(tmp_f2) %>%
      summarise(VAR = var(.data[[var]]), .groups = "drop") %>%
      summarise(diff = VAR[tmp_f2 == f2_levels[2]] -
                       VAR[tmp_f2 == f2_levels[1]]) %>%
      pull(diff)
    
    # --- INTERACTION: shuffle values within cells ---
    df_int <- df %>%
      group_by(.data[[cell]]) %>%
      mutate(perm_var = sample(.data[[var]])) %>%
      ungroup()
    
    df_cells_perm <- df_int %>%
      group_by(.data[[factor1]], .data[[factor2]]) %>%
      summarise(VAR = var(.data[["perm_var"]]), .groups = "drop")
    
    perm_inter[i] <- with(df_cells_perm,
      (VAR[df_cells_perm[[factor1]] == f1_levels[2] & df_cells_perm[[factor2]] == f2_levels[2]] -
       VAR[df_cells_perm[[factor1]] == f1_levels[2] & df_cells_perm[[factor2]] == f2_levels[1]]) -
      (VAR[df_cells_perm[[factor1]] == f1_levels[1] & df_cells_perm[[factor2]] == f2_levels[2]] -
       VAR[df_cells_perm[[factor1]] == f1_levels[1] & df_cells_perm[[factor2]] == f2_levels[1]])
    )
  }
  
  # ----- P-values -----
  p_values <- c(
    main1 = mean(abs(perm_main1) >= abs(obs_main1)),
    main2 = mean(abs(perm_main2) >= abs(obs_main2)),
    interaction = mean(abs(perm_inter) >= abs(obs_inter))
  )
  
  list(
    observed = c(main1 = obs_main1, main2 = obs_main2, interaction = obs_inter),
    p_values = p_values
  )
}


# ----- Multi-variable wrapper -----
perm_2way_VAR_multi <- function(df, vars, factor1 = "Injury", factor2 = "Treatment",
                               cell = "inj_trt_group", nperm = 5000, seed = 123){
  
  results <- map(vars, ~perm_2way_VAR(df, var = .x,
                                      factor1 = factor1,
                                      factor2 = factor2,
                                      cell = cell,
                                      nperm = nperm,
                                      seed = seed))
  names(results) <- vars
  
  map_dfr(names(results), function(v){
    res <- results[[v]]
    tibble(
      variable = v,
      effect = names(res$p_values),
      observed_diff = res$observed[names(res$p_values)],
      p_value = res$p_values
    )
  })
}


# ----- Run -----
results_df <- perm_2way_VAR_multi(res_A$meta, ck_pcs, nperm = 5000)
saveRDS(results_df, file = "1_preprocessed/cyto_PCA_variance.rds")

results_wide <- results_df %>%
  pivot_wider(names_from = effect, values_from = c(observed_diff, p_value))

results_wide
```

# Save
Now save as 2 sheets in the same excel workbook
```{r}
# save df as 2 separate sheets
wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
writeData(wb, "ANOVA", anova_df)
addWorksheet(wb, "posthoc")
writeData(wb, "posthoc", posthoc_df)
saveWorkbook(wb, file = "final/cytokine_PC_stats.xlsx", overwrite = TRUE)
```


### Get loading plots
```{r}
breaks <- c(-0.4, -0.2, 0.0, 0.2, 0.4)
# xlim   <- c(-0.45, 0.45)
weight <- map(ck_pcs, ~ {
  load_df <- data.frame(
    Cytokine = res_A$loadings$Feature,
    Contribution = res_A$loadings[[.x]]
  )
  # xlim <- max(abs(load_df$Contribution)) * 1.2
  plt <- ggplot(load_df, aes(x = reorder(Cytokine, Contribution), y = Contribution)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    theme_minimal() +
    scale_y_continuous(               # x-axis scaling/locations
      # limits = c(xlim),
      breaks = breaks
    ) +
    labs(
      title = paste0(.x, " (", round(var_explained$data[var_explained$data$PC == .x, ]$prop_var * 100, 3), " %)"),
      x = "Cytokine", y = "Loading Value"
      )
  list(plt)
})
names(weight) <- ck_pcs

weight[[1]]
weight[[2]]
weight[[3]]
weight[[4]]
weight[[5]]

# groupwise is already a named list of plots
imap(weight, function(plot_list, nm) {
  ggsave(
    filename = paste0("../figures/cytokine_loading_", nm, ".png"),
    plot     = plot_list[[1]],   # since each element is a list with one plot
    width    = 4, height = 5.5
  )
})

imap(weight, function(plot_list, nm) {
  ggsave(
    filename = paste0("../figures/cytokine_loading_", nm, ".svg"),
    plot     = plot_list[[1]],   # since each element is a list with one plot
    width    = 4, height = 5.5,
    device   = "svg"
  )
})
```

### get group-wise boxplots
```{r}
groupwise <- map(ck_pcs, ~ {
  plt <- box_easy(
    data  = res_A$meta,
    dv    = .x,
    y_lab = paste0(.x, " (", round(var_explained$data[var_explained$data$PC == .x, ]$prop_var * 100, 1), " %)"),
    )
  list(plt)
})
names(groupwise) <- ck_pcs

groupwise[[1]]
groupwise[[2]]
groupwise[[3]]
groupwise[[4]]
groupwise[[5]]

# groupwise is already a named list of plots
imap(groupwise, function(plot_list, nm) {
  ggsave(
    filename = paste0("../figures/cytokine_barplot_", nm, ".png"),
    plot     = plot_list[[1]],   # since each element is a list with one plot
    width    = 5.5, height = 5.5
  )
})

imap(groupwise, function(plot_list, nm) {
  ggsave(
    filename = paste0("../figures/cytokine_barplot_", nm, ".svg"),
    plot     = plot_list[[1]],   # since each element is a list with one plot
    width    = 5.5, height = 5.5,
    device   = "svg"
  )
})
```


### Make a biplot with marginal boxplots
```{r}
# 1. Main biplot WITH legend
p_main <- plot_pca_biplot(res_A, "inj_trt_group", four_colors, i_vectors = 0)

# 2. Extract x and y axes of p_main
build_main <- ggplot_build(p_main)
x_limits <- build_main$layout$panel_scales_x[[1]]$range$range
y_limits <- build_main$layout$panel_scales_y[[1]]$range$range

# 3. Extract its legend
shared_legend <- get_legend(p_main)

# 4. Remove legends from all panels
p_main <- p_main + theme(legend.position = "none")
p_x   <- marginal_boxplot(res_A, PC = "PC1", vertical = TRUE,
                                group_var = "inj_trt_group",
                                group_colors = four_colors) +
               theme(legend.position = "none") +
               scale_y_continuous(limits = x_limits)

p_y   <- marginal_boxplot(res_A, PC = "PC2", vertical = FALSE,
                                group_var = "inj_trt_group",
                                group_colors = four_colors) +
               theme(legend.position = "none") +
               scale_y_continuous(limits = y_limits)

# Align p_main with each marginal, perfectly
### align p_x & p_main on the vertical edge
aligned_p <- cowplot::align_plots(p_x, p_main, align = "v", axis = "l")
p_x_aligned   <- aligned_p[[1]]
p_main_v <- aligned_p[[2]]

### align p_main & p_y on the horizontal edge
aligned_h <- cowplot::align_plots(p_main, p_y, align = "h", axis = "b")
p_main_h  <- aligned_h[[1]]
p_y_aligned    <- aligned_h[[2]]

# First row: p_x_noleg and blank
top_row <- plot_grid(p_x_aligned, shared_legend, ncol = 2, rel_widths  = c(4, 1), align = "hv")
# Second row: p_main and p_y_noleg
bottom_row <- plot_grid(p_main, p_y_aligned, ncol   = 2, rel_widths = c(4, 1))
# Assemble final plot
final_plot <- plot_grid(top_row, bottom_row, ncol  = 1, rel_heights = c(1, 4))            # main 4/5, legend 1/5

# Display
print(final_plot)

# Not going to use this in the paper, but I thought it was cool.
# ggsave("../figures/A_cyto_PCA_biplot_marginal.png", final_plot, width = 6, height = 6)
# ggsave("../figures/A_cytoPCA_biplot_marginal.svg", final_plot, device = "svg", height = 6, width = 6)

```