---
title: "B_plot_final.Rmd"
author: "Alyssa N. Easton"
date: '2025-02-17'
output:
---

Plot results from B_joint_final, save in PDF format

## Setup
```{r "setup", include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
knitr::opts_knit$set(root.dir = file.path(Sys.getenv("VNS_ROOT"), "data"))  # stored VNS_ROOT in ~/.Renviron
options(expressions = 6000)  # Increase allowed recursion depth
options(digits = 4) ## Formats output to 10 digits
library(magrittr)
library(tidyverse)
library(broom)
library(broom.helpers)
library(rlang)
library(glue)
library(janitor)
library(purrr)
library(ggh4x)
library(parallel)
library(parallelly)
library(pbapply)
library(furrr)      # future_pmap comes from here
library(future)
library(limma)
library(ggeffects)
library(patchwork)
library(car)
library(scales)
library(future.apply)
library(ComplexHeatmap)
library(circlize)
library(cowplot)
library(Cairo)
library(grid)
library(gridExtra)
library(ggtext)
library(lme4)
library(lmerTest)
# source("/proj/gibbons/2024_easton_vns/workflow/functions/mdact.R")
# source("/proj/gibbons/2024_easton_vns/workflow/functions/june_functions.R")
# source("/proj/gibbons/2024_easton_vns/workflow/functions/shannon_div_functions.R")
# source("/proj/gibbons/2024_easton_vns/workflow/functions/artifact_correct_functions.R")
# source("/proj/gibbons/2024_easton_vns/workflow/functions/lmer_functions.R")
source("/proj/gibbons/2024_easton_vns/workflow/functions/final_form_functions.R")
```


Plotting functions moved to `final_form_functions.R`

# 2: Analyzing DA and M-DACT results

Since I already ran prior analyses, I can just import these dfs
```{r}
df            <- readRDS("../data/processed/B_clr_df_withalpha_0216.rds")
model_microbe <- readRDS("../data/3_newest/B_microbe_models_0216.rds")
model_outcome <- readRDS("../data/3_newest/B_outcome_models_0216.rds")
all_p_df      <- readRDS("../data/3_newest/B_LRT_p_values_0216.rds")
all_run_df    <- readRDS("../data/3_newest/B_mediation_0216.rds")

coef_df <- readRDS("../data/3_newest/B_microbe_coef_dfs_0217.rds")
m_coef_df <- coef_df$m
y_coef_df <- coef_df$y

ci_df <- readRDS("/proj/gibbons/2024_easton_vns/data/3_newest/B_bootCI_df_0216.rds")

m_lrt <- model_microbe %>%
  transmute(microbe, term, "p" = p.value, "q" = adj.p) %>%
  filter(term %in% c("blast", "vns")) %>%
  pivot_wider(
    id_cols = "microbe", names_from = "term",
    values_from = c("p", "q"),
    names_glue = "{.value}_{term}"
    ) %>%
  mutate(
    blast_signif = ifelse(q_blast < 0.2, "Blast", ""),
    vns_signif = ifelse(q_vns < 0.2, "VNS", ""),
    # microbe = gsub("_", " ", microbe)
  )

da_hits <- m_lrt %>% filter(blast_signif == "Blast" | vns_signif == "VNS")
da_bugs <- da_hits %>% pull(microbe) %>% unique()

med_hits <- all_run_df %>%
  filter(!is.na(signif_at) & signif_at <= 0.2) %>%
  rowwise() %>% mutate(bug_out = paste0(mediator, ";", outcome)) %>%
  dplyr::select(-any_of(starts_with("dact_thr")))

outcome_hits <- map_dfr(model_outcome, function(df) {
  df %>% filter(term == "microbe", adj.p < 0.2) }) %>%
  rowwise() %>% mutate(bug_out = paste0(microbe, ";", outcome)) %>%
  arrange(adj.p) %>%
  mutate(hit_type = factor(ifelse(bug_out %in% med_hits$bug_out, "mediator", "non-mediator")))
```

Set up scales early for microbe abundance + outcome
```{r}
########################  heatmap values of microbes  ########################  
all_plotted_bugs <- c(
  da_bugs, med_hits$mediator, outcome_hits$microbe
) %>% unique()

max_bug_b <- m_coef_df %>%
  filter(microbe %in% all_plotted_bugs) %>% 
  dplyr::select(contains("est_")) %>%
  max(abs(.))

max_bug_value <- max(c(2.418, max_bug_b)) # 2.418 was max in cohort A (larger than here)

bug_color <- colorRamp2(  # formerly col_fun
  seq(-max_bug_value, max_bug_value, length.out = 11),
  rev(RColorBrewer::brewer.pal(11, "RdBu"))
)


########################  heatmap values of outcomes  ########################  
all_plotted_out <- unique(c(outcome_hits$bug_out, med_hits$bug_out))

max_out_b <- y_coef_df %>%
  mutate(bug_out = paste0(microbe, ";", outcome)) %>%
  filter(bug_out %in% all_plotted_out) %>%
  dplyr::select(contains("est_")) %>%
  max(abs(.))

# max_out_value <- max(c(4.043, max_bug_b)) # 4.043 was max in cohort A
max_out_value <- max_out_b

out_color <- colorRamp2(  # formerly col_fun
  seq(-max_out_value, max_out_value, length.out = 11),
  rev(RColorBrewer::brewer.pal(11, "PuOr"))            # changed this so behavioral outcomes are a different color from cytokines
)

########################  FDR colors  ########################  
# fdr_fun <- colorRamp2(c(0.00, 0.30), c("darkgreen", "lightyellow"))
fdr_value_vector <- c(# rev(c(
  "0.100" = 0.1,
  "0.125" = 0.125,
  "0.150" = 0.15,
  "0.175" = 0.175,
  "0.200" = 0.2) #)

# fdr_colors <- structure(
#   RColorBrewer::brewer.pal(length(fdr_value_vector), "BuPu"),
#   names = fdr_value_vector
#   )

fdr_color <- colorRamp2(  # formerly col_fun
  seq(min(fdr_value_vector), max(fdr_value_vector), length.out = 5),
  # RColorBrewer::brewer.pal(5, "BuPu")
  rev(RColorBrewer::brewer.pal(5, "BuPu"))
)

```

Which outcome names do I need to edit?
plotted_outcomes <- all_plotted_out %>% gsub("^.*?;", "", .) %>% unique()

```{r}
max_bug_value
max_out_value
```


Build legends
```{r}
bug_lgd <- Legend(
  col_fun = bug_color,
  title = "Effect on \nMicrobe",
  at = c(-2.5, 0, 2.5)
  )

out_lgd <- Legend(col_fun = out_color, title = "Effect on \nOutcome", at = c(-2, -1, 0, 1, 2))
fdr_lgd <- Legend(col_fun = fdr_color, title = "Mediation \nFDR <",   at = fdr_value_vector,
                  labels = names(fdr_value_vector))

blast_lgd <- Legend(
  labels = c("Blast", "VNS"),
  legend_gp = gpar(fill = c("#0492c2", "#FFc067")),
  title = "LRT FDR < 0.2"
  )

pd <- packLegend(bug_lgd, blast_lgd, out_lgd, fdr_lgd, direction = "vertical")
```

```{r, fig.width = 2, fig.height = 5}
# create a viewport
grid.newpage()
pushViewport(viewport(width = 1, height = 1))

# Draw the whole stack at once, aligned to the left
draw(pd, x = unit(0.5, "cm"), y = unit(0.5, "npc"), just = "left")
```
Save it
```{r}
pdf("../figures/B_heatmap_legends.pdf", width = 2, height = 5)
draw(pd, x = unit(0.5, "cm"), y = unit(0.5, "npc"), just = "left")
dev.off()
```

Taxonomy 
```{r}
med_bugs <- med_hits %>% pull(mediator) %>% unique()

taxmap %>% filter(Species %in% da_bugs)
taxmap %>% filter(Species %in% med_bugs)
```


## Plotting
### Exposure -> microbe
```{r}
m_coef_df %>% filter(microbe %in% da_bugs)

da_blast <- setNames(da_hits$blast_signif, nm = da_hits$microbe)
da_vns   <- setNames(da_hits$vns_signif,   nm = da_hits$microbe)
```

### Microbe Heat map
```{r, fig.width = 5, fig.height = 4.5}
m_coef_mat <- m_coef_df %>%
  filter(microbe %in% da_bugs) %>%
  # mutate(microbe = gsub("_", " ", microbe)) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(contains("est")) %>%
  as.matrix()

colnames(m_coef_mat) <- c("Blast", "VNS", "Blast:VNS")

m_p_mat <- m_coef_df %>%
  filter(microbe %in% da_bugs) %>%
  # mutate(microbe = gsub("_", " ", microbe)) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(contains("p")) %>%
  mutate(
    p_blast = case_when(p_blast < 0.001 ~ "**", p_blast < 0.05 ~ "*", .default = ""),
    p_vns   = case_when(p_vns   < 0.001 ~ "**", p_vns   < 0.05 ~ "*", .default = ""),
    p_bv    = case_when(p_bv    < 0.001 ~ "**", p_bv    < 0.05 ~ "*", .default = "")
    ) %>%
  as.matrix()

da_blast <- da_blast[rownames(m_coef_mat)] # %>% as.factor()
da_vns   <- da_vns[rownames(m_coef_mat)] # %>% as.factor()

blast_color <- c("Blast" = "#0492c2")
vns_color   <- c("VNS"   = "#FFc067") # NA values show up as transparent anyways
```


Where da_vns is a named array denoting whether a microbe was significant for VNS (TRUE/FALSE), and da_blast similarly for Blast.

```{r}
lrt_ann_blast<- rowAnnotation(
  "Blast" = da_blast,
  col = list("Blast" = blast_color),
  show_annotation_name = FALSE,
  show_legend = FALSE,
  simple_anno_size = unit(0.2, "cm"),
  na_col = "#FFFFFF00",
  annotation_legend_param = list(title = NULL)
)

lrt_ann_vns <- rowAnnotation(
  "VNS" = da_vns,
  col = list("VNS" = vns_color),
  show_annotation_name = FALSE,
  show_legend = FALSE,
  simple_anno_size = unit(0.2, "cm"),
  na_col = "#FFFFFF00",
  annotation_legend_param = list(title = NULL)
)

# plot(lrt_ann_blast)
```


```{r, fig.width=6, fig.height=3.5}
wide <- 2 * ncol(m_coef_mat)
high <- 2 * nrow(m_coef_mat)

ht <- Heatmap(
  m_coef_mat,
  name = "Estimate",
  col = bug_color,
  cluster_columns = FALSE,
  show_row_dend = FALSE,
  show_column_dend = FALSE,
  left_annotation = lrt_ann_blast, # since lrt_ann_vns has no hits, remove
  cluster_rows = TRUE,
  row_labels = clean_bug_names(rownames(m_coef_mat)),
  column_names_rot = 45,
  # row_split = da_vns,   # no VNS hits, so we don't need this
  # cluster_row_slices = TRUE,
  
  show_heatmap_legend = FALSE,
  
  width  = unit(wide, "cm"),
  height = unit(high, "cm"),
  row_names_gp = grid::gpar(fontsize = 20, fontface = "italic"),
  column_names_gp = grid::gpar(fontsize = 20),
  
  cell_fun = function(j, i, x, y, w, h, fill) {
    grid::grid.text(
      m_p_mat[i, j],
      x = x,
      y = y,
      gp = grid::gpar(
        fontsize = 24,
        col = "black",
        fontface = "bold"
      )
    )
  })

ht
```

```{r}
pdf("../figures/B_microbe_abundance_fdr20.pdf", width = 6, height = 3.5)
ht_drawn <- draw(ht)
# row_ord <- row_order(ht_drawn)
# da_bug_order <- rownames(trt_effect_matrix)[row_ord]
dev.off()
```


Romboutsia is generally increased by treatment, but especially blast
Clostridiales is generally increased by blast, but especially on VNS background
SGB41718 is generally decreased by blast, but especially on VNS background

Can also plot/test "presence/absence" if reviewers request, but the results did align well when I tested them for sparse features (as in, sparse bugs that were significant in the CLR regressions were actually even more significant if I did a presence/absence logistic regression, because CLR-transforming zeroes is just adding noise).

## Bulk plots

one_residual
```{r, attr.message=FALSE}
pdf_split(
    hit_df        = outcome_hits,
    data_df       = df,
    sub_height    = 5,
    sub_width     = 5.5,
    n_cols        = 4,
    plot_function = "residout_bug",
    filename      = "../figures/B_outcome_resid_plots.pdf"
    )
```

mediation hits, one residual
```{r, attr.message=FALSE}
med_hits_renamed <- med_hits %>% rename(microbe = mediator)

pdf_split(
    hit_df        = med_hits_renamed,
    data_df       = df,
    sub_height    = 5,
    sub_width     = 5.5,
    n_cols        = 4,
    plot_function = "residout_bug",
    filename      = "../figures/B_medhits_scatter_rX_Y.pdf"
    )
```
mediation hits, two residual
```{r, echo=FALSE}
pdf_split(
    hit_df        = med_hits_renamed,
    data_df       = df,
    sub_height    = 5,
    sub_width     = 5.5,
    n_cols        = 4,
    plot_function = "residout_residbug",
    filename      = "../figures/B_medhits_scatter_rX_rY.pdf"
    )
```
Regular box plot
```{r, fig.width = 5, fig.height = 5}
bugs_to_plot <- unique(c(da_bugs, outcome_hits$microbe, med_hits$mediator)) %>% sort() # put in alphabetical order

# check this!!!
levels(df$inj_trt_group)

pdf_split(
    hit_df        = da_bugs,
    data_df       = df,
    plot_function = "box_easier",
    filename      = "../figures/B_dahits_box.pdf",
    sub_height = 5, sub_width = 5, n_cols = 4
    )

pdf_split(
    hit_df        = outcome_hits$microbe,
    data_df       = df,
    plot_function = "box_easier",
    filename      = "../figures/B_outhits_box.pdf",
    sub_height = 5, sub_width = 5, n_cols = 4
    )

pdf_split(
    hit_df        = med_hits$mediator,
    data_df       = df,
    plot_function = "box_easier",
    filename      = "../figures/B_medhits_box.pdf",
    sub_height = 5, sub_width = 5, n_cols = 4
    )

# pdf_split(
#     hit_df        = bugs_to_plot,
#     data_df       = df,
#     plot_function = "box_easier",
#     filename      = "../figures/B_microbe_abundance.pdf",
#     sub_height = 5, sub_width = 5, n_cols = 4
#     )
```

# Big mediaton plot
## Plot coefficients for mediators + annotate with p-values
```{r}
med_bugs <- unique(med_hits$mediator) # luckily they are all unique

med_bugs <- med_hits %>%
  filter(signif_at <= 0.2) %>%
  pull(mediator)

m_coef_mat <- m_coef_df %>%
  filter(microbe %in% med_bugs) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(contains("est")) %>%
  as.matrix()

m_p_mat <- m_coef_df %>%
  filter(microbe %in% med_bugs) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(contains("p")) %>%
  mutate(
    p_blast = case_when(p_blast < 0.001 ~ "**", p_blast < 0.05 ~ "*", .default = ""),
    p_vns   = case_when(p_vns   < 0.001 ~ "**", p_vns   < 0.05 ~ "*", .default = ""),
    p_bv    = case_when(p_bv    < 0.001 ~ "**", p_bv    < 0.05 ~ "*", .default = "")
    ) %>%
  as.matrix()

y_coef_mat <- y_coef_df %>%
  semi_join(med_hits, by = c("outcome", "microbe" = "mediator")) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(contains("est")) %>%
  as.matrix()

y_p_mat <- y_coef_df %>%
  semi_join(med_hits, by = c("outcome", "microbe" = "mediator")) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(contains("p")) %>%
  mutate(
    p_blast = case_when(p_blast < 0.001 ~ "**", p_blast < 0.05 ~ "*", .default = ""),
    p_vns   = case_when(p_vns   < 0.001 ~ "**", p_vns   < 0.05 ~ "*", .default = ""),
    p_bv    = case_when(p_bv    < 0.001 ~ "**", p_bv    < 0.05 ~ "*", .default = ""),
    p_microbe = case_when(p_microbe < 0.001 ~ "**", p_microbe    < 0.05 ~ "*", .default = "")
    ) %>%
  as.matrix()

row_order <- rownames(m_coef_mat)
y_coef_mat <- y_coef_mat[row_order, c(colnames(m_coef_mat), "est_microbe")]
y_p_mat <- y_p_mat[row_order, c(colnames(m_p_mat), "p_microbe")]

row_outcome <- med_hits %>%
  distinct(mediator, outcome) %>%
  filter(mediator %in% rownames(m_coef_mat)) %>%
  tibble::deframe()

row_exposure <- med_hits %>%
  distinct(mediator, exposure) %>%
  filter(mediator %in% rownames(m_coef_mat)) %>%
  tibble::deframe()

row_fdr <- med_hits$signif_at
names(row_fdr) <- med_hits$mediator

row_outcome <- row_outcome[rownames(m_coef_mat)]
row_exposure <- row_exposure[rownames(m_coef_mat)]
row_fdr <- row_fdr[rownames(m_coef_mat)]


stopifnot(all(names(row_outcome) == rownames(m_coef_mat)))
stopifnot(all(names(row_outcome) == rownames(y_coef_mat)))
stopifnot(all(names(row_fdr) == rownames(row_outcome)))

exposure_color <- c("pm_blast" = "#0492c2", "pm_vns" = "#FFc067")
```


```{r, fig.width = 1, fig.height = 1}
fdr_annotation <- rowAnnotation(
  "FDR <" = row_fdr,
  col = list("FDR <" = fdr_color),
  simple_anno_size = unit(0.2, "cm"),
  show_annotation_name = FALSE,
  show_legend = FALSE,
  annotation_legend_param = list(title = NULL)
)

which_exposure <- rowAnnotation(
  "Exposure" = row_exposure,
  col = list("Exposure" = exposure_color),
  simple_anno_size = unit(0.2, "cm"),
  show_annotation_name = FALSE,
  show_legend = FALSE,
  annotation_legend_param = list(title = NULL)
)

plot(fdr_annotation + which_exposure)
```


```{r, fig.width = 7, fig.height = 6}
wide <- 2 * ncol(m_coef_mat)
high <- 2 * nrow(m_coef_mat)
row_exposure_clean <- row_exposure %>% gsub("pm_blast", "Blast", .) %>% gsub("pm_vns", "VNS", .)

# heatmap model M coefficients
ht_m <- Heatmap(
  m_coef_mat,
  name = "M Estimate",
  col = bug_color,
  left_annotation = which_exposure,
  
  row_split = row_exposure_clean,
  row_title = NULL,
  
  cluster_rows = TRUE,
  cluster_row_slices = TRUE,
  show_row_dend = FALSE,
  row_labels = clean_bug_names(rownames(m_coef_mat)),
  row_names_gp = grid::gpar(fontsize = 20, fontface = "italic"),
  
  
  column_labels = c("Blast", "VNS", "Blast:VNS"),
  column_names_rot = 45,
  column_names_gp = grid::gpar(fontsize = 20),
  show_heatmap_legend = FALSE,
  show_column_dend = FALSE,
  cluster_columns = FALSE,
  
  # row_title_rot = 0,
  # row_title_gp = grid::gpar(fontsize = 20), 
  
  width  = unit(wide, "cm"),
  height = unit(high, "cm"),
  
  cell_fun = function(j, i, x, y, w, h, fill) {
    grid::grid.text(
      m_p_mat[i, j],
      x = x,
      y = y,
      gp = grid::gpar(
        fontsize = 24,
        col = "black",
        fontface = "bold"
      )
    )
  })

ht_m
```

```{r, fig.width = 7, fig.height = 6}
wide <- 2 * ncol(y_coef_mat)
high <- 2 * nrow(y_coef_mat)


# heatmap model Y coefficients
ht_y <- Heatmap(
  y_coef_mat,
  name = "Y Estimate",
  col = out_color,
  show_heatmap_legend = FALSE,
  left_annotation = fdr_annotation,
  
  row_split = row_exposure_clean,
  row_title = NULL,
  cluster_rows = TRUE,
  cluster_row_slices = TRUE,
  show_row_dend = FALSE,
  row_labels = clean_bug_names(rownames(y_coef_mat)),
  row_names_gp = grid::gpar(fontsize = 20, fontface = "italic"),
  
  column_labels = c("Blast", "VNS", "Blast:VNS", "microbe"),
  cluster_columns = FALSE,
  show_column_dend = FALSE,
  column_names_rot = 45,
  column_names_gp = grid::gpar(fontsize = 20),
  
  width  = unit(wide, "cm"),
  height = unit(high, "cm"),
  
  cell_fun = function(j, i, x, y, w, h, fill) {
    grid::grid.text(
      y_p_mat[i, j],
      x = x,
      y = y,
      gp = grid::gpar(
        fontsize = 24,
        col = "black",
        fontface = "bold"
      )
    )
  })

ht_y
```


```{r, fig.width = 10, fig.height = 6}
ht_list <- ht_m + ht_y

draw(ht_list, main_heatmap = "M Estimate")
```

Now save
```{r}
pdf("../figures/B_med_coef_heatmap.pdf", width = 10, height = 6)
ht_drawn <- draw(ht_list, main_heatmap = "M Estimate")
rows <- row_order(ht_drawn)
dev.off()
```

## Plotting blast mediation hits
*Doing this after heat maps so I can use same order as clustering*
```{r}
ht_row_order <- c(unlist(rows))
names(ht_row_order) <- NULL
ht_row_order

ci_row_order <- row_order[ht_row_order]
ci_row_order
```


```{r}
ci_plt <- ci_df %>%
  filter(mediator %in% ci_row_order) %>%
  mutate(
    exposure = factor(exposure, levels = c("blast", "vns")),
    mediator = factor(mediator, levels = ci_row_order),     # don't reverse
    outcome  = factor(outcome, levels = c("OFB_Clatency", "OFB_Cdistance", "PD_choice_filt", "EtOH_pref_24h_final")),
    exp_med_out  = as.character(interaction(exposure, mediator, outcome, sep = ";"))
    ) %>%
  pivot_longer(
    cols = matches("^(ACME|ADE|Total)_"),
    names_to = c("effect", "cond", ".value"),
    names_pattern = "^([^_]+)_([^_]+)_(.+)$"
  ) %>%
  mutate(
    effect = factor(effect, levels = c("ACME", "ADE", "Total")),
    cond   = factor(cond, levels = c("cond0", "cond1")),        # fct_recode(., "new" = "old")
    # exp_med_out = factor(paste0(exposure, ";", outcome, ";", mediator)),
    exp_med_out_cond = factor(paste0(exp_med_out, ";", cond))
    # microbe = factor(mediator)
  ) %>%
  arrange(mediator)

shade_bands <- ci_plt %>%
  group_by(mediator) %>% 
  arrange(by = mediator) %>%
  mutate(
    band_id = as.numeric(mediator),
    fill = ifelse(band_id %% 2 == 1, "grey95", "transparent")
    )

effect_colors <- c(
  "ADE"   = "#9C2121",
  "ACME"  = "#689835",
  "Total" = "#070E00"
)
```

```{r, fig.width = 7, fig.height = 3.5}

ci_absmax <- max(abs(c(ci_plt$upper, ci_plt$lower))) * 1.1

plt <- ggplot(ci_plt, aes(x = est, y = effect, color = effect)) +
  geom_rect(data = shade_bands, aes(
      xmin = -Inf, xmax = Inf,
      ymin = -Inf, ymax = Inf,
      fill = fill
      ), inherit.aes = FALSE, color = NA) +
  geom_point(size = 1.5) +
  geom_linerange(aes(xmin = lower, xmax = upper), linewidth = 0.4) +
  scale_fill_identity() +
  # geom_errorbarh(aes(xmin = mean - se, xmax = mean + se), linewidth = 0.3, height = 0.5) +
  scale_color_manual(values = effect_colors) +
  facet_grid(
    rows = vars(mediator), # if multiple exposure/outcomes, use exp_med_out
    cols   = vars(cond),
    space  = "free_y",
    # switch = "y", scales = "free_y",
    ) +
  scale_x_continuous(limits = c(-ci_absmax, ci_absmax)) +
  labs(x = NULL, y = NULL) +
  guides(fill = guide_legend(reverse = TRUE)) +
  theme_minimal(base_size = 14) +
  theme(
    strip.placement   = "outside",
    panel.spacing.y   = unit(0.1, "lines"),   # try 1-3 "lines" of spacing
    panel.spacing.x   = unit(0.1, "lines"),   # put close together
    axis.text.y       = element_blank(),
    axis.ticks.y      = element_blank(),
    axis.ticks.x      = element_line(color = "black", linewidth = 0.3),
    axis.line.x       = element_line(colour = "black", linewidth = 0.4),
    panel.grid        = element_blank(),
    panel.background  = element_blank()
  )

plt1 <- plt + theme(
  legend.position    = "none",
  strip.text.y.right = element_text(
    angle = 0, margin = margin(l = 4, unit = "pt"), hjust = 0
    )
  # strip.text.x.top = c("")
  )

plt2 <- plt + theme(legend.position = "none", strip.text.y.right = element_blank())

plt_leg <- plt + theme(strip.text.y.right = element_blank())
legend_only <- get_legend(plt_leg)

plt1
```

do as device instead (so I can save plot with legend)
```{r}
row_height  <- 0.5
panel_width <- 2

pdf("../figures/B_med_blast_cond_blocked.pdf",
    width = (panel_width*2) + 3,
    height = (nlevels(shade_bands$mediator) * row_height))

print(plt1)
plot(legend_only)
dev.off()
```

## Plot Blast (+/- VNS) and VNS (+/- Blast) separately
```{r}
blast_row_order <- ci_row_order[ci_row_order %in% ci_df$mediator[ci_df$exposure == "blast"]]

blast_ci_plt <- ci_df %>%
  filter(
    mediator %in% blast_row_order
    ) %>%
  mutate(
    exposure = factor(exposure, levels = c("blast", "vns")),
    mediator = factor(mediator, levels = blast_row_order),     # don't reverse
    outcome  = factor(outcome),  # , #levels = c("OFB_Clatency", "OFB_Cdistance", "PD_choice_filt", "EtOH_pref_24h_final")),
    exp_med_out  = as.character(interaction(exposure, mediator, outcome, sep = ";"))
    ) %>%
  pivot_longer(
    cols = matches("^(ACME|ADE|Total)_"),
    names_to = c("effect", "cond", ".value"),
    names_pattern = "^([^_]+)_([^_]+)_(.+)$"
  ) %>%
  mutate(
    effect = factor(effect, levels = c("ACME", "ADE", "Total")),
    cond   = factor(cond, levels = c("cond0", "cond1")),        # fct_recode(., "new" = "old")
    # exp_med_out = factor(paste0(exposure, ";", outcome, ";", mediator)),
    exp_med_out_cond = factor(paste0(exp_med_out, ";", cond))
    # microbe = factor(mediator)
  ) %>%
  arrange(mediator)

blast_shade_bands <- blast_ci_plt %>%
  group_by(mediator) %>% 
  arrange(by = mediator) %>%
  mutate(
    band_id = as.numeric(mediator),
    fill = ifelse(band_id %% 2 == 1, "grey95", "transparent")
    )

line_data <- data.frame(
  col_group = c("cond0", "cond1"),
  x_center  = c(0, 0) # These are your "centers"
)
```

```{r, fig.width = 7, fig.height = 3.5}

ci_absmax <- max(abs(c(ci_plt$upper, ci_plt$lower))) * 1.1

blast_cond <- c("VNS (-)", "VNS (+)")

blast_plt <- ggplot(blast_ci_plt, aes(x = est, y = effect, color = effect)) +
  geom_rect(data = blast_shade_bands, aes(
      xmin = -Inf, xmax = Inf,
      ymin = -Inf, ymax = Inf,
      fill = fill
      ), inherit.aes = FALSE, color = NA) +
  geom_point(size = 1.5) +
  geom_linerange(aes(xmin = lower, xmax = upper), linewidth = 0.4) +
  scale_fill_identity() +
  # geom_errorbarh(aes(xmin = mean - se, xmax = mean + se), linewidth = 0.3, height = 0.5) +
  scale_color_manual(values = effect_colors) +
  facet_grid(
    rows = vars(mediator), # if multiple exposure/outcomes, use exp_med_out
    cols   = vars(cond),
    labeller = labeller(cond = c("cond0" = "VNS (-)", "cond1" = "VNS (+)")),
    space  = "free_y",
    # switch = "y", scales = "free_y",
    ) +
  scale_x_continuous(limits = c(-ci_absmax, ci_absmax)) +
  labs(x = NULL, y = NULL) +
  guides(fill = guide_legend(reverse = TRUE)) +
  
  geom_vline(data = line_data, 
             aes(xintercept = x_center), 
             linetype = "dashed", 
             color = "darkgrey", 
             size = 0.5, alpha = 0.5) +
  
  theme_minimal(base_size = 14) +
  theme(
    strip.placement   = "outside",
    panel.spacing.y   = unit(0.1, "lines"),   # try 1-3 "lines" of spacing
    panel.spacing.x   = unit(0.1, "lines"),   # put close together
    axis.text.y       = element_blank(),
    axis.ticks.y      = element_blank(),
    axis.ticks.x      = element_line(color = "black", linewidth = 0.3),
    axis.line.x       = element_line(colour = "black", linewidth = 0.4),
    panel.grid        = element_blank(),
    panel.background  = element_blank()
  )

blast_plt1 <- blast_plt + theme(
  legend.position    = "none",
  strip.text.y.right = element_text(
    angle = 0, margin = margin(l = 4, unit = "pt"), hjust = 0
    ),
  strip.text.x.top = element_text()
  )

blast_plt2 <- blast_plt + theme(legend.position = "none", strip.text.y.right = element_blank())

blast_plt_leg <- blast_plt + theme(strip.text.y.right = element_blank())
blast_legend_only <- get_legend(blast_plt_leg)

blast_plt1
```

Repeat for VNS
```{r}
vns_row_order <- ci_row_order[ci_row_order %in% ci_df$mediator[ci_df$exposure == "vns"]]

vns_ci_plt <- ci_df %>%
  filter(
    mediator %in% vns_row_order
    ) %>%
  mutate(
    exposure = factor(exposure, levels = c("blast", "vns")),
    mediator = factor(mediator, levels = vns_row_order),     # don't reverse
    outcome  = factor(outcome),  # , #levels = c("OFB_Clatency", "OFB_Cdistance", "PD_choice_filt", "EtOH_pref_24h_final")),
    exp_med_out  = as.character(interaction(exposure, mediator, outcome, sep = ";"))
    ) %>%
  pivot_longer(
    cols = matches("^(ACME|ADE|Total)_"),
    names_to = c("effect", "cond", ".value"),
    names_pattern = "^([^_]+)_([^_]+)_(.+)$"
  ) %>%
  mutate(
    effect = factor(effect, levels = c("ACME", "ADE", "Total")),
    cond   = factor(cond, levels = c("cond0", "cond1")),        # fct_recode(., "new" = "old")
    # exp_med_out = factor(paste0(exposure, ";", outcome, ";", mediator)),
    exp_med_out_cond = factor(paste0(exp_med_out, ";", cond))
    # microbe = factor(mediator)
  ) %>%
  arrange(mediator)

vns_shade_bands <- vns_ci_plt %>%
  group_by(mediator) %>% 
  arrange(by = mediator) %>%
  mutate(
    band_id = as.numeric(mediator),
    fill = ifelse(band_id %% 2 == 1, "grey95", "transparent")
    )

line_data <- data.frame(
  col_group = c("cond0", "cond1"),
  x_center  = c(0, 0) # These are your "centers"
)
```

```{r, fig.width = 7, fig.height = 1.5}

ci_absmax <- max(abs(c(ci_plt$upper, ci_plt$lower))) * 1.1

vns_cond <- c("Blast (-)", "Blast (+)")

vns_plt <- ggplot(vns_ci_plt, aes(x = est, y = effect, color = effect)) +
  geom_rect(data = vns_shade_bands, aes(
      xmin = -Inf, xmax = Inf,
      ymin = -Inf, ymax = Inf,
      fill = fill
      ), inherit.aes = FALSE, color = NA) +
  geom_point(size = 1.5) +
  geom_linerange(aes(xmin = lower, xmax = upper), linewidth = 0.4) +
  scale_fill_identity() +
  # geom_errorbarh(aes(xmin = mean - se, xmax = mean + se), linewidth = 0.3, height = 0.5) +
  scale_color_manual(values = effect_colors) +
  facet_grid(
    rows = vars(mediator), # if multiple exposure/outcomes, use exp_med_out
    cols   = vars(cond),
    labeller = labeller(cond = c("cond0" = "Blast (-)", "cond1" = "Blast (+)")),
    space  = "free_y",
    # switch = "y", scales = "free_y",
    ) +
  scale_x_continuous(limits = c(-ci_absmax, ci_absmax)) +
  labs(x = NULL, y = NULL) +
  guides(fill = guide_legend(reverse = TRUE)) +
  
  geom_vline(data = line_data, 
             aes(xintercept = x_center), 
             linetype = "dashed", 
             color = "darkgrey", 
             size = 0.5, alpha = 0.5) +
  
  theme_minimal(base_size = 14) +
  theme(
    strip.placement   = "outside",
    panel.spacing.y   = unit(0.1, "lines"),   # try 1-3 "lines" of spacing
    panel.spacing.x   = unit(0.1, "lines"),   # put close together
    axis.text.y       = element_blank(),
    axis.ticks.y      = element_blank(),
    axis.ticks.x      = element_line(color = "black", linewidth = 0.3),
    axis.line.x       = element_line(colour = "black", linewidth = 0.4),
    panel.grid        = element_blank(),
    panel.background  = element_blank()
  )

vns_plt1 <- vns_plt + theme(
  legend.position    = "none",
  strip.text.y.right = element_text(
    angle = 0, margin = margin(l = 4, unit = "pt"), hjust = 0
    ),
  strip.text.x.top = element_text()
  )

# vns_plt2 <- vns_plt + theme(legend.position = "none", strip.text.y.right = element_blank())
# 
# vns_plt_leg <- vns_plt + theme(strip.text.y.right = element_blank())
# vns_legend_only <- get_legend(vns_plt_leg)

vns_plt1
```

Save both
```{r}
row_height  <- 0.5
panel_width <- 2

# 1. Calculate heights
h1 <- nlevels(blast_shade_bands$mediator) * row_height
h2 <- nlevels(vns_shade_bands$mediator) * row_height

combined_plot <- vns_plt1 / blast_plt1 +
  plot_layout(heights = c(h2, h1))
```

```{r, fig.width = 7, fig.height = 3.1}
print(combined_plot)
```

```{r}
pdf_width <- (panel_width * 2) + 3  # add wiggle room for row labels
pdf_height <- h1 + h2 + 1 # add wiggle room for column labels + x-axis text

# Open PDF with TOTAL height
pdf("../figures/B_CIs_combined.pdf",
    width = pdf_width,
    height = pdf_height) # Sum of heights

print(combined_plot)
dev.off()
```


# Random
### Testing plots
```{r, fig.width = 5.5, fig.height = 5}
# testplot <- residout_residbug(med_hits[[1, "mediator"]], df, med_hits[[1, "outcome"]])
# print(testplot)
# 
# testplot <- box_easier(data = df, dv = "GGB28883_SGB41564", y_lab = "GGB28883_SGB41564")
# print(testplot)

x_y <- plot_adjusted_scatter(
  data = df,
  x_var = med_hits[[1, "mediator"]],  # e.g., "bug_name"
  y_var = med_hits[[1, "outcome"]],   # e.g., "outcome_name"
  covariate_formula = "~ blast + vns + blast_and_vns + (1|Round_main)",
  residualize_x = FALSE,
  residualize_y = FALSE
)

x_y

x_ry <- plot_adjusted_scatter(
  data = df,
  x_var = med_hits[[1, "mediator"]],  # e.g., "bug_name"
  y_var = med_hits[[1, "outcome"]],   # e.g., "outcome_name"
  covariate_formula = "~ blast + vns + blast_and_vns + (1|Round_main)",
  residualize_x = FALSE,
  residualize_y = TRUE
)

x_ry

rx_ry <- plot_adjusted_scatter(
  data = df,
  x_var = med_hits[[1, "mediator"]],  # e.g., "bug_name"
  y_var = med_hits[[1, "outcome"]],   # e.g., "outcome_name"
  covariate_formula = "~ blast + vns + blast_and_vns + (1|Round_main)",
  residualize_x = TRUE,
  residualize_y = TRUE
)

rx_ry


clean_out_names("OFB_Cdistance")
grepl("\n", clean_out_names("OFB_Cdistance"))
nchar(sub("\n.*", "", clean_out_names("OFB_Cdistance")))


```

OK so SGB41455 is associated with increased startle_habituation (+0.402),
which is about the same effect size as blast, *but significant independent of blast*
```{r}
# form <- as.formula("startle_habituation ~ GGB28802_SGB41455 + blast + vns + blast_and_vns + (1 | Round_main)")
# m    <- lmer(form, data = df, REML = FALSE, na.action = "na.exclude")
# summary(m)


form <- as.formula("startle_PPI ~ GGB42598_SGB59794 + blast + vns + blast_and_vns + (1 | Round_main)")
m    <- lmer(form, data = df, REML = FALSE, na.action = "na.exclude")
summary(m)
```

We could use bootstrapping to estimate these effects (slopes) as well. 
##### Minimum microbe p-value for each outcome?
```{r, fig.width = 4, fig.height = 4}
for (out in model_outcome) {
  p_df <- data.frame(
    microbe = out$microbe[out$term == "microbe"],
    p.value = out$p.value[out$term == "microbe"]
  )
  
  min_p <- min(p_df$p.value)
  outcome_name <- unique(out$outcome)
  
  p_hist <- ggplot(data.frame(p_df), aes(x = p.value)) +
    geom_histogram(bins = 20, color = "white") +
    labs(
      title = paste0(outcome_name, ", min p = ", round(min_p, digits = 4)),
      x     = "p_microbe (outcome ~ microbe + treatments)",
      y     = "count"
    ) + theme_bw()
  
  print(p_hist)
}
```

### Plotting blast mediation hits (OLD)
```{r}
ci_plt <- ci_df %>%
  pivot_longer(
    cols = matches("^(ACME|ADE|Total)_"),
    names_to = c("effect", "cond", ".value"),
    names_pattern = "^([^_]+)_([^_]+)_(.+)$"
  ) %>%
  mutate(
    effect = factor(effect, levels = c("ACME", "ADE", "Total")),
    cond   = factor(cond, levels = c("cond0", "cond1")),
    exp_med_out = factor(paste0(exposure, ";", outcome, ";", mediator)),
    exp_med_out_cond = factor(paste0(exp_med_out, ";", cond)),
    microbe = factor(mediator)
  ) %>%
  arrange(effect)

shade_bands <- ci_plt %>%
  group_by(exp_med_out) %>% 
  arrange(by = exp_med_out) %>%
  mutate(
    band_id = as.numeric(exp_med_out),
    fill = ifelse(band_id %% 2 == 1, "grey95", "transparent")
    )

effect_colors <- c(
  "ADE"   = "#9C2121",
  "ACME"  = "#689835",
  "Total" = "#070E00"
)
```

```{r, fig.width = 7, fig.height = 3}
plt <- ggplot(ci_plt, aes(x = est, y = effect, color = effect)) +
  geom_rect(data = shade_bands, aes(
      xmin = -Inf, xmax = Inf,
      ymin = -Inf, ymax = Inf,
      fill = fill
      ), inherit.aes = FALSE, color = NA) +
  geom_point(size = 1.5) +
  geom_linerange(aes(xmin = lower, xmax = upper), linewidth = 0.4) +
  scale_fill_identity() +
  # geom_errorbarh(aes(xmin = mean - se, xmax = mean + se), linewidth = 0.3, height = 0.5) +
  scale_color_manual(values = effect_colors) +
  facet_grid(
    rows = vars(exp_med_out), # if multiple exposure/outcomes, use exp_med_out
    cols   = vars(cond),
    space  = "free_y",
    # switch = "y", scales = "free_y",
    ) +
  # scale_x_continuous(limits = c(xmin, xmax)) +
  labs(x = NULL, y = NULL) +
  guides(fill = guide_legend(reverse = TRUE)) +
  theme_minimal(base_size = 14) +
  theme(
    strip.placement   = "outside",
    panel.spacing.y   = unit(0.1, "lines"),   # try 1-3 "lines" of spacing
    panel.spacing.x   = unit(0.1, "lines"),   # put close together
    axis.text.y       = element_blank(),
    axis.ticks.y      = element_blank(),
    axis.ticks.x      = element_line(color = "black", linewidth = 0.3),
    axis.line.x       = element_line(colour = "black", linewidth = 0.4),
    panel.grid        = element_blank(),
    panel.background  = element_blank()
  )

plt1 <- plt + theme(
  legend.position    = "none",
  strip.text.y.right = element_text(
    angle = 0, margin = margin(l = 4, unit = "pt"), hjust = 0
  ))

plt2 <- plt + theme(legend.position = "none", strip.text.y.right = element_blank())

plt_leg <- plt + theme(strip.text.y.right = element_blank())
legend_only <- get_legend(plt_leg)

plt1
```


do as device instead (so I can save plot with legend)
```{r}
row_height  <- 0.5
panel_width <- 2

pdf("../figures/B_med_cond_cases.pdf",
    width = (panel_width*2) + 3,
    height = (nlevels(shade_bands$exp_med_out) * row_height))

print(plt1)
plot(legend_only)
dev.off()
```


### Make heatmap to go along with this (OLD)

```{r, fig.width = 5, fig.height = 4}
order_bugs <- c(
  "bacterium_1XD42_54", "Clostridiales_bacterium", "GGB29002_SGB41718",
  "GGB3171_SGB4185", "GGB28898_SGB41580"
)

trt_effect_matrix <- model_microbe %>%
  filter(microbe %in% med_hits$mediator) %>%
  mutate(microbe = factor(microbe, levels = order_bugs)) %>%
  dplyr::select(microbe, term, estimate) %>%
  pivot_wider(names_from = term, values_from = estimate) %>%
  arrange(microbe) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(c("blast", "vns", "blast_and_vns")) %>%
  as.matrix()

trt_p_matrix <- model_microbe %>%
  filter(microbe %in% med_hits$mediator) %>%
  mutate(
    microbe = factor(microbe, levels = order_bugs),
    p.ann = case_when(
      p.value < 0.005 ~ "**",
      p.value < 0.05 ~ "*",
      .default = ""
      )
    ) %>%
  dplyr::select(microbe, term, p.ann) %>%
  pivot_wider(names_from = term, values_from = p.ann) %>%
  arrange(microbe) %>%
  column_to_rownames("microbe") %>%
  as.matrix()

absmax <- max(abs(trt_effect_matrix))
col_fun <- colorRamp2(
  c(-absmax, 0, absmax),
  c("blue", "white", "red")
)

ht <- Heatmap(trt_effect_matrix, name = "Estimate", col = col_fun,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_dend = FALSE,
  show_column_dend = FALSE,
  row_order = order_bugs,
  cell_fun = function(j, i, x, y, w, h, fill) {
    grid::grid.text(
      trt_p_matrix[i, j],
      x = x,
      y = y,
      gp = grid::gpar(
        fontsize = 14,
        col = "black",
        fontface = "bold"
      )
    )
  })
```


```{r, fig.width = 5, fig.height = 5}
ht
```

```{r}
pdf("../figures/med_coef_heatmap.pdf", width = 5, height = 5)
ht_drawn <- draw(ht)
dev.off()
```

#### Also plot residual outcome vs. residual bug for the mediators
I want to plot residual outcome vs. bug
```{r}
residout_residbug <- function(bug_name, data, outcome_name) {
  plt <- data.frame(
    row.names     = data$seq_name,
    bug_clr       = data[[bug_name]],
    inj_trt_group = data$inj_trt_group,
    blast         = data$blast,
    vns           = data$vns,
    blast_and_vns = data$blast_and_vns,
    Round_main    = data$Round_main,
    outcome       = data[[outcome_name]]
  ) %>%
    filter(!is.na(outcome))
  
  formx <- as.formula("bug_clr ~ blast + vns + blast_and_vns + (1 | Round_main)")
  formy <- as.formula("outcome ~ blast + vns + blast_and_vns + (1 | Round_main)")
  fitx <- lmer(formula = formx, data = plt)
  fity <- lmer(formula = formy, data = plt)
  
  plt$bug_resid <- resid(fitx)
  plt$out_resid <- resid(fity)
  
  ggplot(plt, aes(x = bug_resid, y = out_resid)) +     # fill = inj_trt_group, shape = Round_main
    geom_point(aes(color = inj_trt_group), size = 4, alpha = 0.7) +
    scale_fill_manual(values = injtrt_colors) +
    # scale_shape_manual(values = batch_shapes) +
    # geom_smooth(aes(color = inj_trt_group), method = "lm", se = FALSE, alpha = 0.2) +
    scale_color_manual(values = injtrt_colors) +
    geom_smooth(method = "lm", se = FALSE, color = "black",linetype = "dashed") +
    labs(
      x = paste0(bug_name, " Abundance"),
      y = paste0("Residual normalized ", outcome_name)
    ) +
    theme_bw() + theme(legend.position = "none")
}
```

Save all these to a single pdf
```{r, fig.width = 4, fig.height = 4}
resid_resid_plots <- list()
pdf("../figures/B_outcome_resid_resid_plots.pdf", width = 5, height = 5)

for (i in 1:nrow(med_hits)) {
  resid_resid_plots[[i]] <- residout_residbug(med_hits[[i, "mediator"]], df, med_hits[[i, "outcome"]])
  print(resid_resid_plots[[i]])
}

# Close the PDF device to save
dev.off()
```
