---
title: "microbe_summary_plot.Rmd"
author: "Alyssa N. Easton"
date:   '2025-10-15'
output: 
---

# Setup
```{r "setup", include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
knitr::opts_knit$set(root.dir = file.path(Sys.getenv("VNS_ROOT"), "data"))  # stored VNS_ROOT in ~/.Renviron
options(expressions = 6000)  # Increase allowed recursion depth
options(digits = 10) ## Formats output to 10 digits
library(magrittr)
library(tidyverse)
library(moments)
library(broom)
library(broom.helpers)
library(HDMT)
library(rlang)
library(glue)
library(janitor)
library(openxlsx)
library(purrr)
library(cowplot)
library(ggh4x)
library(svglite)
library(ggforestplot)
library(ComplexHeatmap)
source("../scripts/mdact.R")
source("../scripts/june_functions.R")
source("/proj/gibbons/2024_easton_vns/scripts/shannon_div_functions.R")
```

Data imports
```{r}
data <- readRDS("/proj/gibbons/2024_easton_vns/data/1_preprocessed/species_october.rds")
# lmer <- readRDS("/proj/gibbons/2024_easton_vns/data/lmer/all_diffabund_new.rds")
lmer_a <- readRDS("/proj/gibbons/2024_easton_vns/data/lmer/A_microbe_2var.rds")
lmer_b <- readRDS("/proj/gibbons/2024_easton_vns/data/lmer/B_microbe_2var.rds")
pca <- readRDS("PCA/pca_A_cytokines.rds")
```

```{r}

```


# Make function
```{r}
summary_plot <- function(bugs, clr_data, meta) {
  bug_data <- clr_data %>%
    dplyr::select(all_of(bugs)) %>%
    scale() %>% as.data.frame() %>%       # scale
    rownames_to_column("seq_name") %>%
    left_join(
      meta %>%
        dplyr::select(c("seq_name", "inj_trt_group")),
      by = "seq_name"
      ) %>%
    column_to_rownames("seq_name")
    
  plot_bug <- bug_data %>%
    pivot_longer(
      cols = -inj_trt_group,
      names_to = "microbe",
      values_to = "relab"
    )

  summary_bug <- plot_bug %>%
    group_by(microbe, inj_trt_group) %>%
    summarise(
      mean = mean(relab, na.rm = TRUE),
      se   = sd(relab, na.rm = TRUE) / sqrt(n()),
      .groups = "drop"
    )
  
  summary_bug <- summary_bug %>%
    # define custom factor order for bugs
    mutate(
      # make an ordered factor so all groups of a microbe stay together
      y_label = factor(
        paste(microbe, inj_trt_group, sep = "_"),
        levels = summary_bug %>%
          arrange(microbe, desc(inj_trt_group)) %>%                       # descending groups
          transmute(lvl = paste(microbe, inj_trt_group, sep = "_")) %>%
          pull(lvl)
      ),
      inj_trt_group = fct_rev(inj_trt_group),                        # reverse order groups (top-bottom)
      microbe       = factor(microbe, levels = bugs)                 # don't need to reverse order of microbes?
    )
  
  # compute midpoints for each variable's block
  microbe_labels <- summary_bug %>%
    group_by(microbe) %>%
    summarise(y_mid = mean(as.numeric(y_label)))
  
  # Define alternating shading bands
  shade_bands <- summary_bug %>%
    group_by(microbe) %>% 
    arrange(by = microbe) %>%
    mutate(
      band_id = as.numeric(microbe),
      fill = ifelse(band_id %% 2 == 1, "grey95", "white")
      )
  
  # calulate limits 
  # xlim <- max(abs(c(summary_bug$mean + summary_bug$se, summary_bug$mean - summary_bug$se))) * 1.01
  xmax <- max(summary_bug$mean + summary_bug$se) * 1.01
  xmin <- min(summary_bug$mean - summary_bug$se) * 1.01 # assuming negative
  # plot
  plt <- ggplot(summary_bug, aes(x = mean, y = inj_trt_group, color = inj_trt_group)) +
    geom_rect(
      data = shade_bands,
      aes(
        xmin = -Inf, xmax = Inf,
        ymin = -Inf, ymax = Inf,
        fill = fill
      ),
      inherit.aes = FALSE,
      color = NA
    ) +
    geom_vline(xintercept =  0.0, color = "grey25", alpha = 0.7, linetype = "dashed", linewidth = 0.2) +
    geom_vline(xintercept =  1.0, color = "grey50", alpha = 0.7, linetype = "dashed", linewidth = 0.2) +
    geom_vline(xintercept = -1.0, color = "grey50", alpha = 0.7, linetype = "dashed", linewidth = 0.2) +
    geom_vline(xintercept =  0.5, color = "grey75", alpha = 0.5, linetype = "dashed", linewidth = 0.2) +
    geom_vline(xintercept = -0.5, color = "grey75", alpha = 0.5, linetype = "dashed", linewidth = 0.2) +
    # geom_vline(xintercept =  1.5, color = "grey75", alpha = 0.5, linetype = "dashed", linewidth = 0.2) +
    # geom_vline(xintercept = -1.5, color = "grey75", alpha = 0.5, linetype = "dashed", linewidth = 0.2) +
    geom_point(size = 1.0) +
    geom_linerange(aes(xmin = mean - se, xmax = mean + se), linewidth = 0.25) +
    scale_fill_identity() +
    # geom_errorbarh(aes(xmin = mean - se, xmax = mean + se), linewidth = 0.3, height = 0.5) +
    scale_color_manual(values = four_colors) +
    facet_grid(
      rows   = vars(microbe),
      # cols   = vars(location),
      # scales = "free_y",
      space  = "free_y",
      switch = "y"            # plot y-labels on left-hand side (instead of right)
      ) +
    scale_x_continuous(limits = c(xmin, xmax)) +
    labs(x = NULL, y = NULL) +
    theme_minimal(base_size = 14) +
    theme(
      strip.placement   = "outside",
      strip.text.y.left = element_text(angle = 0, hjust = 1),
      panel.spacing.y   = unit(0.1, "lines"),   # try 1-3 "lines" of spacing
      panel.spacing.x   = unit(0.1, "lines"),   # put close together
      axis.text.y       = element_blank(),
      axis.ticks.y      = element_blank(),
      axis.line.x       = element_line(colour = "black", linewidth = 0.4),
      panel.grid        = element_blank(),
      panel.background  = element_blank(),
      legend.position   = "none"
    )
  
  return(plt)
}

# # removed from function, can be added afterwards
# geom_vline(xintercept =  1.5, color = "grey75", alpha = 0.5, linetype = "dashed", linewidth = 0.2) +
# geom_vline(xintercept = -1.5, color = "grey75", alpha = 0.5, linetype = "dashed", linewidth = 0.2) +


```

Test the function
```{r}
a_med_bugs <- c("GGB28379_SGB40959", "Streptococcus_danieliae", "GGB29685_SGB42494")

plt <- summary_plot(
  bugs     = a_med_bugs,
  clr_data = data$a_clr,
  meta     = data$a_meta
) + theme(strip.text.y.left = element_blank())

# ggsave("../figures/a_med_bugs.png", plt, width = 1.5, height = 1.2)
# ggsave("../figures/a_med_bugs.svg", plt, width = 1.5, height = 1.2, device = "svg")
```

```{r}
b_med_bug <- "bacterium_1XD42_54"

plt <- summary_plot(
  bugs     = b_med_bug,
  clr_data = data$b_clr,
  meta     = data$b_meta
) + theme(strip.text.y.left = element_blank())

# ggsave("../figures/b_med_bugb.png", plt, width = 1.5, height = 0.7)
# ggsave("../figures/b_med_bug.svg", plt, width = 1.5, height = 0.7, device = "svg")
```

Get more info about these bugs (Family-level, more annotation)
```{r}
a_med_bugs
data$taxonomy_map %>%
  filter(Species %in% a_med_bugs)
```
They are all firmicutes. We have Clostridia, Bacilli, and unknown Firmicutes. On the Family-level, this is Eubacteriaceae, Streptococcaceae, and an unknown Family of Firmicutes.

All species I've plotted in manuscript
```{r}
allspec <- c(
  "Clostridiales_bacterium",
  "Romboutsia_ilealis",
  "Clostridium_cocleatum",
  "Lachnospiraceae_bacterium_MD308",
  "bacterium_1XD42_54",
  "Alistipes_sp_DSM_112343",
  "Lactobacillus_johnsonii",
  "Lachnospiraceae_bacterium",
  "Lachnospiraceae_bacterium_MD308",
  "Eubacteriaceae_bacterium",
  "Muribaculaceae_bacterium",
  "GGB28991_SGB41705",
  "GGB29685_SGB42494",
  "GGB30461_SGB63209",
  "GGB30463_SGB43537",
  "GGB46159_SGB63927",
  "GGB46159_SGB63927",
  "GGB70852_SGB95305",
  "GGB27878_SGB40312",
  "GGB27918_SGB40356",
  "GGB28379_SGB40959",
  "Streptococcus_danieliae",
  "GGB29685_SGB42494"
)

plot_bugs <- data$taxonomy_map %>%
  filter(Species %in% allspec)

unique(plot_bugs$Order)
```

# Just summary plot
## cohort A: which microbes do we want to plot?
```{r}
bugs <- lmer_a$lmer_res %>%  # lmer$a_main$lmer_res %>%
  mutate(
    y = gsub("s__", "", y)
  ) %>%
  filter(
    term == "blast",
    signif == TRUE
    ) %>%
  pull(y)

med_bugs <- c("GGB28379_SGB40959", "Streptococcus_danieliae")

bugs <- c(bugs, med_bugs)
bugs
```

Out of hits from A, how present are they in A versus B?
```{r}
a_ab <- data$a_relabund %>%
  dplyr::select(all_of(bugs)) %>%
  mutate(across(everything(), ~replace_na(., 0)))

b_ab <- data$b_relabund %>%
  dplyr::select(any_of(bugs)) %>%
  mutate(across(everything(), ~replace_na(., 0)))


return.presence(a_ab, colnames(a_ab))
return.presence(b_ab, colnames(b_ab))
```


```{r}
bug_data <- data$a_clr %>%
  dplyr::select(all_of(bugs)) %>%
  scale() %>% as.data.frame() %>%       # scale
  rownames_to_column("seq_name") %>%
  left_join(
    data$a_meta %>%
      dplyr::select(c("seq_name", "inj_trt_group")),
    by = "seq_name"
    ) %>%
  column_to_rownames("seq_name")
  
plot_bug <- bug_data %>%
  pivot_longer(
    cols = -inj_trt_group,
    names_to = "microbe",
    values_to = "relab"
  )

head(plot_bug)
```

```{r}
summary_bug <- plot_bug %>%
  group_by(microbe, inj_trt_group) %>%
  summarise(
    mean = mean(relab, na.rm = TRUE),
    se   = sd(relab, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

summary_bug <- summary_bug %>%
  # define custom factor order for bugs
  mutate(
    # make an ordered factor so all groups of a microbe stay together
    y_label = factor(
      paste(microbe, inj_trt_group, sep = "_"),
      levels = summary_bug %>%
        arrange(microbe, desc(inj_trt_group)) %>%                       # descending groups
        transmute(lvl = paste(microbe, inj_trt_group, sep = "_")) %>%
        pull(lvl)
    ),
    inj_trt_group = fct_rev(inj_trt_group),      # reverse order groups (top-bottom)
  )

head(summary_bug)
```
Labels for y-axis
```{r}
# compute midpoints for each variable's block
microbe_labels <- summary_bug %>%
  group_by(microbe) %>%
  summarise(y_mid = mean(as.numeric(y_label)))

# Define alternating shading bands
shade_bands <- summary_bug %>%
  group_by(microbe) %>% 
  arrange(by = microbe) %>%
  mutate(
    microbe = as.factor(microbe),
    band_id = as.numeric(microbe),
    fill = ifelse(band_id %% 2 == 1, "grey95", "white")
    )
```

Plot
```{r}
xlim <- max(abs(c(summary_bug$mean + summary_bug$se, summary_bug$mean - summary_bug$se))) * 1.01

plt <- ggplot(summary_bug, aes(x = mean, y = inj_trt_group, color = inj_trt_group)) +
  geom_rect(
    data = shade_bands,
    aes(
      xmin = -Inf, xmax = Inf,
      ymin = -Inf, ymax = Inf,
      fill = fill
    ),
    inherit.aes = FALSE,
    color = NA
  ) +
  # geom_vline(xintercept =  0.0, color = "grey25", alpha = 0.7, linetype = "dashed", linewidth = 0.2) +
  # geom_vline(xintercept =  1.0, color = "grey50", alpha = 0.7, linetype = "dashed", linewidth = 0.2) +
  # geom_vline(xintercept = -1.0, color = "grey50", alpha = 0.7, linetype = "dashed", linewidth = 0.2) +
  # geom_vline(xintercept =  0.5, color = "grey75", alpha = 0.5, linetype = "dashed", linewidth = 0.2) +
  # geom_vline(xintercept = -0.5, color = "grey75", alpha = 0.5, linetype = "dashed", linewidth = 0.2) +
  # geom_vline(xintercept =  1.5, color = "grey75", alpha = 0.5, linetype = "dashed", linewidth = 0.2) +
  # geom_vline(xintercept = -1.5, color = "grey75", alpha = 0.5, linetype = "dashed", linewidth = 0.2) +
  geom_point(size = 1.0) +
  geom_linerange(aes(xmin = mean - se, xmax = mean + se), linewidth = 0.25) +
  scale_fill_identity() +
  # geom_errorbarh(aes(xmin = mean - se, xmax = mean + se), linewidth = 0.3, height = 0.5) +
  scale_color_manual(values = four_colors) +
  facet_grid(
    rows   = vars(microbe),
    # cols   = vars(location),
    # scales = "free_y",
    space  = "free_y",
    switch = "y"            # plot y-labels on left-hand side (instead of right)
    ) +
  scale_x_continuous(limits = c(-xlim, xlim)) +
  geom_vline(xintercept =  0.0, color = "grey25", alpha = 0.7, linetype = "dashed", linewidth = 0.2) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 14) +
  theme(
    strip.placement   = "outside",
    strip.text.y.left = element_text(angle = 0, hjust = 1),
    panel.spacing.y   = unit(0.1, "lines"),   # try 1-3 "lines" of spacing
    panel.spacing.x   = unit(0.1, "lines"),   # put close together
    axis.text.y       = element_blank(),
    axis.ticks.y      = element_blank(),
    axis.line.x       = element_line(colour = "black", linewidth = 0.4),
    panel.grid        = element_blank(),
    panel.background  = element_blank(),
    legend.position   = "none"
  )

plt

# ggsave("../figures/bug_summary_test.png", plt, width = 4.5, height = 7)
# ggsave("../figures/bug_summary.svg", plt, width = 4.5, height = 7, device = "svg")

```
# summary + lmer plot
## Try adding lmer plot as well!

```{r}
bugs <- lmer_a$lmer_res %>% # lmer$a_main$lmer_res %>%
  mutate(y = gsub("s__", "", y)) %>%
  filter(
    term == "blast",
    signif == TRUE
    ) %>%
  pull(y)

bugs
```


Group mean abundance dataframe
```{r}
group_abund <- data$a_clr %>%
  dplyr::select(all_of(bugs)) %>%
  scale() %>% as.data.frame() %>%       # scale/center CLR values
  rownames_to_column("seq_name") %>%
  left_join(                            # join with metadata
    data$a_meta %>%
      dplyr::select(c("seq_name", "inj_trt_group")),
    by = "seq_name"
    ) %>%
  column_to_rownames("seq_name") %>%
  pivot_longer(
    cols = -inj_trt_group,
    names_to = "microbe",
    values_to = "relab"
  ) %>%
  group_by(microbe, inj_trt_group) %>%
  summarise(
    mean = mean(relab, na.rm = TRUE),
    se   = sd(relab, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  mutate(
    panel = "GroupMeans",
    y_label = paste(microbe, inj_trt_group, sep = "_")
  )

head(group_abund)
```
Blast effect CI dataframe
```{r}
blastCI <- lmer_a$lmer_res %>% # lmer$a_main$lmer_res %>%
  mutate(y = gsub("s__", "", y)) %>%
  filter(
    term == "blast",
    y %in% bugs
    ) %>%
  mutate(microbe = fct_reorder(factor(y), estimate)) %>%
  transmute(
    microbe,
    # inj_trt_group = term,                     # for consistency in grouping/color
    mean = estimate,                            # match variable name
    se   = se,
    y_label = paste(microbe, term, sep = "_"),  # this is for if you have multiple terms
    panel = "Regression"                        # tag for facet column
  )

# save this bug order?
microbe_levels <- levels(blastCI$microbe)

head(blastCI)
```

get order-level annotations
```{r}
data$taxonomy_map %>%
  filter(Species %in% microbe_levels)

microbe_levels
```

```{r}
combined <- bind_rows(group_abund, blastCI) %>%
  mutate(
    microbe = factor(microbe, levels = microbe_levels),
    inj_trt_group = fct_rev(inj_trt_group)
    )


# Define alternating shading bands
shade_bands <- combined %>%
  group_by(microbe) %>% 
  arrange(by = microbe) %>%
  mutate(
    microbe = as.factor(microbe),
    band_id = as.numeric(microbe),
    fill = ifelse(band_id %% 2 == 1, "grey95", "white")
    )

head(combined)
head(shade_bands)
```
Define alternating shading bands
cyto_bands <- cyto_summary %>%
  distinct(variable, location, inj_trt_group) %>%
  group_by(variable) %>% 
  arrange(by = variable) %>%
  mutate(
    band_id = as.numeric(variable),
    fill = ifelse(band_id %% 2 == 1, "grey95", "white")
  ) %>%
  ungroup()

```{r}
plt <- ggplot(combined, aes(x = mean, y = inj_trt_group, color = inj_trt_group)) +
  geom_rect(
    data = shade_bands,
    aes(
      xmin = -Inf, xmax = Inf,
      ymin = -Inf, ymax = Inf,
      fill = fill
    ),
    inherit.aes = FALSE,
    color = NA
  ) +
  geom_point(size = 1.0) +
  geom_linerange(aes(xmin = mean - se, xmax = mean + se), linewidth = 0.25) +
  scale_fill_identity() +
  scale_color_manual(values = four_colors) +
  facet_grid(
    rows   = vars(microbe),
    cols   = vars(panel),
    space  = "free_y",
    scales = "free_x",
    switch = "y"
  ) +
  # scale_x_continuous(limits = c(-xlim, xlim)) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 14) +
  theme(
    strip.placement   = "outside",
    strip.text.y.left = element_text(angle = 0, hjust = 1),
    panel.spacing     = unit(0.1, "lines"),
    axis.text.y       = element_blank(),
    axis.ticks.y      = element_blank(),
    axis.ticks.x      = element_line(color = "black", linewidth = 0.4),
    axis.line.x       = element_line(colour = "black", linewidth = 0.4),
    panel.grid        = element_blank(),
    panel.background  = element_blank(),
    legend.position   = "none"
  )

plt
```
Fiddling with scaling parameters
```{r}
row_height <- 0.4  # inches per microbe row (adjust for your density)
n_microbes <- nlevels(combined$microbe)
height_in <- n_microbes * row_height

ggsave(
  "../figures/A_group_DA.svg",
  plot = plt,
  width = 6,           # fixed width (inches)
  height = height_in,  # scales with number of rows
  units = "in",
  device = "svg",
  dpi = 300
)

```

# Cohort B
```{r}
bugs <- lmer_b$lmer_res %>% # lmer$a_main$lmer_res %>%
  mutate(y = gsub("s__", "", y)) %>%
  filter(
    term == "blast",
    signif == TRUE
    ) %>%
  pull(y)

bugs
```

Group mean abundance dataframe
```{r}
group_abund <- data$b_clr %>%
  dplyr::select(all_of(bugs)) %>%
  scale() %>% as.data.frame() %>%       # scale/center CLR values
  rownames_to_column("seq_name") %>%
  left_join(                            # join with metadata
    data$b_meta %>%
      dplyr::select(c("seq_name", "inj_trt_group")),
    by = "seq_name"
    ) %>%
  column_to_rownames("seq_name") %>%
  pivot_longer(
    cols = -inj_trt_group,
    names_to = "microbe",
    values_to = "relab"
  ) %>%
  group_by(microbe, inj_trt_group) %>%
  summarise(
    mean = mean(relab, na.rm = TRUE),
    se   = sd(relab, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  mutate(
    panel = "GroupMeans",
    y_label = paste(microbe, inj_trt_group, sep = "_")
  )

head(group_abund)
```
Blast effect CI dataframe
```{r}
blastCI <- lmer_b$lmer_res %>% # lmer$a_main$lmer_res %>%
  mutate(y = gsub("s__", "", y)) %>%
  filter(
    term == "blast",
    y %in% bugs
    ) %>%
  mutate(microbe = fct_reorder(factor(y), estimate)) %>%
  transmute(
    microbe,
    # inj_trt_group = term,                     # for consistency in grouping/color
    mean = estimate,                            # match variable name
    se   = se,
    y_label = paste(microbe, term, sep = "_"),  # this is for if you have multiple terms
    panel = "Regression"                        # tag for facet column
  )

# save this bug order?
microbe_levels <- levels(blastCI$microbe)

head(blastCI)
```

```{r}
combined <- bind_rows(group_abund, blastCI) %>%
  mutate(
    microbe = factor(microbe, levels = microbe_levels),
    inj_trt_group = fct_rev(inj_trt_group)
    )


# Define alternating shading bands
shade_bands <- combined %>%
  group_by(microbe) %>% 
  arrange(by = microbe) %>%
  mutate(
    microbe = as.factor(microbe),
    band_id = as.numeric(microbe),
    fill = ifelse(band_id %% 2 == 1, "grey95", "white")
    )

head(combined)
head(shade_bands)
```

```{r}
plt <- ggplot(combined, aes(x = mean, y = inj_trt_group, color = inj_trt_group)) +
  geom_rect(
    data = shade_bands,
    aes(
      xmin = -Inf, xmax = Inf,
      ymin = -Inf, ymax = Inf,
      fill = fill
    ),
    inherit.aes = FALSE,
    color = NA
  ) +
  geom_point(size = 1.0) +
  geom_linerange(aes(xmin = mean - se, xmax = mean + se), linewidth = 0.25) +
  scale_fill_identity() +
  scale_color_manual(values = four_colors) +
  facet_grid(
    rows   = vars(microbe),
    cols   = vars(panel),
    space  = "free_y",
    scales = "free_x",
    switch = "y"
  ) +
  # scale_x_continuous(limits = c(-xlim, xlim)) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 14) +
  theme(
    strip.placement   = "outside",
    strip.text.y.left = element_text(angle = 0, hjust = 1),
    panel.spacing     = unit(0.1, "lines"),
    axis.text.y       = element_blank(),
    axis.ticks.y      = element_blank(),
    axis.ticks.x      = element_line(color = "black", linewidth = 0.4),
    axis.line.x       = element_line(colour = "black", linewidth = 0.4),
    panel.grid        = element_blank(),
    panel.background  = element_blank(),
    legend.position   = "none"
  )

plt
```
Fiddling with scaling parameters
```{r}
row_height <- 0.4  # inches per microbe row (adjust for your density)
n_microbes <- nlevels(combined$microbe)
height_in <- n_microbes * row_height

ggsave(
  "../figures/B_group_DA.svg",
  plot = plt,
  width = 6,           # fixed width (inches)
  height = height_in,  # scales with number of rows
  units = "in",
  device = "svg",
  dpi = 300
)

ggsave(
  "../figures/B_group_DA.png",
  plot = plt,
  width = 6,           # fixed width (inches)
  height = height_in,  # scales with number of rows
  units = "in",
  # device = "svg",
  dpi = 300
)
```

# Workspace
Code from `cytokine_plot.Rmd`
```{r}
cyto_plot <- ggplot(cyto_summary, aes(x = mean, y = inj_trt_group, color = inj_trt_group)) +
  geom_rect(
    data = cyto_bands,
    aes(
      xmin = -Inf, xmax = Inf,
      ymin = -Inf, ymax = Inf,
      fill = fill
    ),
    inherit.aes = FALSE,
    color = NA
  ) +
  geom_vline(xintercept =  0.0, color = "grey25", alpha = 0.7, linetype = "dashed", linewidth = 0.2) +
  geom_point(size = 1.0) +
  geom_linerange(aes(xmin = mean - se, xmax = mean + se), linewidth = 0.25) +
  scale_fill_identity() +
  # geom_errorbarh(aes(xmin = mean - se, xmax = mean + se), linewidth = 0.3, height = 0.5) +
  scale_color_manual(values = four_colors) +
  facet_grid(
    rows   = vars(variable),
    cols   = vars(location),
    # scales = "free_y",
    space  = "free_y",
    switch = "y"            # plot y-labels on left-hand side (instead of right)
    ) +
  scale_x_continuous(limits = c(-1.6, 1.6)) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 14) +
  theme(
    strip.placement   = "outside",
    strip.text.y.left = element_text(angle = 0, hjust = 1),
    panel.spacing.y   = unit(0.1, "lines"),   # try 1-3 "lines" of spacing
    panel.spacing.x   = unit(0.1, "lines"),   # put close together
    axis.text.y       = element_blank(),
    axis.ticks.y      = element_blank(),
    axis.line.x       = element_line(colour = "black", linewidth = 0.4),
    panel.grid        = element_blank(),
    panel.background  = element_blank(),
    legend.position   = "none"
  )

```

# Tried to make it better with Chat, but it got more complicated due to facet grid
```{r}
# Create numeric y positions
group_levels <- levels(group_abund$inj_trt_group)
n_groups <- length(group_levels)

combined <- bind_rows(group_abund, blastCI) %>%
  mutate(
    microbe = factor(microbe, levels = microbe_levels),
    inj_trt_group = fct_rev(inj_trt_group),
    y_pos = case_when(                  # numeric y position: 1, 2, 3, 4 for groups; midpoint for regression
      panel == "GroupMeans" ~ scales::rescale(as.numeric(inj_trt_group),
                                              to = c(1.2, n_groups - 0.2)),  # add top/bottom margins
      panel == "Regression" ~ (n_groups + 1) / 2,
      TRUE ~ NA_real_
    )
  )

# Shading bands (optional)
shade_bands <- combined %>%
  distinct(microbe) %>%
  mutate(
    band_id = as.numeric(microbe),
    fill = ifelse(band_id %% 2 == 1, "grey95", "white")
  )

# Plot
plt <- ggplot(combined, aes(x = mean, y = y_pos)) +
  geom_rect(
    data = shade_bands,
    aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf, fill = fill),
    inherit.aes = FALSE, color = NA
  ) +
  geom_linerange(
    aes(
      xmin = mean - se, xmax = mean + se,
      color = if_else(panel == "GroupMeans", inj_trt_group, NA_character_)
    ),
    linewidth = 0.25
  ) +
  geom_point(
    aes(color = if_else(panel == "GroupMeans", inj_trt_group, NA_character_)),
    size = 1
  ) +
  geom_point(
    data = combined %>% filter(panel == "Regression"),
    color = "black", size = 1
  ) +
  geom_linerange(
    data = combined %>% filter(panel == "Regression"),
    aes(xmin = mean - se, xmax = mean + se),
    color = "grey40", linewidth = 0.25
  ) +
  scale_fill_identity() +
  scale_color_manual(values = four_colors, na.value = "black") +
  facet_grid(
    rows   = vars(microbe),
    cols   = vars(panel),
    space  = "free_y",
    scales = "free_x",
    switch = "y"
  ) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 14) +
  theme(
    strip.placement   = "outside",
    strip.text.y.left = element_text(angle = 0, hjust = 1),
    panel.spacing.y   = unit(0.4, "lines"),
    panel.spacing.x   = unit(0.2, "lines"),
    axis.text.y       = element_blank(),
    axis.ticks.y      = element_blank(),
    axis.line.x       = element_line(colour = "black", linewidth = 0.4),
    panel.grid        = element_blank(),
    panel.background  = element_blank(),
    legend.position   = "none"
  )

plt
```

