---
title: "VNS_tables.Rmd"
author: "Alyssa N. Easton"
date: '2025-04-07'
output:
---

Read in dataframes I saved as RDS, convert to excel spreadsheet format

## Setup
```{r "setup", include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
knitr::opts_knit$set(root.dir = file.path(Sys.getenv("VNS_ROOT"), "data"))  # stored VNS_ROOT in ~/.Renviron
options(expressions = 6000)  # Increase allowed recursion depth
options(digits = 4) ## Formats output to 10 digits
library(magrittr)
library(tidyverse)
library(broom)
library(broom.helpers)
library(rlang)
library(glue)
library(janitor)
library(purrr)
library(ggeffects)
library(patchwork)
library(car)
library(scales)
library(future.apply)
library(ComplexHeatmap)
library(circlize)
library(cowplot)
library(Cairo)
library(grid)
library(gridExtra)
library(ggtext)
library(openxlsx)

# source("/proj/gibbons/2024_easton_vns/workflow/functions/mdact.R")
# source("/proj/gibbons/2024_easton_vns/workflow/functions/june_functions.R")
# source("/proj/gibbons/2024_easton_vns/workflow/functions/shannon_div_functions.R")
# source("/proj/gibbons/2024_easton_vns/workflow/functions/artifact_correct_functions.R")
# source("/proj/gibbons/2024_easton_vns/workflow/functions/lmer_functions.R")
source("/proj/gibbons/2024_easton_vns/workflow/functions/final_form_functions.R")
all_data <- readRDS("/proj/gibbons/2024_easton_vns/data/1_preprocessed/species_october.rds")
taxmap <- all_data$taxonomy_map %>%
  transmute(Species, Genus, Family, Order, Class, Phylum, Kingdom, full_name)
```

# Start with part A
```{r}
model_microbe <- readRDS("../data/3_newest/A_microbe_models_0212.rds") # LRT p-values
model_outcome <- readRDS("../data/3_newest/A_outcome_models_0212.rds")  # LRT p-values (should be equivalent to coef_df)
coef_df <- readRDS("../data/3_newest/A_microbe_coef_dfs_0218.rds")   # full lm() model specifications for M and Y
all_run_df <- readRDS("../data/3_newest/A_mediation_0212.rds")
ci_df <- readRDS("../data/3_newest/A_bootCI_df_0212.rds")

coef_microbe <- readRDS("../data/3_newest/A_microbe_coef_pvals_0206.rds") # regular coefficient p-values
coef_outcome <- readRDS("../data/3_newest/A_outcome_coef_pvals_0206.rds") # regular coef p-vals
```


```{r}
colnames(model_microbe) 
colnames(coef_microbe)
```

```{r}
# mod_M <- model_microbe %>%
#   transmute(term, estimate, "SE" = std.error, "F" = statistic, "Pr(>F)" = p.value, microbe, "n_samples" = n, "Adjusted p" = adj.p)

ttest_m <- coef_microbe %>%
  transmute(microbe, term, "est"= estimate, "se" = std.error, "t" = statistic, "t_p" = p.value, "t_q" = adj.p) %>%
  mutate(term = case_when(
    term == "blast_and_vns" ~ "interact",
    term == "(Intercept)" ~ "intercept",
    term == "Round_main3" ~ "batch",
    .default = term
    )) %>% 
  pivot_wider(
    id_cols = microbe,
    names_from = term,
    values_from = c("est", "se", "t", "t_p", "t_q"),
    names_glue =  "{.value}_{term}"
  ) %>% 
  dplyr::select(-t_q_intercept)

all_m <- model_microbe %>%
  transmute(microbe, term, "LRT_F" = statistic, "LRT_p" = p.value, n, "LRT_q" = adj.p) %>%
  mutate(term = ifelse(term == "blast_and_vns", "interact", term)) %>% 
  pivot_wider(
    id_cols = c(microbe, n),
    names_from = term,
    values_from = c("LRT_F", "LRT_p", "LRT_q"),
    names_glue =  "{.value}_{term}"
  ) %>%
  left_join(ttest_m, by = "microbe") %>%
  rename(Species = microbe) %>%
  arrange(Species) %>%
  left_join(taxmap, by = "Species") %>%
  mutate(Species = case_when(
    Species == "berger_parker" ~ "Berger Parker Index",
    Species == "richness" ~ "Richness (# observed features)",
    Species == "shannon" ~ "Shannon Index",
    .default = Species
  ))

all_m
```

one row for each model M regression.
n=27 for all of these. I can just write that somewhere. 
F(blast) and F(VNS) are df=2, and F(interact) is df=1

```{r}
ttest_y <- coef_outcome %>%
  map_dfr(function(one) {
    one %>%
      mutate(outcome = unique(one$outcome))
  }) %>%
  transmute(
    outcome, microbe, term, "est" = estimate, "se" = std.error,
    "t" = statistic, "t_p" = p.value, "t_q" = adj.p
    ) %>%
  mutate(
    term = case_when(
    term == "blast_and_vns" ~ "interact",
    term == "(Intercept)" ~ "intercept",
    term == "Round_main3" ~ "batch",
    .default = term
      )) %>% 
  pivot_wider(
    id_cols = c("outcome", "microbe"),
    names_from = term,
    values_from = c("est", "se", "t", "t_p", "t_q"),
    names_glue =  "{.value}_{term}"
  ) %>%
  dplyr::select(-t_q_intercept)

ttest_y
```

```{r}
all_y <- model_outcome %>%
  map_dfr(function(one) {
    one %>%
      mutate(outcome = unique(one$outcome))
  }) %>%
  transmute(outcome, microbe, term, "LRT_F" = statistic, "LRT_p" = p.value, n, "LRT_q" = adj.p) %>%
  filter(term == "microbe") %>%
  pivot_wider(
    id_cols = c(outcome, microbe, n),
    names_from = term,
    values_from = c("LRT_F", "LRT_p", "LRT_q"),
    names_glue =  "{.value}_{term}"
  ) %>%
  left_join(ttest_y, by = c("outcome", "microbe")) %>%
  rename(Species = microbe) %>%
  arrange(outcome, Species) %>%
  left_join(taxmap, by = "Species") %>%
  mutate(Species = case_when(
    Species == "berger_parker" ~ "Berger Parker Index",
    Species == "richness" ~ "Richness (# observed features)",
    Species == "shannon" ~ "Shannon Index",
    .default = Species
  ))

all_y %>% transmute(outcome, Species, LRT_p_microbe, t_p_microbe) # LRT p-value is equivalent to coefficient p-value here.
```

Now onto M-DACT results
```{r}
mdact <- all_run_df %>%
  transmute(
    exposure, mediator, outcome, p_m, p_y, p_composite, dact_stat, dact_thr_10, dact_thr_15, dact_thr_20, signif_at
  ) %>%
  mutate(
    exposure = ifelse(exposure == "pm_blast", "blast", "vns"),
    Species = mediator
    ) %>%
  arrange(outcome, mediator, exposure) %>%
  left_join(taxmap, by = "Species") %>%
  mutate(mediator = case_when(
    mediator == "berger_parker" ~ "Berger Parker Index",
    mediator == "richness" ~ "Richness (# observed features)",
    mediator == "shannon" ~ "Shannon Index",
    .default = mediator
  ))

# mdact %>% filter(signif_at == 0.250)

mdact
```
```{r}
mdact %>% filter(mediator == "Shannon Index")
```

Finally, the CI df
```{r}
acme <- ci_df %>%
  transmute(
    exposure, mediator, outcome,
    ACME_cond0_lower, ACME_cond0_est, ACME_cond0_upper,
    ADE_cond0_lower, ADE_cond0_est, ADE_cond0_upper,
    Total_cond0_lower, Total_cond0_est, Total_cond0_upper,
    ACME_cond1_lower, ACME_cond1_est, ACME_cond1_upper,
    ADE_cond1_lower, ADE_cond1_est, ADE_cond1_upper,
    Total_cond1_lower, Total_cond1_est, Total_cond1_upper
  )

acme
```

```{r}
ci_blast <- acme %>% filter(exposure == "blast") %>%
  mutate(
    `ACME (VNS-)` = paste0("[", round(ACME_cond0_lower, 3), ", ", round(ACME_cond0_upper, 3), "]"),
    `ADE (VNS-)` = paste0("[", round(ADE_cond0_lower, 3), ", ", round(ADE_cond0_upper, 3), "]"),
    `Total (VNS-)`= paste0("[", round(Total_cond0_lower, 3), ", ", round(Total_cond0_upper, 3), "]"),
    `ACME (VNS+)` = paste0("[", round(ACME_cond1_lower, 3), ", ", round(ACME_cond1_upper, 3), "]"),
    `ADE (VNS+)` = paste0("[", round(ADE_cond1_lower, 3), ", ", round(ADE_cond1_upper, 3), "]"),
    `Total (VNS+)`= paste0("[", round(Total_cond1_lower, 3), ", ", round(Total_cond1_upper, 3), "]")
  ) %>%
  transmute(exposure, mediator, outcome,
            `ACME (VNS-)`, `ADE (VNS-)`, `Total (VNS-)`,
            `ACME (VNS+)`, `ADE (VNS+)`, `Total (VNS+)`
  )

ci_vns <- acme %>% filter(exposure == "vns") %>%
  mutate(
    `ACME (Blast-)` = paste0("[", round(ACME_cond0_lower, 3), ", ", round(ACME_cond0_upper, 3), "]"),
    `ADE (Blast-)` = paste0("[", round(ADE_cond0_lower, 3), ", ", round(ADE_cond0_upper, 3), "]"),
    `Total (Blast-)`= paste0("[", round(Total_cond0_lower, 3), ", ", round(Total_cond0_upper, 3), "]"),
    `ACME (Blast+)` = paste0("[", round(ACME_cond1_lower, 3), ", ", round(ACME_cond1_upper, 3), "]"),
    `ADE (Blast+)` = paste0("[", round(ADE_cond1_lower, 3), ", ", round(ADE_cond1_upper, 3), "]"),
    `Total (Blast+)`= paste0("[", round(Total_cond1_lower, 3), ", ", round(Total_cond1_upper, 3), "]")
  ) %>%
  transmute(exposure, mediator, outcome,
            `ACME (Blast-)`, `ADE (Blast-)`, `Total (Blast-)`,
            `ACME (Blast+)`, `ADE (Blast+)`, `Total (Blast+)`
  )


ci_blast
ci_vns
```

```{r}
a_tables <- list(
  "lm_m" = all_m,
  "lm_y" = all_y,
  "mdact_mediation" = mdact,
  "acme_estimation" = acme, 
  "blast_ci" = ci_blast,
  "vns_ci" = ci_vns
)
```


# Repeat for cohort B things
```{r}
model_microbe <- readRDS("../data/3_newest/B_microbe_models_0216.rds") # LRT p-values
model_outcome <- readRDS("../data/3_newest/B_outcome_models_0216.rds")  # LRT p-values (should be equivalent to coef_df)
coef_df <- readRDS("../data/3_newest/B_microbe_coef_dfs_0217.rds")   # full lm() model specifications for M and Y
all_run_df <- readRDS("../data/3_newest/B_mediation_0216.rds")
ci_df <- readRDS("../data/3_newest/B_bootCI_df_0216.rds")

coef_microbe <- readRDS("../data/3_newest/B_microbe_coef_pvals_0216.rds") # regular coefficient p-values
coef_outcome <- readRDS("../data/3_newest/B_outcome_coefs_0216.rds") # regular coef p-vals
```


```{r}
colnames(model_microbe) 
colnames(coef_microbe)

model_microbe
coef_microbe
```

```{r}
# mod_M <- model_microbe %>%
#   transmute(term, estimate, "SE" = std.error, "F" = statistic, "Pr(>F)" = p.value, microbe, "n_samples" = n, "Adjusted p" = adj.p)

ttest_m <- coef_microbe %>%
  transmute(microbe, term, "est"= estimate, "se" = std.error, "t" = statistic, "t_p" = p.value, "t_q" = adj.p) %>%
  mutate(term = case_when(
    term == "blast_and_vns" ~ "interact",
    term == "(Intercept)" ~ "intercept",
    term == "Round_main3" ~ "batch",     # there is no main term for batch here because it was a random effect
    .default = term
    )) %>% 
  pivot_wider(
    id_cols = microbe,
    names_from = term,
    values_from = c("est", "se", "t", "t_p", "t_q"),
    names_glue =  "{.value}_{term}"
  ) %>% 
  dplyr::select(-t_q_intercept)

all_m <- model_microbe %>%
  transmute(microbe, term, "LRT_F" = statistic, "LRT_p" = p.value, n, "LRT_q" = adj.p) %>%
  mutate(term = ifelse(term == "blast_and_vns", "interact", term)) %>% 
  pivot_wider(
    id_cols = c(microbe, n),
    names_from = term,
    values_from = c("LRT_F", "LRT_p", "LRT_q"),
    names_glue =  "{.value}_{term}"
  ) %>%
  left_join(ttest_m, by = "microbe") %>%
  rename(Species = microbe) %>%
  arrange(Species) %>%
  left_join(taxmap, by = "Species") %>%
  mutate(Species = case_when(
    Species == "berger_parker" ~ "Berger Parker Index",
    Species == "richness" ~ "Richness (# observed features)",
    Species == "shannon" ~ "Shannon Index",
    .default = Species
  ))

all_m
```


```{r}
ttest_y <- coef_outcome %>%
  map_dfr(function(one) {
    one %>%
      mutate(outcome = unique(one$outcome))
  }) %>%
  transmute(
    outcome, microbe, term, "est" = estimate, "se" = std.error,
    "t" = statistic, "t_p" = p.value, "t_q" = adj.p
    ) %>%
  mutate(
    term = case_when(
    term == "blast_and_vns" ~ "interact",
    term == "(Intercept)" ~ "intercept",      # I don't seem to have intercept coefficients saved here
    term == "Round_main3" ~ "batch",
    .default = term
      )) %>% 
  pivot_wider(
    id_cols = c("outcome", "microbe"),
    names_from = term,
    values_from = c("est", "se", "t", "t_p", "t_q"),
    names_glue =  "{.value}_{term}"
  )

ttest_y
```

```{r}
all_y <- model_outcome %>%
  map_dfr(function(one) {
    one %>%
      mutate(outcome = unique(one$outcome))
  }) %>%
  transmute(outcome, microbe, term, "LRT_Chi" = statistic, "LRT_p" = p.value, n, "LRT_q" = adj.p) %>%
  filter(term == "microbe") %>%
  pivot_wider(
    id_cols = c(outcome, microbe, n),
    names_from = term,
    values_from = c("LRT_Chi", "LRT_p", "LRT_q"),
    names_glue =  "{.value}_{term}"
  ) %>%
  left_join(ttest_y, by = c("outcome", "microbe")) %>%
  rename(Species = microbe) %>%
  arrange(outcome, Species) %>%
  left_join(taxmap, by = "Species") %>%
  mutate(Species = case_when(
    Species == "berger_parker" ~ "Berger Parker Index",
    Species == "richness" ~ "Richness (# observed features)",
    Species == "shannon" ~ "Shannon Index",
    .default = Species
  ))

all_y %>% transmute(outcome, Species, LRT_p_microbe, t_p_microbe) # LRT p-value is equivalent to coefficient p-value here.
```

The LRT p-values are just slightly different from the lmer p-values, maybe because they're Swatterwaithe vs. Chi (asymptotically equivalent), as opposed to Student's t-test vs. F, which are the same? But they're close enough that it really doesn't matter. 

Now onto M-DACT results
```{r}
mdact <- all_run_df %>%
  transmute(
    exposure, mediator, outcome, p_m, p_y, p_composite, dact_stat, dact_thr_10, dact_thr_15, dact_thr_20, signif_at
  ) %>%
  mutate(
    exposure = ifelse(exposure == "pm_blast", "blast", "vns"),
    Species = mediator
    ) %>%
  arrange(outcome, mediator, exposure) %>%
  left_join(taxmap, by = "Species") %>%
  mutate(mediator = case_when(
    mediator == "berger_parker" ~ "Berger Parker Index",
    mediator == "richness" ~ "Richness (# observed features)",
    mediator == "shannon" ~ "Shannon Index",
    .default = mediator
  ))

# mdact %>% filter(signif_at == 0.250)

mdact
```


```{r}
mdact %>% filter(!is.na(signif_at))
```

Finally, the CI df
```{r}
acme <- ci_df %>%
  transmute(
    exposure, mediator, outcome,
    ACME_cond0_lower, ACME_cond0_est, ACME_cond0_upper,
    ADE_cond0_lower, ADE_cond0_est, ADE_cond0_upper,
    Total_cond0_lower, Total_cond0_est, Total_cond0_upper,
    ACME_cond1_lower, ACME_cond1_est, ACME_cond1_upper,
    ADE_cond1_lower, ADE_cond1_est, ADE_cond1_upper,
    Total_cond1_lower, Total_cond1_est, Total_cond1_upper
  ) %>%
  arrange(outcome, mediator, exposure) %>%
  mutate(Species = mediator) %>% left_join(taxmap, by = "Species") %>%
  mutate(mediator = case_when(
    mediator == "berger_parker" ~ "Berger Parker Index",
    mediator == "richness" ~ "Richness (# observed features)",
    mediator == "shannon" ~ "Shannon Index",
    .default = mediator
  )) %>%
  dplyr::select(-Species)

ci_blast <- acme %>% filter(exposure == "blast") %>%
  mutate(
    `ACME (VNS-)` = paste0("[", round(ACME_cond0_lower, 3), ", ", round(ACME_cond0_upper, 3), "]"),
    `ADE (VNS-)` = paste0("[", round(ADE_cond0_lower, 3), ", ", round(ADE_cond0_upper, 3), "]"),
    `Total (VNS-)`= paste0("[", round(Total_cond0_lower, 3), ", ", round(Total_cond0_upper, 3), "]"),
    `ACME (VNS+)` = paste0("[", round(ACME_cond1_lower, 3), ", ", round(ACME_cond1_upper, 3), "]"),
    `ADE (VNS+)` = paste0("[", round(ADE_cond1_lower, 3), ", ", round(ADE_cond1_upper, 3), "]"),
    `Total (VNS+)`= paste0("[", round(Total_cond1_lower, 3), ", ", round(Total_cond1_upper, 3), "]")
  ) %>%
  transmute(
    exposure, mediator, outcome,
    `ACME (VNS-)`, `ADE (VNS-)`, `Total (VNS-)`,
    `ACME (VNS+)`, `ADE (VNS+)`, `Total (VNS+)`,
    Genus, Family, Order, Class, Phylum, Kingdom, full_name
  )

ci_vns <- acme %>% filter(exposure == "vns") %>%
  mutate(
    `ACME (Blast-)` = paste0("[", round(ACME_cond0_lower, 3), ", ", round(ACME_cond0_upper, 3), "]"),
    `ADE (Blast-)` = paste0("[", round(ADE_cond0_lower, 3), ", ", round(ADE_cond0_upper, 3), "]"),
    `Total (Blast-)`= paste0("[", round(Total_cond0_lower, 3), ", ", round(Total_cond0_upper, 3), "]"),
    `ACME (Blast+)` = paste0("[", round(ACME_cond1_lower, 3), ", ", round(ACME_cond1_upper, 3), "]"),
    `ADE (Blast+)` = paste0("[", round(ADE_cond1_lower, 3), ", ", round(ADE_cond1_upper, 3), "]"),
    `Total (Blast+)`= paste0("[", round(Total_cond1_lower, 3), ", ", round(Total_cond1_upper, 3), "]")
  ) %>%
  transmute(
    exposure, mediator, outcome,
    `ACME (Blast-)`, `ADE (Blast-)`, `Total (Blast-)`,
    `ACME (Blast+)`, `ADE (Blast+)`, `Total (Blast+)`,
    Genus, Family, Order, Class, Phylum, Kingdom, full_name
  )


ci_blast
ci_vns
```


```{r}
b_tables <- list(
  "lmer_m" = all_m,
  "lmer_y" = all_y,
  "mdact_mediation" = mdact,
  "acme_estimation" = acme, 
  "blast_ci" = ci_blast,
  "vns_ci" = ci_vns
)
```


# Save as sheets in the an excel workbook
```{r}
# save df as 2 separate sheets
wb <- createWorkbook()
addWorksheet(wb, "ANOVA")
writeData(wb, "ANOVA", anova_df)
addWorksheet(wb, "posthoc")
writeData(wb, "posthoc", posthoc_df)
saveWorkbook(wb, file = "final/cytokine_PC_stats.xlsx", overwrite = TRUE)
```

```{r}
# wb <- createWorkbook()
# addWorksheet(wb, "acute_microbe_model")
# addWorksheet(wb, "acute_outcome_model")
# addWorksheet(wb, "acute_MDACT_mediation")
# addWorksheet(wb, "acute_CI_raw")
# addWorksheet(wb, "acute_CI_formatted")
# 
# addWorksheet(wb, "chronic_microbe_model")
# addWorksheet(wb, "chronic_outcome_model")
# addWorksheet(wb, "chronic_MDACT_mediation")
# addWorksheet(wb, "chronic_CI_raw")
# addWorksheet(wb, "chronic_CI_formatted")
# 
# writeData(wb, "acute_microbe_model", a_tables$lm_m)
# writeData(wb, "acute_outcome_model", a_tables$lm_y)
# writeData(wb, "acute_MDACT_mediation", a_tables$mdact_mediation)
# writeData(wb, "acute_CI_raw", a_tables$acme_estimation)
# writeData(wb, "acute_CI_formatted", a_tables$blast_ci, startRow = 1, colNames = TRUE)
# writeData(wb, "acute_CI_formatted", a_tables$vns_ci, startRow = nrow(a_tables$blast_ci) + 4, colNames = TRUE)
# 
# writeData(wb, "chronic_microbe_model", b_tables$lmer_m)
# writeData(wb, "chronic_outcome_model", b_tables$lmer_y)
# writeData(wb, "chronic_MDACT_mediation", b_tables$mdact_mediation)
# writeData(wb, "chronic_CI_raw", b_tables$acme_estimation)
# writeData(wb, "chronic_CI_formatted", b_tables$blast_ci, startRow = 1, colNames = TRUE)
# writeData(wb, "chronic_CI_formatted", b_tables$vns_ci, startRow = nrow(b_tables$blast_ci) + 4, colNames = TRUE)
# 
# saveWorkbook(wb, file = "final/lm_lmer_tables.xlsx", overwrite = TRUE)
```

```{r}
wb <- createWorkbook()

# Create style
header_style <- createStyle(textDecoration = c("bold", "underline"))

# Add sheets
addWorksheet(wb, "acute_microbe_model")
addWorksheet(wb, "acute_outcome_model")
addWorksheet(wb, "acute_MDACT_mediation")
addWorksheet(wb, "acute_CI_raw")
addWorksheet(wb, "acute_CI_formatted")

addWorksheet(wb, "chronic_microbe_model")
addWorksheet(wb, "chronic_outcome_model")
addWorksheet(wb, "chronic_MDACT_mediation")
addWorksheet(wb, "chronic_CI_raw")
addWorksheet(wb, "chronic_CI_formatted")

# ---- Acute ----
writeData(wb, "acute_microbe_model", a_tables$lm_m, headerStyle = header_style)
setColWidths(wb, "acute_microbe_model", cols = 1:ncol(a_tables$lm_m), widths = "auto")

writeData(wb, "acute_outcome_model", a_tables$lm_y, headerStyle = header_style)
setColWidths(wb, "acute_outcome_model", cols = 1:ncol(a_tables$lm_y), widths = "auto")

writeData(wb, "acute_MDACT_mediation", a_tables$mdact_mediation, headerStyle = header_style)
setColWidths(wb, "acute_MDACT_mediation", cols = 1:ncol(a_tables$mdact_mediation), widths = "auto")

writeData(wb, "acute_CI_raw", a_tables$acme_estimation, headerStyle = header_style)
setColWidths(wb, "acute_CI_raw", cols = 1:ncol(a_tables$acme_estimation), widths = "auto")

# stacked tables
writeData(wb, "acute_CI_formatted", a_tables$blast_ci,
          startRow = 1, headerStyle = header_style)

writeData(wb, "acute_CI_formatted", a_tables$vns_ci,
          startRow = nrow(a_tables$blast_ci) + 4,
          headerStyle = header_style)

setColWidths(wb, "acute_CI_formatted",
             cols = 1:max(ncol(a_tables$blast_ci), ncol(a_tables$vns_ci)),
             widths = "auto")

# ---- Chronic ----
writeData(wb, "chronic_microbe_model", b_tables$lmer_m, headerStyle = header_style)
setColWidths(wb, "chronic_microbe_model", cols = 1:ncol(b_tables$lmer_m), widths = "auto")

writeData(wb, "chronic_outcome_model", b_tables$lmer_y, headerStyle = header_style)
setColWidths(wb, "chronic_outcome_model", cols = 1:ncol(b_tables$lmer_y), widths = "auto")

writeData(wb, "chronic_MDACT_mediation", b_tables$mdact_mediation, headerStyle = header_style)
setColWidths(wb, "chronic_MDACT_mediation", cols = 1:ncol(b_tables$mdact_mediation), widths = "auto")

writeData(wb, "chronic_CI_raw", b_tables$acme_estimation, headerStyle = header_style)
setColWidths(wb, "chronic_CI_raw", cols = 1:ncol(b_tables$acme_estimation), widths = "auto")

# stacked tables
writeData(wb, "chronic_CI_formatted", b_tables$blast_ci,
          startRow = 1, headerStyle = header_style)

writeData(wb, "chronic_CI_formatted", b_tables$vns_ci,
          startRow = nrow(b_tables$blast_ci) + 4,
          headerStyle = header_style)

setColWidths(wb, "chronic_CI_formatted",
             cols = 1:max(ncol(b_tables$blast_ci), ncol(b_tables$vns_ci)),
             widths = "auto")

# Save
saveWorkbook(wb, "../data/final/lm_lmer_tables.xlsx", overwrite = TRUE)
```

