---
title: "XX_A_regress_mediate.Rmd"
author: "Alyssa N. Easton"
date: '2026-07-07'
output:
---

Testing the hypothesis:
“Microbe mediates ANY component of the blast → outcome relationship”

## Setup
```{r "setup", include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
knitr::opts_knit$set(root.dir = file.path(Sys.getenv("VNS_ROOT"), "data"))  # stored VNS_ROOT in ~/.Renviron
options(expressions = 6000)  # Increase allowed recursion depth
options(digits = 4) ## Formats output to 10 digits
library(magrittr)
library(tidyverse)
library(broom)
library(broom.helpers)
library(rlang)
library(glue)
library(janitor)
library(purrr)
library(ggh4x)
library(zCompositions)
library(parallel)
library(parallelly)
library(pbapply)
library(furrr)      # future_pmap comes from here
library(future)
library(limma)
library(ggeffects)
library(patchwork)
library(lme4)
library(lmerTest)
library(performance)
library(MuMIn)
library(car)
library(scales)
library(future.apply)
library(ComplexHeatmap)
library(circlize)
library(cowplot)
library(Cairo)
source("./functions/mdact.R")
source("./functions/june_functions.R")
source("./functions/shannon_div_functions.R")
source("./functions/artifact_correct_functions.R")
source("./functions/lmer_functions.R")
source("./functions/mediation_utils.R")   # this overrides a lot of mdact.R
```

## Import and preprocess data
    metadata, microbe presence/absence, sample-wise presence/absence, batch-wise presence/absence
    compile into a summary `bug_info`
```{r}
all_data <- readRDS("/proj/gibbons/2024_easton_vns/data/1_preprocessed/species_october.rds")
alpha_a <- readRDS("/proj/gibbons/2024_easton_vns/data/diversity/A_alpha.rds")
bug_info <- readRDS("../workflow/A_bug_info.rds")

presence_df <- all_data$a_counts %>%
  mutate(across(everything(), ~ . != 0)) %>%
  rownames_to_column("seq_name")

per_sample_presence <- all_data$a_counts %>%
  mutate(across(everything(), ~ . != 0)) %>%
  {rowSums(.) / ncol(.)}

# all_joint_regr <- readRDS("../B_pM_lmer_0106.rds")
# out_list <- readRDS("../B_outcome_lmer_0106.rds")

outcome_cols <- c(
  "acute_cyto_PC1",  "acute_cyto_PC2", "acute_cyto_PC3",
  "acute_cyto_PC4",   "acute_cyto_PC5"
  )
```

# 1: Statistical models

### Processing
```{r}
args <- list(
  total_presence      = 0.3,
  batch_presence      = 0.1,
  impute_method       = "GBM",
  filt_before_clr     = TRUE,
  ctrl_batch          = TRUE
  )

filt_bugs <- bug_info %>% filter(
  pres_total     >= args$total_presence,
  pres_batch1    >= args$batch_presence,
  pres_batch3    >= args$batch_presence) %>%
  pull(microbe)

if (args$filt_before_clr) {
  filt_counts <- all_data$a_counts %>%
    dplyr::select(any_of(filt_bugs))
} else {
  filt_counts <- all_data$a_counts
}

imputed_counts <- cmultRepl(filt_counts, label = 0, method = args$impute_method, z.delete = FALSE)
gm <- exp(rowMeans(log(imputed_counts)))
clr_mat <- log(imputed_counts) - log(gm)
clr_df <- as_tibble(clr_mat) %>% scale() %>% as.data.frame() # NOW i've scaled it
colnames(clr_df) <- colnames(filt_counts)
rownames(clr_df) <- rownames(clr_mat)


adiv_features <- alpha_a %>%
  transmute(shannon, richness, berger_parker) %>%
  rownames_to_column("seq_name") %>%
  mutate(
    shannon = scale(shannon),
    richness = scale(richness),
    berger_parker = scale(log(berger_parker))
    )

hist(alpha_a$berger_parker)
hist(adiv_features$berger_parker)
```

Merge all data into one df 
  Treatments (exposures) are binary
  Microbes are filtered, CLR-transformed, and then z-scored
  Outcome variables are z-scored
  *Alpha diversity is z-scored!*
  
```{r}
df <- readRDS("/proj/gibbons/2024_easton_vns/data/1_preprocessed/meta_outcomePC_all.rds") %>%
  dplyr::select(all_of(c(
    "seq_name", "inj_trt_group", "blast", "vns", "blast_and_vns", "Round_main", outcome_cols
  ))) %>%
  mutate(seq_name = as.character(seq_name)) %>%
  right_join((clr_df %>% rownames_to_column("seq_name")), by = "seq_name") %>%
  left_join(adiv_features, by = "seq_name")

filt_bugs <- c(filt_bugs, "shannon", "richness", "berger_parker")

# saveRDS(df, "../data/processed/A_clr_df_withalpha_0206.rds")
# df <- readRDS("../data/processed/A_clr_df_withalpha_0206.rds")
```


## Model microbe (M)

Differential abundance testing (contains p_m)
  > Added in testing for VNS effect given blast == 1
  
```{r}
model_microbe <- map_dfr(filt_bugs, function(bug_name) {
  
  form <- as.formula(paste0(bug_name, " ~ blast + vns + blast_and_vns + Round_main"))
  terms <- c("blast", "vns", "blast_and_vns") #  "vns_on_blast")
  n_samples <- sum(!is.na(df[[bug_name]]))

  tryCatch({

    m <- lm(form, data = df, na.action = "na.exclude")
    sm <- summary(m)$coefficients
    
    ################## Fixed effects ##################
    fe <- tibble(
      term      = rownames(sm),
      estimate  = sm[, "Estimate"],
      std.error = sm[, "Std. Error"],
      statistic = sm[, "t value"]
    ) %>%
      filter(term %in% c("blast", "vns", "blast_and_vns"))  # not sure if we want blast_and_vns
    
    ################## P-values ##################
    # Exact F test is available for lm(), so I will use that instead of asymptotic Chisq
    pvals <- tibble(
      term = terms,
      p.value = c(
        anova(update(m, . ~ . - blast - blast_and_vns),m, test = "F")$`Pr(>F)`[2],  # any blast effect
        anova(update(m, . ~ . - vns   - blast_and_vns), m, test = "F")$`Pr(>F)`[2], # any vns effect
        anova(update(m, . ~ .         - blast_and_vns), m, test = "F")$`Pr(>F)`[2] # interaction
        # car::linearHypothesis(m, "vns + blast_and_vns = 0")$`Pr(>F)`[2]  # ⭐ VNS effect when blast = 1 (linear contrast)
      )
    )

    ################## Combine ################## # removed linear hypothesis from output
    # bind_rows(
    #   fe %>% dplyr::select(term, estimate, std.error, statistic),
    #   tibble(
    #     term = "vns_on_blast",
    #     estimate = NA_real_,
    #     std.error = NA_real_,
    #     statistic = NA_real_
    #   )) %>%
    bind_rows(
      fe %>%
        dplyr::select(term, estimate, std.error, statistic) %>%
        left_join(pvals, by = "term") %>%
        mutate(
          microbe = bug_name,
          n = n_samples
          )
    )

  }, error = function(e) {

    tibble(
      term      = terms,
      estimate  = NA_real_,
      std.error = NA_real_,
      statistic = NA_real_,
      p.value   = NA_real_,
      microbe   = bug_name,
      n         = n_samples
    )
  })
}) %>%
  group_by(term) %>%
  mutate(adj.p = p.adjust(p.value, method = "BH")) %>%
  ungroup()

# saveRDS(model_microbe, "../A_microbe_models_0109.rds")
# saveRDS(model_microbe, "../A_microbe_models_0113.rds")
saveRDS(model_microbe, "../data/3_newest/A_microbe_models_0212.rds")
head(model_microbe)
```

Separately, run to get regular ("Wald-style" t-test) coefficient p-values (for heat maps)
```{r}
# testlm <- lm(
#   Lachnospiraceae_unclassified_SGB41414 ~ blast + vns + blast_and_vns + Round_main,
#   data = df
#   )
# 
# summary(testlm)$coefficients

coef_microbe <- map_dfr(filt_bugs, function(bug_name) {
  
  form <- as.formula(paste0(bug_name, " ~ blast + vns + blast_and_vns + Round_main"))
  terms <- c("blast", "vns", "blast_and_vns", "Round_main3", "(Intercept)")
  n_samples <- sum(!is.na(df[[bug_name]]))

  tryCatch({

    m <- lm(form, data = df, na.action = "na.exclude")
    sm <- summary(m)$coefficients
    
    ################## Fixed effects ##################
    fe <- tibble(
      term      = rownames(sm),
      estimate  = sm[, "Estimate"],
      std.error = sm[, "Std. Error"],
      statistic = sm[, "t value"],
      p.value   = sm[, "Pr(>|t|)"]
    ) %>%
      filter(term %in% terms)
  
    bind_rows(fe %>% mutate(microbe = bug_name, n = n_samples))
    
  }, error = function(e) {

    tibble(
      term      = terms,
      estimate  = NA_real_,
      std.error = NA_real_,
      statistic = NA_real_,
      p.value   = NA_real_,
      microbe   = bug_name,
      n         = n_samples
    )
  })
}) %>%
  group_by(term) %>%
  mutate(adj.p = ifelse(
    term == "(Intercept)", NA_real_,
    p.adjust(p.value, method = "BH")
    )) %>%
  ungroup()

# saveRDS(coef_microbe, "../data/3_newest/A_microbe_coef_pvals_0206.rds")

head(coef_microbe)
```

## Model outcome (Y)

Test whether microbes explain any variance in outcome not explained by treatment
```{r}
model_outcome <- outcome_cols %>% # used two_col to test
  set_names() %>%
  map(function(out) {
    map_dfr(filt_bugs, function(bug_name) {
      df$microbe <- df[[bug_name]]
      # df <- df %>% filter(!is.na(out))      # added this to remove NA outcome samples
      df <- df %>% filter(!is.na(.data[[out]])) # changed to this because out is a column name
      form <- as.formula(paste0(out, " ~ microbe + blast + vns + blast_and_vns + Round_main"))
      terms <- c("blast", "vns", "blast_and_vns", "microbe") # ,"vns_on_blast")
      n_samples <- sum(!is.na(df[[bug_name]])) # count non-missing samples for this bug

      tryCatch({
        m <- lm(form, data = df, na.action = "na.exclude")
        sm <- summary(m)$coefficients   ## ---- fixed-effect estimates ----
        fe <- tibble(
          term      = rownames(sm),
          estimate  = sm[, "Estimate"],
          std.error = sm[, "Std. Error"],
          statistic = sm[, "t value"]
        ) %>%
          filter(term %in% terms)
        
        ################## P-values ##################
        pvals <- tibble(                ## ---- LRT p-values ----
          term = terms,
          p.value = c(      # I am using an exact F-test instead of Chi-squared due to small sample size
            anova(update(m, . ~ . - blast - blast_and_vns), m, test = "F")$`Pr(>F)`[2], # any blast effect
            anova(update(m, . ~ . - vns - blast_and_vns), m, test = "F")$`Pr(>F)`[2],    # any vns effect
            anova(update(m, . ~ . - blast_and_vns), m, test = "F")$`Pr(>F)`[2], # interaction / synergy
            anova(update(m, . ~ . - microbe), m, test = "F")$`Pr(>F)`[2] # , # microbe
            # car::linearHypothesis(m, "vns + blast_and_vns = 0")$`Pr(>F)`[2]   # vns on blast
          ))
        
        ################## Combine ##################
        bind_rows(
          fe %>% dplyr::select(term, estimate, std.error, statistic)
          # tibble(
          #   term = "vns_on_blast",
          #   estimate = NA_real_,
          #   std.error = NA_real_,
          #   statistic = NA_real_
          # )
        ) %>%
          left_join(pvals, by = "term") %>%
          mutate(
            microbe = bug_name,
            n = n_samples
          )

        fe %>%
          left_join(pvals, by = "term") %>%
          mutate(
            microbe = bug_name,
            n = n_samples
          )
      }, error = function(e) {
        tibble(
          term      = terms,
          estimate  = NA_real_,
          std.error = NA_real_,
          statistic = NA_real_,
          p.value   = NA_real_,
          microbe   = bug_name,
          n         = n_samples
        )
    })
  }) %>%
    mutate(outcome = out) %>%
    group_by(term) %>%
    mutate(adj.p = p.adjust(p.value, method = "BH")) %>%
    ungroup()
})

# saveRDS(model_outcome, "../A_outcome_models_0113.rds")
saveRDS(model_outcome, "../data/3_newest/A_outcome_models_0212.rds")
head(model_outcome)
```

```{r}
coef_outcome <- outcome_cols %>% # used two_col to test
  set_names() %>%
  map(function(out) {
    
    map_dfr(filt_bugs, function(bug_name) {
    
      df$microbe <- df[[bug_name]]
      df <- df %>% filter(!is.na(out))      # added this to remove NA outcome samples
      form <- as.formula(paste0(out, " ~ microbe + blast + vns + blast_and_vns + Round_main"))
      terms <- c("blast", "vns", "blast_and_vns", "microbe", "Round_main3", "(Intercept)")
      n_samples <- sum(!is.na(df[[bug_name]])) # count non-missing samples for this bug
  
      tryCatch({
        m <- lm(form, data = df, na.action = "na.exclude")
        sm <- summary(m)$coefficients
        
        ################## Fixed effects ##################
        fe <- tibble(
          term      = rownames(sm),
          estimate  = sm[, "Estimate"],
          std.error = sm[, "Std. Error"],
          statistic = sm[, "t value"],
          p.value   = sm[, "Pr(>|t|)"]
        ) %>%
          filter(term %in% terms)
      
        bind_rows(fe %>% mutate(microbe = bug_name, n = n_samples))
        
      }, error = function(e) {
    
        tibble(
          term      = terms,
          estimate  = NA_real_,
          std.error = NA_real_,
          statistic = NA_real_,
          p.value   = NA_real_,
          microbe   = bug_name,
          n         = n_samples
        )
      })
      
    }) %>%
      mutate(outcome = out) %>%
      group_by(term) %>%
      mutate(adj.p = ifelse(term == "(Intercept)", NA_real_, p.adjust(p.value, method = "BH"))) %>%
      ungroup()
  })

# saveRDS(coef_outcome, "../A_outcome_coef_pvals_0206.rds")
# head(coef_outcome)
```

Organize p-values for mediation analysis
```{r}
p_m_df <- model_microbe %>%
  pivot_wider(id_cols = "microbe", names_from = "term",
    values_from = "p.value", names_prefix = "pm_"
    )

p_y_df <- model_outcome %>%
  map_dfr(function(one) {
    one %>%
      dplyr::filter(term == "microbe") %>%
      dplyr::select(microbe, p.value) %>%
      dplyr::mutate(outcome = unique(one$outcome))
  }) %>%
  pivot_wider(
    id_cols = "microbe",
    names_from = "outcome",
    values_from = "p.value"
  )

all_p_df <- p_m_df %>%
  left_join(p_y_df, by = "microbe")

# head(all_p_df)
# saveRDS(all_p_df, "../data/3_newest/A_p_values_0212.rds")
```

Also merge regular coefficient p-values
```{r}
m_coef_df <- coef_microbe %>%
  mutate(
    model = "m",
    term = ifelse(term == "blast_and_vns", "bv", term)
    ) %>%
  transmute(microbe, term, "est" = estimate, "p" = p.value) %>%
  filter(term %in% c("blast", "vns", "bv")) %>%
  pivot_wider(
    id_cols = "microbe", names_from = "term",
    values_from = c("est", "p"),
    names_glue = "{.value}_{term}"
    )

y_coef_df <- coef_outcome %>%
  map_dfr(function(one) {
    one %>%
      mutate(
        outcome = unique(one$outcome),
        model = "y",
        term = ifelse(term == "blast_and_vns", "bv", term)
      )
  }) %>%
  transmute(microbe, outcome, term, "est" = estimate, "p" = p.value) %>%
  filter(term %in% c("blast", "vns", "bv", "microbe")) %>%
  pivot_wider(
    id_cols = c("microbe", "outcome"),
    names_from = "term",
    values_from = c("est", "p"),
    names_glue = "{.value}_{term}"
    )

head(m_coef_df)
head(y_coef_df)

coef_df <- list(
  "m" = m_coef_df,
  "y" = y_coef_df
)

saveRDS(coef_df, "../data/3_newest/A_microbe_coef_dfs_0218.rds")
```

## M-DACT

```{r}
df            <- readRDS("../data/processed/A_clr_df_withalpha_0206.rds")
model_microbe <- readRDS("../data/3_newest/A_microbe_models_0212.rds")
model_outcome <- readRDS("../data/3_newest/A_outcome_models_0212.rds")
all_p_df      <- readRDS("../data/3_newest/A_p_values_0212.rds")
```

`one_mdact` just uses all_p_df hard-coded
```{r}
i <- c("pm_blast", "pm_vns")
j <- outcome_cols

mdact_runs <- expand.grid(exp_name = i, out_name = j, stringsAsFactors = FALSE)

# run everything at FDR = 0.10
all_runs <- pmap_dfr(
  mdact_runs,
  function(exp_name, out_name) {
    one_mdact(
      all_p_df = all_p_df,
      pm = exp_name,
      py = out_name,
      fdr = 0.1,
      diagnostic_root = TRUE
    )
  }
)
```

Run at other FDR levels to get a nice range
```{r}
fdr_vec <- c(0.100, 0.125, 0.150, 0.175, 0.200, 0.225, 0.250)

hit_grid <- expand.grid(
  exp_name = i,
  out_name = j,
  fdr = fdr_vec,
  stringsAsFactors = FALSE
) %>%
  pmap_dfr(function(exp_name, out_name, fdr) {
    one_mdact(
      all_p_df = all_p_df,
      pm = exp_name,
      py = out_name,
      fdr = fdr,
      diagnostic_root = TRUE
    )
  })

# summarize first FDR level at which each mediator is selected
signif_by_fdr <- hit_grid %>%
  group_by(exposure, outcome, mediator) %>%
  summarise(
    signif_at = if (any(selected, na.rm = TRUE)) {
      min(fdr_target[selected], na.rm = TRUE)
    } else {
      NA_real_
    },
    .groups = "drop"
  )

# create a threshold table with both discrete and diagnostic thresholds
thr_wide <- hit_grid %>%
  distinct(
    exposure, outcome, fdr_target,
    dact_thr_discrete, dact_thr_root
  ) %>%
  mutate(
    fdr_lab = paste0("fdr", gsub("\\.", "_", sprintf("%.3f", fdr_target)))
  ) %>%
  dplyr::select(
    exposure, outcome, fdr_lab,
    dact_thr_discrete, dact_thr_root
  ) %>%
  pivot_wider(
    names_from = fdr_lab,
    values_from = c(dact_thr_discrete, dact_thr_root),
    names_glue = "{.value}_{fdr_lab}"
  )

# merge into final object
all_run_df <- all_runs %>%
  left_join(
    signif_by_fdr,
    by = c("exposure", "outcome", "mediator")
  ) %>%
  left_join(
    thr_wide,
    by = c("exposure", "outcome")
  )

saveRDS(all_run_df, "../data/3_newest/A_mediation_0707.rds")
```

## Sensitivity check: does uniroot() change selected mediaators?
```{r}
root_sensitivity <- hit_grid %>%
  mutate(
    selected_root = ifelse(
      is.finite(dact_thr_root),
      dact_stat <= dact_thr_root,
      FALSE
    ),
    disagrees_with_discrete = selected_root != selected
  ) %>%
  group_by(exposure, outcome, fdr_target) %>%
  summarise(
    n_selected_discrete = sum(selected, na.rm = TRUE),
    n_selected_root = sum(selected_root, na.rm = TRUE),
    n_disagreements = sum(disagrees_with_discrete, na.rm = TRUE),
    max_abs_thr_diff = max(abs(dact_thr_root - dact_thr_discrete), na.rm = TRUE),
    .groups = "drop"
  )

root_sensitivity
```
```{r}
root_discrepant_mediators <- hit_grid %>%
  mutate(
    selected_root = ifelse(
      is.finite(dact_thr_root),
      dact_stat <= dact_thr_root,
      FALSE
    ),
    disagrees_with_discrete = selected_root != selected
  ) %>%
  filter(disagrees_with_discrete) %>%
  arrange(exposure, outcome, fdr_target, dact_stat) %>%
  dplyr::select(
    exposure, outcome, fdr_target, mediator,
    dact_stat, selected, selected_root,
    dact_thr_discrete, dact_thr_root
  )

root_discrepant_mediators
```
I AM HERE

```{r}
df <- df %>% filter(!is.na(.data[[out]]))
```


It's hard to get hits for VNS --> PC1 because although it has the lowest DACT stats,
it also has really low threshold for FDR significance due to the p-value distribution.

# 2: Analyzing DA and M-DACT results

Since I already ran prior analyses, I can just import these dfs
```{r}
df            <- readRDS("../data/processed/A_clr_df_withalpha_0206.rds")
model_microbe <- readRDS("../data/3_newest/A_microbe_models_0212.rds")
model_outcome <- readRDS("../data/3_newest/A_outcome_models_0212.rds")
all_p_df      <- readRDS("../data/3_newest/A_p_values_0212.rds")
all_run_df    <- readRDS("../data/3_newest/A_mediation_0707.rds")
```

## Parsing LMER results
### Exposure -> microbe
```{r}
da_bugs <- model_microbe %>%
  filter(adj.p < 0.2, term %in% c("blast", "vns")) %>%
  pull(microbe) %>%
  unique()

lrt_pvals <- model_microbe %>%
  filter(microbe %in% da_bugs) %>%
  transmute(microbe, term, "p" = p.value, "q" = adj.p) %>%
  filter(term %in% c("blast", "vns")) %>%
  pivot_wider(
    id_cols = "microbe", names_from = "term",
    values_from = c("p", "q"),
    names_glue = "{.value}_{term}"
    ) %>%
  mutate(
    blast_signif = ifelse(q_blast < 0.2, TRUE, FALSE),
    vns_signif = ifelse(q_vns < 0.2, TRUE, FALSE),
    microbe = gsub("_", " ", microbe)
  )

da_blast <- setNames(lrt_pvals$blast_signif, nm = lrt_pvals$microbe)
da_vns   <- setNames(lrt_pvals$vns_signif, nm = lrt_pvals$microbe)

```

There are actually hits for blast and VNS!

I think the boxplots are most informative, but I could also plot the estimates as a heat map.

```{r, fig.width = 5, fig.height = 4.5}

m_coef_mat <- m_coef_df %>%
  filter(microbe %in% da_bugs) %>%
  mutate(microbe = gsub("_", " ", microbe)) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(contains("est")) %>%
  as.matrix()

colnames(m_coef_mat) <- c("Blast", "VNS", "Blast:VNS")

m_p_mat <- m_coef_df %>%
  filter(microbe %in% da_bugs) %>%
  mutate(microbe = gsub("_", " ", microbe)) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(contains("p")) %>%
  mutate(
    p_blast = case_when(p_blast < 0.001 ~ "**", p_blast < 0.05 ~ "*", .default = ""),
    p_vns   = case_when(p_vns   < 0.001 ~ "**", p_vns   < 0.05 ~ "*", .default = ""),
    p_bv    = case_when(p_bv    < 0.001 ~ "**", p_bv    < 0.05 ~ "*", .default = "")
    ) %>%
  as.matrix()

absmax <- max(abs(m_coef_mat))
col_fun <- colorRamp2(
  c(-absmax, 0, absmax),
  c("blue", "white", "red")
)

da_blast <- da_blast[rownames(m_coef_mat)]
da_vns   <- da_vns[rownames(m_coef_mat)]

blast_color <- c("TRUE" = "#0492c2", "FALSE" = "#FFFFFF00")
vns_color   <- c("TRUE" = "#FFc067", "FALSE" = "#FFFFFF00")

```

```{r}
lrt_ann_blast<- rowAnnotation(
  "LRT Blast FDR < 0.2" = da_blast,
  col = list("LRT Blast FDR < 0.2" = blast_color),
  show_annotation_name = FALSE,
  simple_anno_size = unit(0.2, "cm")
)

lrt_ann_vns<- rowAnnotation(
  "LRT VNS FDR < 0.2" = da_vns,
  col = list("LRT VNS FDR < 0.2" = vns_color),
  show_annotation_name = FALSE,
  simple_anno_size = unit(0.2, "cm")
)

plot(lrt_ann_vns +lrt_ann_blast)
```


```{r, fig.width=6, fig.height=12}

ht <- Heatmap(m_coef_mat, name = "Estimate", col = col_fun,
  
  cluster_columns = FALSE,
  show_row_dend = FALSE,
  show_column_dend = FALSE,
  left_annotation = c(lrt_ann_vns, lrt_ann_blast),
  
  row_split = da_vns,
  cluster_rows = TRUE,
  cluster_row_slices = TRUE,
  
  cell_fun = function(j, i, x, y, w, h, fill) {
    grid::grid.text(
      m_p_mat[i, j],
      x = x,
      y = y,
      gp = grid::gpar(
        fontsize = 14,
        col = "black",
        fontface = "bold"
      )
    )
  })

ht
```
We see a lot more "effect saturation" here than in the chronic cohort.

```{r}
pdf("../figures/A_microbe_abundance_fdr20.pdf", width = 6, height = 12)
ht_drawn <- draw(ht)
# row_ord <- row_order(ht_drawn)
# da_bug_order <- rownames(trt_effect_matrix)[row_ord]
dev.off()
```

```{r}
pdf("../figures/A_microbe_abundance_fdr20_short.pdf", width = 6, height = 6)
ht_drawn <- draw(ht)
# row_ord <- row_order(ht_drawn)
# da_bug_order <- rownames(trt_effect_matrix)[row_ord]
dev.off()
```

Can also plot/test "presence/absence" if reviewers request, but the results did align well when I tested them for sparse features (as in, sparse bugs that were significant in the CLR regressions were actually even more significant if I did a presence/absence logistic regression, because CLR-transforming zeroes is just adding noise).

### Microbe  -> outcome (independent of exposure)
```{r}
outcome_hits <- map_dfr(model_outcome, function(df) {
  df %>% filter(term == "microbe", adj.p < 0.15)
})

outcome_hits
```

*GGB28456_SGB41039* is associated withacute_cyto_PC2 independent of exposures

These bugs may or may not be affected by treatment, but they explain additional variance in *outcomes*
```{r}
form <- as.formula("acute_cyto_PC2 ~ GGB28456_SGB41039 + blast + vns + blast_and_vns + Round_main")
m    <- lm(form, data = df, na.action = "na.exclude")
summary(m)
```
OK so GGB28456_SGB41039 is associated with higher cytokine PC2 (+1.48),
which is the opposite direction as blast.

### Plot bug vs. residual outcome
```{r}
residout_bug <- function(bug_name, data, outcome_name) {
  plt <- data.frame(
    row.names     = data$seq_name,
    bug_clr       = data[[bug_name]],
    inj_trt_group = factor(data$inj_trt_group, levels = c("Sham + Sham", "Blast + Sham", "Sham + VNS", "Blast + VNS")),
    blast         = data$blast,
    vns           = data$vns,
    blast_and_vns = data$blast_and_vns,
    Round_main    = data$Round_main,
    outcome       = data[[outcome_name]]
  ) %>%
    filter(!is.na(outcome))
  
  residy <- lm(outcome ~ blast + vns + blast_and_vns + Round_main, data = plt)
  plt$out_resid <- resid(residy)
  
  ggplot(plt, aes(x = bug_clr, y = out_resid)) +     # fill = inj_trt_group, shape = Round_main
    geom_point(aes(color = inj_trt_group), size = 4, alpha = 0.7) +
    # scale_fill_manual(values = injtrt_colors) +
    # scale_shape_manual(values = batch_shapes) +
    # geom_smooth(aes(color = inj_trt_group), method = "lm", se = FALSE, alpha = 0.2) +
    scale_color_manual(values = injtrt_colors) +
    geom_smooth(method = "lm", se = FALSE, color = "black",linetype = "dashed") +
    labs(
      x = paste0(gsub("_", " ", bug_name), " Abundance (4h)"),
      y = paste0("Residual", gsub("acute_cyto_", "", outcome_name), " (4h)")
    ) +
    theme_bw() + 
    theme(
      legend.position = "none",
      axis.text = element_text(size = 20),
      axis.title = element_text(size = 20)
      )
}
```
Could use bootstrapping to estimate these effects (slopes) as well. 

Save all these to a single pdf
```{r}
residout_plots <- list()
pdf("../figures/A_not_mediators.pdf", width = 5, height = 5)

for (i in 1:nrow(outcome_hits)) {
  residout_plots[[i]] <- residout_bug(outcome_hits[[i, "microbe"]], df, outcome_hits[[i, "outcome"]])
  print(residout_plots[[i]])
}

# Close the PDF device to save
dev.off()

print(residout_plots[[1]])
```

function to plot multiple to 1 page
```{r}
pdf_split <- function(hit_df, data_df, sub_height, sub_width, n_cols, plot_function, filename) {
  n_plts <- nrow(hit_df)
  n_rows <- ceiling(n_plts / n_cols)
  
  pdf_width <- sub_width * n_cols
  pdf_height <- sub_height * n_rows
  
  plot_vector <- vector("list", n_plts)
  
  for (i in seq_len(nrow(hit_df))) {
    plot_vector[[i]] <- residout_bug(hit_df[[i, "microbe"]], data_df, hit_df[[i, "outcome"]])
  }
  
  pdf(filename, width = pdf_width, height = pdf_height)
  
  grid.arrange(
    grobs = plot_vector,
    ncol = n_cols   # adjust layout
  )
  
  dev.off()
}
```


### Exposure -> microbe -> outcome 
```{r}
med_hits <- all_run_df %>% filter(!is.na(signif_at))
med_hits
```

#### Group-wise box-plot (microbes)
```{r, fig.width = 5, fig.height = 5}
bugs_to_plot <- unique(c(da_bugs, outcome_hits$microbe, med_hits$mediator))
bug_names <- bugs_to_plot %>%
  str_replace_all("sp", "sp.") %>%
  str_replace_all("_", " ")

bug_boxplots <- list()

df$inj_trt_group <- factor(
  df$inj_trt_group,
  levels = c("Sham + Sham", "Blast + Sham", "Sham + VNS", "Blast + VNS")
  )
```

Save all these to a single pdf
```{r}
# Open the PDF device
pdf("../figures/A_microbe_abundance.pdf", width = 5, height = 5) # You can adjust dimensions
# each loop adds a new pdf page
for (i in 1:length(bugs_to_plot)) {
  bug_boxplots[[i]] <- box_easier(df, bugs_to_plot[i], bug_names[i])
  print(bug_boxplots[[i]])
}

# Close the PDF device to save
dev.off()
```



## ACME Bootstrap CI

Estimate mediation effect magnitude with bootstrapping
```{r}
med_hits <- all_run_df %>% filter(!is.na(signif_at))

med_ci_input <- med_hits %>%
  transmute(exposure, mediator, outcome) %>% # no p_method
  mutate(
    exposure = gsub("pm_", "", exposure),  # this assumes no exposures are "vns_on_blast"
    condition = case_when(
      exposure == "blast" ~ "vns",
      exposure == "vns" ~ "blast"
    )
  ) %>%
  split(seq_len(nrow(.))) %>% 
  map(as.list)

med_ci_input[[1]]
```

Had to change some things from `boot_med_lmer` which does a cluster bootstrap. I instead just did a regular (row) resampling bootstrap
```{r}
plan(multisession, workers = 6) # because med_ci_input is a list of 12

ci_list <- future_lapply(med_ci_input, function(x) {
  strat_boot_lm(
    data     = df,
    exposure = x$exposure,
    mediator = x$mediator,
    outcome  = x$outcome,
    n_boot   = 999
    ) }, future.seed = 125)

ci_df <- bind_rows(ci_list)

saveRDS(ci_df, "/proj/gibbons/2024_easton_vns/data/3_newest/A_bootCI_df_0212.rds")
ci_df
```

## Plot coefficients for mediators + annotate with p-values
  Get the order of microbes
### new code
```{r}
# med_bugs <- unique(med_hits$mediator) # luckily they are all unique

med_bugs <- med_hits %>%
  filter(signif_at <= 0.2) %>%
  pull(mediator)

m_coef_mat <- m_coef_df %>%
  filter(microbe %in% med_bugs) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(contains("est")) %>%
  as.matrix()

m_p_mat <- m_coef_df %>%
  filter(microbe %in% med_bugs) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(contains("p")) %>%
  mutate(
    p_blast = case_when(p_blast < 0.001 ~ "**", p_blast < 0.05 ~ "*", .default = ""),
    p_vns   = case_when(p_vns   < 0.001 ~ "**", p_vns   < 0.05 ~ "*", .default = ""),
    p_bv    = case_when(p_bv    < 0.001 ~ "**", p_bv    < 0.05 ~ "*", .default = "")
    ) %>%
  as.matrix()

y_coef_mat <- y_coef_df %>%
  semi_join(med_hits, by = c("outcome", "microbe" = "mediator")) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(contains("est")) %>%
  as.matrix()

y_p_mat <- y_coef_df %>%
  semi_join(med_hits, by = c("outcome", "microbe" = "mediator")) %>%
  column_to_rownames("microbe") %>%
  dplyr::select(contains("p")) %>%
  mutate(
    p_blast = case_when(p_blast < 0.001 ~ "**", p_blast < 0.05 ~ "*", .default = ""),
    p_vns   = case_when(p_vns   < 0.001 ~ "**", p_vns   < 0.05 ~ "*", .default = ""),
    p_bv    = case_when(p_bv    < 0.001 ~ "**", p_bv    < 0.05 ~ "*", .default = ""),
    p_microbe = case_when(p_microbe < 0.001 ~ "**", p_microbe    < 0.05 ~ "*", .default = "")
    ) %>%
  as.matrix()

row_order <- rownames(m_coef_mat)
y_coef_mat <- y_coef_mat[row_order, c(colnames(m_coef_mat), "est_microbe")]
y_p_mat <- y_p_mat[row_order, c(colnames(m_p_mat), "p_microbe")]

row_outcome <- med_hits %>%
  distinct(mediator, outcome) %>%
  filter(mediator %in% rownames(m_coef_mat)) %>%
  tibble::deframe()

row_fdr <- med_hits$signif_at
names(row_fdr) <- med_hits$mediator

row_outcome <- row_outcome[rownames(m_coef_mat)]
row_fdr <- row_fdr[rownames(m_coef_mat)]


stopifnot(all(names(row_outcome) == rownames(m_coef_mat)))
stopifnot(all(names(row_outcome) == rownames(y_coef_mat)))
stopifnot(all(names(row_fdr) == rownames(row_outcome)))

# fdr_fun <- colorRamp2(c(0.00, 0.30), c("darkgreen", "lightyellow"))
fdr_value_vector <- rev(c(0.1, 0.125, 0.15, 0.175, 0.20))

fdr_colors <- structure(
  RColorBrewer::brewer.pal(length(fdr_value_vector), "Greens"),
  names = fdr_value_vector
  )

absmax <- max(abs(c(m_coef_mat, y_coef_mat)))
col_fun <- colorRamp2(
  c(-absmax, 0, absmax),
  c("blue", "white", "red")
)

```

Annotation first
```{r}
fdr_annotation <- rowAnnotation(
  "FDR <" = row_fdr,
  col = list("FDR <" = fdr_colors),
  show_annotation_name = FALSE,
  simple_anno_size = unit(0.2, "cm")
)

plot(fdr_annotation)
```

```{r}
# heatmap model M coefficients
ht_m <- Heatmap(
  m_coef_mat,
  name = "M Estimate",
  col = col_fun,
  left_annotation = fdr_annotation,
  
  row_split = row_outcome,
  cluster_rows = TRUE,
  cluster_row_slices = TRUE,
  cluster_columns = FALSE,
  show_row_dend = FALSE,
  show_column_dend = FALSE,
  cell_fun = function(j, i, x, y, w, h, fill) {
    grid::grid.text(
      m_p_mat[i, j],
      x = x,
      y = y,
      gp = grid::gpar(
        fontsize = 14,
        col = "black",
        fontface = "bold"
      )
    )
  })

# ht_m
```

```{r}
# heatmap model Y coefficients
ht_y <- Heatmap(
  y_coef_mat,
  name = "Y Estimate",
  col = col_fun,
  
  row_split = row_outcome,
  cluster_rows = TRUE,
  cluster_row_slices = TRUE,
  
  cluster_columns = FALSE,
  show_row_dend = FALSE,
  show_column_dend = FALSE,
  cell_fun = function(j, i, x, y, w, h, fill) {
    grid::grid.text(
      y_p_mat[i, j],
      x = x,
      y = y,
      gp = grid::gpar(
        fontsize = 14,
        col = "black",
        fontface = "bold"
      )
    )
  })

# ht_y
```


```{r, fig.width = 8, fig.height = 7}
ht_list <- ht_m + ht_y

draw(ht_list, main_heatmap = "M Estimate")
draw(ht_list, main_heatmap = "Y Estimate")
```

Now save
```{r}
pdf("../figures/A_med_coef_heatmap.pdf", width = 8, height = 7)
ht_drawn <- draw(ht_list, main_heatmap = "M Estimate")
rows <- row_order(ht_drawn)
dev.off()
```

## Plotting blast mediation hits
*Doing this after heat maps so I can use same order as clustering*
```{r}
ht_row_order <- c(unlist(rows[1]), unlist(rows[2]))
names(ht_row_order) <- NULL
ht_row_order

ci_row_order <- row_order[ht_row_order]
ci_row_order
```


```{r}
ci_plt <- ci_df %>%
  filter(mediator %in% ci_row_order) %>%
  mutate(
    exposure = factor(exposure, levels = c("blast", "vns")),
    mediator = factor(mediator, levels = ci_row_order),     # don't reverse
    outcome  = factor(outcome, levels = c("acute_cyto_PC3", "acute_cyto_PC5")),
    exp_med_out  = as.character(interaction(exposure, mediator, outcome, sep = ";"))
    ) %>%
  pivot_longer(
    cols = matches("^(ACME|ADE|Total)_"),
    names_to = c("effect", "cond", ".value"),
    names_pattern = "^([^_]+)_([^_]+)_(.+)$"
  ) %>%
  mutate(
    effect = factor(effect, levels = c("ACME", "ADE", "Total")),
    cond   = factor(cond, levels = c("cond0", "cond1")),
    # exp_med_out = factor(paste0(exposure, ";", outcome, ";", mediator)),
    exp_med_out_cond = factor(paste0(exp_med_out, ";", cond))
    # microbe = factor(mediator)
  ) %>%
  arrange(mediator)

shade_bands <- ci_plt %>%
  group_by(mediator) %>% 
  arrange(by = mediator) %>%
  mutate(
    band_id = as.numeric(mediator),
    fill = ifelse(band_id %% 2 == 1, "grey95", "transparent")
    )

effect_colors <- c(
  "ADE"   = "#9C2121",
  "ACME"  = "#689835",
  "Total" = "#070E00"
)
```

```{r, fig.width = 7, fig.height = 7}

plt <- ggplot(ci_plt, aes(x = est, y = effect, color = effect)) +
  geom_rect(data = shade_bands, aes(
      xmin = -Inf, xmax = Inf,
      ymin = -Inf, ymax = Inf,
      fill = fill
      ), inherit.aes = FALSE, color = NA) +
  geom_point(size = 1.5) +
  geom_linerange(aes(xmin = lower, xmax = upper), linewidth = 0.4) +
  scale_fill_identity() +
  # geom_errorbarh(aes(xmin = mean - se, xmax = mean + se), linewidth = 0.3, height = 0.5) +
  scale_color_manual(values = effect_colors) +
  facet_grid(
    rows = vars(mediator), # if multiple exposure/outcomes, use exp_med_out
    cols   = vars(cond),
    space  = "free_y",
    # switch = "y", scales = "free_y",
    ) +
  # scale_x_continuous(limits = c(xmin, xmax)) +
  labs(x = NULL, y = NULL) +
  guides(fill = guide_legend(reverse = TRUE)) +
  theme_minimal(base_size = 14) +
  theme(
    strip.placement   = "outside",
    panel.spacing.y   = unit(0.1, "lines"),   # try 1-3 "lines" of spacing
    panel.spacing.x   = unit(0.1, "lines"),   # put close together
    axis.text.y       = element_blank(),
    axis.ticks.y      = element_blank(),
    axis.ticks.x      = element_line(color = "black", linewidth = 0.3),
    axis.line.x       = element_line(colour = "black", linewidth = 0.4),
    panel.grid        = element_blank(),
    panel.background  = element_blank()
  )

plt1 <- plt + theme(
  legend.position    = "none",
  strip.text.y.right = element_text(
    angle = 0, margin = margin(l = 4, unit = "pt"), hjust = 0
  ))

plt2 <- plt + theme(legend.position = "none", strip.text.y.right = element_blank())

plt_leg <- plt + theme(strip.text.y.right = element_blank())
legend_only <- get_legend(plt_leg)

plt1
```

do as device instead (so I can save plot with legend)
```{r}
row_height  <- 0.5
panel_width <- 2

pdf("../figures/A_med_blast_cond_blocked.pdf",
    width = (panel_width*2) + 3,
    height = (nlevels(shade_bands$mediator) * row_height))

print(plt1)
plot(legend_only)
dev.off()
```




#### Also plot residual outcome vs. residual bug for the mediators
I want to plot residual outcome vs. bug
```{r}
residout_residbug <- function(bug_name, data, outcome_name) {
  plt <- data.frame(
    row.names     = data$seq_name,
    bug_clr       = data[[bug_name]],
    inj_trt_group = data$inj_trt_group,
    blast         = data$blast,
    vns           = data$vns,
    blast_and_vns = data$blast_and_vns,
    Round_main    = data$Round_main,
    outcome       = data[[outcome_name]]
  ) %>%
    filter(!is.na(outcome))
  
  residy <- lm(outcome ~ blast + vns + blast_and_vns + Round_main, data = plt)
  residx <- lm(bug_clr ~ blast + vns + blast_and_vns + Round_main, data = plt)
  plt$out_resid <- resid(residy)
  plt$bug_resid <- resid(residx)
  
  ggplot(plt, aes(x = bug_resid, y = out_resid)) +     # fill = inj_trt_group, shape = Round_main
    geom_point(aes(color = inj_trt_group), size = 4, alpha = 0.7) +
    # scale_fill_manual(values = injtrt_colors) +
    # scale_shape_manual(values = batch_shapes) +
    # geom_smooth(aes(color = inj_trt_group), method = "lm", se = FALSE, alpha = 0.2) +
    scale_color_manual(values = injtrt_colors) +
    geom_smooth(method = "lm", se = FALSE, color = "black",linetype = "dashed") +
    labs(
      x = paste("Residual \n", gsub("_", " ", bug_name)),  # use expression(italic()) to italicize
      y = paste0("Residual ", gsub("acute_cyto_", "", outcome_name))
    ) +
    theme_bw() +
    theme(
      legend.position = "none",
      axis.text = element_text(size = 14),
      axis.title = element_text(size = 20)
      )
}
```

```{r, fig.width = 5, fig.height = 5}
testplot <- residout_residbug(med_hits[[1, "mediator"]], df, med_hits[[1, "outcome"]])
print(testplot)
```


Save all these to a single pdf
```{r, fig.width = 4, fig.height = 4}
resid_resid_plots <- list()
pdf("../figures/A_outcome_resid_resid_plots.pdf", width = 5, height = 5)

for (i in 1:nrow(med_hits)) {
  resid_resid_plots[[i]] <- residout_residbug(med_hits[[i, "mediator"]], df, med_hits[[i, "outcome"]])
  print(resid_resid_plots[[i]])
}

# Close the PDF device to save
dev.off()
```


